---
title: "Food Tree (Figure 1D)"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)   # data wrangling + readr/tidyr/dplyr/stringr
library(data.tree)   # build hierarchical tree from paths
library(phytools)    # read.newick() for sanity-check loading
```

# Overview

This notebook builds a circular food taxonomy tree (Figure 1D) from the
combined diet table, using GraPhlAn for the final rendering.

The pipeline is:

1.  Build a 7-level food taxonomy from each `Food_code` (one digit per
    level).
2.  Write the taxonomy out as a Newick tree (one for IDs, one for
    names).
3.  Build GraPhlAn annotation files (per-meal average dehydrated weight
    becomes the outer ring; food group → ring color).
4.  Call GraPhlAn (via Docker) to render the PDF.

Code adapted from <https://github.com/knights-lab/Food_Tree>.

# Inputs / outputs

```{r paths}
# ---- inputs ----
dtb_path        <- "../data/152_combined_DTB.csv"          # diet table (per-portion records)
nodes_path      <- "../data/NodeLabelsMCT.txt"             # MCT level-code → human-readable label
fg_color_path   <- "../data/food_group_color_key_final.csv"# food-group → ring color
annot_base_path <- "../data/annotation.base.txt"           # GraPhlAn global style header

# ---- outputs ----
out_table_csv   <- "../data/final_table_for_writing_out_to_newick.csv"
out_tree_id     <- "../data/output_food_tree_datatree.newick"        # leaves = Food_code
out_tree_name   <- "../data/output_food_tree_datatree_name.newick"   # leaves = description
ring_annot_tsv  <- "../data/ring_annot.txt"
ring_style_tsv  <- "../data/ring_itself.txt"

N_LEVELS <- 7   # Food_code is 8 digits; we use the first 7 as taxonomy levels
```

# 1. Load the food database

Take all unique `(Food_code, description)` pairs from the diet table and
sanitize the description (Newick / GraPhlAn dislike spaces, commas,
parentheses, semicolons, slashes, quotes).

```{r load-foods}
dtb <- read_csv(dtb_path)

food_database <- dtb %>%
  distinct(Food_code, description) %>%
  distinct(Food_code, .keep_all = TRUE) %>%   # guard against same code w/ multiple descriptions
  rename(FoodID = Food_code) %>%
  mutate(
    description = description %>%
      str_replace_all("[ ,\\-()%/\";]", "_") %>%
      str_remove_all("'")
  )
```

# 2. Build the 7-level taxonomy table

Each digit position of `FoodID` is a taxonomy level. We look up each
level code in the MCT node table to attach a human-readable label.

```{r build-taxonomy}
nodes <- read_tsv(nodes_path, col_types = cols(.default = "c"))

# Build columns L1..L7 by progressively taking the first 1..N_LEVELS digits of FoodID.
main <- food_database %>%
  select(FoodID, description)

for (k in seq_len(N_LEVELS)) {
  main[[paste0("L", k)]] <- stringr::str_sub(main$FoodID, 1, k)
}

# Replace each level code with its human-readable label from `nodes`.
main_named <- main %>%
  pivot_longer(starts_with("L"), names_to = "Level", values_to = "Level.code") %>%
  left_join(nodes, by = "Level.code") %>%
  select(-Level.code) %>%
  pivot_wider(names_from = Level, values_from = Main.food.description) %>%
  mutate(across(starts_with("L"), ~ replace_na(.x, "")))

# Newick path string (leaves = FoodID, internal nodes = level labels).
final_table <- main_named %>%
  mutate(
    newickstring      = str_glue("foodtreeroot/{L1}/{L2}/{L3}/{L4}/{L5}/{L6}/{FoodID}"),
    newickstring_name = str_glue("foodtreeroot/L1_{L1}/L2_{L2}/L3_{L3}/L4_{L4}/L5_{L5}/L6_{L6}/{description}"),
    taxonomy          = pmap_chr(across(starts_with("L")), ~ paste(c(...), collapse = ";"))
  )

dir.create(dirname(out_table_csv), showWarnings = FALSE, recursive = TRUE)
write_csv(final_table, out_table_csv)
```

# 3. Write Newick trees

Two trees are produced from the same taxonomy: one with `Food_code`
leaves (used downstream for joining numeric data) and one with
`description` leaves (used by GraPhlAn so labels are readable).

```{r write-newick}
write_newick <- function(df, path_col, out_path) {
  tree_obj <- data.tree::as.Node(df, pathName = path_col)
  dir.create(dirname(out_path), showWarnings = FALSE, recursive = TRUE)
  cat(data.tree::ToNewick(tree_obj), file = out_path)
}

write_newick(final_table, "newickstring",      out_tree_id)
write_newick(final_table, "newickstring_name", out_tree_name)
```

# 4. Per-food summaries for the outer ring

Outer-ring height = average dehydrated weight per meal for each food.
Outer-ring color = food-group color (first digit of `Food_code`).

```{r ring-data}
# Total number of distinct meals across the whole cohort (denominator).
n_meal <- dtb %>%
  distinct(pid, fdrt, Meal) %>%
  nrow()

# Tree leaves, in tree order — we need our ring rows to match this order.
leaves <- read.newick(out_tree_id)$tip.label

# Food-group color key, e.g. fgrp1 = "1" → fg_milk → "#..."
fg_colors <- read_csv(fg_color_path, col_types = "ccccc")

# Custom ordering for the food-group factor (drives the ring color sweep).
fg_levels <- c("fg_egg", "fg_oils", "fg_legume", "fg_fruit", "fg_sweets",
               "fg_milk", "fg_meat", "fg_veggie", "fg_grain")

ring_data <- dtb %>%
  group_by(Food_code) %>%
  summarise(fc_sum = sum(dehydrated_weight), .groups = "drop") %>%
  mutate(Food_code = as.character(Food_code)) %>%
  filter(Food_code %in% leaves) %>%
  mutate(
    per_meal_ave = round(fc_sum / n_meal, 3),
    fgrp1        = str_sub(Food_code, 1, 1)
  ) %>%
  left_join(fg_colors, by = "fgrp1") %>%
  mutate(fg1_name = factor(fg1_name, levels = fg_levels)) %>%
  arrange(fg1_name)
```

# 5. GraPhlAn annotation files

GraPhlAn expects a long-format TSV with columns
`<leaf_label>  <key>  <ring_level>  <value>`. Leaf labels here are food
*descriptions*, because that's what the `_name.newick` tree carries.

```{r ring-annotation}
# Map Food_code → sanitized description (the leaf label in the _name tree).
fcode_to_desc <- final_table %>%
  select(FoodID, description) %>%
  mutate(FoodID = as.character(FoodID))

# One row per (food, attribute): height, color, width.
ring_annot <- ring_data %>%
  select(Food_code, ring_height = per_meal_ave, ring_color = color) %>%
  left_join(fcode_to_desc, by = c("Food_code" = "FoodID")) %>%
  mutate(
    ring_width  = 0.5,
    # cast all ring attributes to character so pivot_longer can stack them
    across(c(ring_height, ring_color, ring_width), as.character)
  ) %>%
  pivot_longer(c(ring_height, ring_color, ring_width),
               names_to = "key", values_to = "value") %>%
  mutate(ring_level = 1L) %>%
  select(description, key, ring_level, value)

dir.create(dirname(ring_annot_tsv), showWarnings = FALSE, recursive = TRUE)
write_tsv(ring_annot, ring_annot_tsv, col_names = FALSE)
```

```{r ring-style}
# Per-ring styling (separators + label font). Single ring → one row per attribute.
ring_style <- tribble(
  ~key,                                 ~value,
  "ring_internal_separator_thickness",  "0",
  "ring_external_separator_thickness",  "0",
  "ring_label_font_size",               "5",
  "ring_separator_color",               "black",
) %>%
  mutate(ring_level = 1L) %>%
  select(key, ring_level, value)

write_tsv(ring_style, ring_style_tsv, col_names = FALSE)
```

# 6. Concatenate annotations and render with GraPhlAn

`Annot_tree4.txt` = global header + per-ring style + per-leaf ring
values.

GraPhlAn is run via the `shengwei/graphlan` Docker image. On Apple
Silicon you'll see a platform warning — harmless, but you can silence it
with `--platform linux/amd64`.

```{bash render, eval = FALSE}
# Build the combined annotation file
cat ../data/annotation.base.txt \
    ../data/ring_itself.txt \
    ../data/ring_annot.txt \
  > ../data/Annot_tree4.txt
```


```{bash render, eval = FALSE}
# 1) annotate tree → xml
# 2) xml → pdf
docker run --rm --platform linux/amd64 -v "$(pwd)/../data":/data shengwei/graphlan:latest \
  graphlan_annotate.py \
    --annot /data/Annot_tree4.txt \
    /data/output_food_tree_datatree_name.newick \
    /data/guide_1_new.xml

docker run --rm --platform linux/amd64 -v "$(pwd)/../data":/data shengwei/graphlan:latest \
  graphlan.py \
    /data/guide_1_new.xml \
    /data/F1_D_food_tree_004.pdf \
    --dpi 300 --size 2.5 --pad 0
```

> See <https://huttenhower.sph.harvard.edu/graphlan/> for GraPhlAn
> install notes. 



