---
title: "Modernizing Procrustes"
output: html_document
date: "2026-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(vegan)
```

# 1. Load inputs

```{r load}
dtb  <- read_csv("../data/152_combined_DTB.csv",  show_col_types = FALSE)
meta <- read_csv("../data/153_combined_META.csv", show_col_types = FALSE)
asv  <- read_csv("../data/171_quality_asv_relab_pident97_genus.csv", show_col_types = FALSE)
```

# 2. Keep samples with food records on ALL 5 prior days

```{r filter-samples}
# Daily per-patient food intake aggregated to (pid, fdrt, Food_code)
diet_day <- dtb %>%
  group_by(pid, fdrt, Food_code) %>%
  summarise(dehydrated_weight = sum(dehydrated_weight), .groups = "drop")

# Days that have any food record
diet_days <- diet_day %>% distinct(pid, fdrt) %>% mutate(has = TRUE)

# A sample passes if (pid, sdrt-k) exists for k = 1..5
prior_ok <- meta %>%
  select(pid, sdrt, sampleid) %>%
  tidyr::crossing(k = 1:5) %>%
  mutate(fdrt = sdrt - k) %>%
  left_join(diet_days, by = c("pid", "fdrt")) %>%
  group_by(sampleid) %>%
  summarise(n_ok = sum(!is.na(has)), .groups = "drop") %>%
  filter(n_ok == 5)

meta_keep <- meta %>% semi_join(prior_ok, by = "sampleid")
cat("Samples kept:", nrow(meta_keep), "/", nrow(meta), "\n")
```

# 3. Build the five diet tables and the five matching genus tables

```{r build-tables}
# Long table: for each kept sample, food records on each of the 5 prior days
long_diet <- meta_keep %>%
  select(pid, sdrt, sampleid) %>%
  tidyr::crossing(offset = 1:5) %>%
  mutate(fdrt = sdrt - offset) %>%
  inner_join(diet_day, by = c("pid", "fdrt"))

# Genus relab per (sampleid, genus) — multiple ASVs collapsed by sum
genus_long <- asv %>%
  filter(!is.na(genus)) %>%
  group_by(sampleid, genus) %>%
  summarise(count_relative = sum(count_relative), .groups = "drop")

sids_ordered <- sort(unique(meta_keep$sampleid))

build_pair <- function(W) {
  # Diet: rows = Food_code, cols = sampleid, value = daily avg over W days
  food_tbl <- long_diet %>%
    filter(offset <= W) %>%
    group_by(sampleid, Food_code) %>%
    summarise(daily_avg = sum(dehydrated_weight) / W, .groups = "drop") %>%
    pivot_wider(names_from = sampleid, values_from = daily_avg, values_fill = 0)

  # Ensure every kept sampleid is a column, in order
  for (s in setdiff(sids_ordered, names(food_tbl)[-1])) food_tbl[[s]] <- 0
  food_tbl <- food_tbl[, c("Food_code", sids_ordered)]
  food_tbl <- food_tbl[rowSums(as.matrix(food_tbl[, -1])) > 0, ]

  # Genus table: same sampleid columns
  g_tbl <- genus_long %>%
    filter(sampleid %in% sids_ordered) %>%
    pivot_wider(names_from = sampleid, values_from = count_relative, values_fill = 0)

  for (s in setdiff(sids_ordered, names(g_tbl)[-1])) g_tbl[[s]] <- 0
  g_tbl <- g_tbl[, c("genus", sids_ordered)]
  g_tbl <- g_tbl[rowSums(as.matrix(g_tbl[, -1])) > 0, ]

  list(food = food_tbl, genus = g_tbl)
}

# Write all 10 TSVs with the '#OTU ID' header that biom convert expects
for (W in 1:5) {
  out_dir <- file.path("../data/006_paired_for_procrustes", paste0("p", W, "day"))
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  pair <- build_pair(W)

  food <- pair$food; names(food)[1] <- "#OTU ID"
  write_tsv(food, file.path(out_dir, paste0("d", W, "_diet_foodcode_dehydrated_weight.tsv")))

  g <- pair$genus; names(g)[1] <- "#OTU ID"
  write_tsv(g, file.path(out_dir, paste0("d", W, "_stool_relab_genus.tsv")))

  cat(sprintf("W=%d  food=%d x %d   genus=%d x %d\n",
              W, nrow(food), ncol(food) - 1, nrow(g), ncol(g) - 1))
}
```
```{r}
# Build the five macronutrient tables (rows = macro, cols = sampleid)
MACROS <- c("Protein_g", "Fat_g", "Carbohydrates_g", "Fibers_g", "Sugars_g")

# Aggregate each macronutrient per (pid, fdrt) — daily totals per patient
diet_day <- dtb %>%
  group_by(pid, fdrt) %>%
  summarise(across(all_of(MACROS), sum), .groups = "drop")

diet_days <- diet_day %>% distinct(pid, fdrt) %>% mutate(has = TRUE)

# Long table: kept samples × 5 prior days × per-day macronutrient totals
long_macro <- meta_keep %>%
  select(pid, sdrt, sampleid) %>%
  tidyr::crossing(offset = 1:5) %>%
  mutate(fdrt = sdrt - offset) %>%
  inner_join(diet_day, by = c("pid", "fdrt"))
```


```{r}
build_macro <- function(W) {
  # daily average over W prior days, per sample, per macronutrient
  long_macro %>%
    filter(offset <= W) %>%
    group_by(sampleid) %>%
    summarise(across(all_of(MACROS), ~ sum(.x) / W), .groups = "drop") %>%
    pivot_longer(all_of(MACROS), names_to = "macro", values_to = "daily_avg") %>%
    pivot_wider(names_from = sampleid, values_from = daily_avg, values_fill = 0) %>%
    # ensure every kept sampleid is a column, in order
    { tbl <- .
      for (s in setdiff(sids_ordered, names(tbl)[-1])) tbl[[s]] <- 0
      tbl[, c("macro", sids_ordered)]
    }
}

for (W in 1:5) {
  out_dir <- file.path("../data/006_paired_for_procrustes", paste0("p", W, "day"))

  m <- build_macro(W)
  names(m)[1] <- "#OTU ID"   # biom convert needs this literal header

  write_tsv(m, file.path(out_dir, paste0("d", W, "_diet_macronutrients_per_pt.tsv")))

  cat(sprintf("W=%d  macro=%d x %d\n", W, nrow(m), ncol(m) - 1))
}
```

# 4. Configure Docker / QIIME2

```{r docker-config}
DATA_ABS     <- normalizePath("../data")
QIIME2_IMAGE <- "quay.io/qiime2/qiime2:2026.4"   # adjust to your tag
TREE_QZA     <- "../data/output_food_tree_datatree.qza"  # path inside container, relative to /data

DOCKER_RUN <- sprintf("docker run --rm --platform linux/amd64 -v %s:/data -w /data %s", DATA_ABS, QIIME2_IMAGE)

Sys.setenv(DOCKER_RUN = DOCKER_RUN, TREE_QZA = TREE_QZA)
cat(DOCKER_RUN, "\n")
```

> The food UniFrac step needs `../data/output_food_tree_datatree.qza`

# 5. Run QIIME2 for all five windows in one bash loop

```{bash run-qiime, eval=FALSE}
set -euo pipefail

for W in 1 2 3 4 5; do
  T=d${W}
  P=../data/006_paired_for_procrustes/p${W}day        # path inside the container (under /data)

  echo "===== W=${W} ====="

  # ---------- FOOD: tsv -> biom -> qza -> unweighted UniFrac -> PCoA -> export ----------
  $DOCKER_RUN biom convert \
      -i $P/${T}_diet_foodcode_dehydrated_weight.tsv \
      -o $P/${T}_diet_foodcode_dehydrated_weight.biom \
      --to-hdf5 --table-type="Table"

  $DOCKER_RUN qiime tools import \
      --input-path  $P/${T}_diet_foodcode_dehydrated_weight.biom \
      --output-path $P/${T}_diet_foodcode_dehydrated_weight.qza \
      --type "FeatureTable[Frequency]"

  $DOCKER_RUN qiime diversity beta-phylogenetic \
      --i-table     $P/${T}_diet_foodcode_dehydrated_weight.qza \
      --i-phylogeny $TREE_QZA \
      --p-metric    unweighted_unifrac \
      --o-distance-matrix $P/${T}_diet_foodcode_dehydrated_weight_unweighted_unifrac_distance_matrix.qza

  $DOCKER_RUN qiime diversity pcoa \
      --i-distance-matrix $P/${T}_diet_foodcode_dehydrated_weight_unweighted_unifrac_distance_matrix.qza \
      --o-pcoa            $P/${T}_diet_foodcode_dehydrated_weight_unweighted_unifrac_pcoa.qza

  $DOCKER_RUN qiime tools export \
      --input-path  $P/${T}_diet_foodcode_dehydrated_weight_unweighted_unifrac_pcoa.qza \
      --output-path $P/${T}_diet_foodcode_dehydrated_weight_unweighted_unifrac_pcoa

  # ---------- STOOL: tsv -> biom -> qza -> Bray–Curtis -> PCoA -> export ----------
  $DOCKER_RUN biom convert \
      -i $P/${T}_stool_relab_genus.tsv \
      -o $P/${T}_stool_relab_genus.biom \
      --to-hdf5 --table-type="Table"

  $DOCKER_RUN qiime tools import \
      --input-path  $P/${T}_stool_relab_genus.biom \
      --output-path $P/${T}_stool_relab_genus.qza \
      --type "FeatureTable[Frequency]"

  $DOCKER_RUN qiime diversity beta \
      --i-table  $P/${T}_stool_relab_genus.qza \
      --p-metric braycurtis \
      --o-distance-matrix $P/${T}_stool_relab_genus_braycurtis_distance_matrix.qza

  $DOCKER_RUN qiime diversity pcoa \
      --i-distance-matrix $P/${T}_stool_relab_genus_braycurtis_distance_matrix.qza \
      --o-pcoa            $P/${T}_stool_relab_genus_braycurtis_pcoa.qza

  $DOCKER_RUN qiime tools export \
      --input-path  $P/${T}_stool_relab_genus_braycurtis_pcoa.qza \
      --output-path $P/${T}_stool_relab_genus_braycurtis_pcoa
done
```

```{bash run-qiime-macro, eval=FALSE}
# 5. Run QIIME2 Bray–Curtis + PCoA on macronutrients for W = 1..5
set -euo pipefail

for W in 1 2 3 4 5; do
  T=d${W}
  P=../data/006_paired_for_procrustes/p${W}day        # path inside the container (under /data)

  echo "===== W=${W} ====="

  # tsv -> biom
  $DOCKER_RUN biom convert \
      -i $P/${T}_diet_macronutrients_per_pt.tsv \
      -o $P/${T}_diet_macronutrients_per_pt.biom \
      --to-hdf5 --table-type="Table"

  # biom -> qza (FeatureTable[Frequency])
  $DOCKER_RUN qiime tools import \
      --input-path  $P/${T}_diet_macronutrients_per_pt.biom \
      --output-path $P/${T}_diet_macronutrients_per_pt.qza \
      --type "FeatureTable[Frequency]"

  # Bray–Curtis distance
  $DOCKER_RUN qiime diversity beta \
      --i-table  $P/${T}_diet_macronutrients_per_pt.qza \
      --p-metric braycurtis \
      --o-distance-matrix $P/${T}_diet_macronutrients_per_pt_braycurtis_distance_matrix.qza

  # PCoA
  $DOCKER_RUN qiime diversity pcoa \
      --i-distance-matrix $P/${T}_diet_macronutrients_per_pt_braycurtis_distance_matrix.qza \
      --o-pcoa            $P/${T}_diet_macronutrients_per_pt_braycurtis_pcoa.qza

  # Export PCoA (writes ordination.txt inside the directory)
  $DOCKER_RUN qiime tools export \
      --input-path  $P/${T}_diet_macronutrients_per_pt_braycurtis_pcoa.qza \
      --output-path $P/${T}_diet_macronutrients_per_pt_braycurtis_pcoa
done
```

# 6. Procrustes score calculations

## 1. Helper: read an exported QIIME2 ordination.txt into a Site × PC matrix

```{r helper}
read_pcoa_matrix <- function(fn) {
  read_tsv(fn, skip = 9, col_names = FALSE, show_col_types = FALSE) %>%
    filter(!X1 %in% c("Biplot", "Site constraints")) %>%
    column_to_rownames("X1") %>%
    as.matrix()
}
```

## 2. Build the lookup table of ordination paths for each W


# FOR NOW USE THE results calcualted the previous way!   

Each `qiime tools export` call landed in a folder containing an `ordination.txt`.

```{r build-fns}
Ws <- 1:5

fns <- tibble(
  W      = Ws,
  ID     = paste0("p", Ws, "d"),
  food   = sprintf("/Users/daia1/Work/projects/MSS_pipeline-/scripts/food_tree/data/finalized/paired/pNday/allstool_p%dd_diet_foodID_dehydrated_weight_per_pt_unweighted_unifrac_pcoa/ordination.txt", Ws, Ws),
  macro  = sprintf("/Users/daia1/Work/projects/MSS_pipeline-/scripts/food_tree/data/finalized/paired/pNday/allstool_p%dd_diet_macro_dehydrated_weight_per_pt_bc_pcoa/ordination.txt", Ws, Ws),
  stool  = sprintf("/Users/daia1/Work/projects/MSS_pipeline-/scripts/food_tree/data/finalized/paired/pNday/allstool_p%dd_scts_stool_relab_species_braycurtis_pcoa/ordination.txt", Ws, Ws)
)
fns
```

## 3. Procrustes sum-of-squares for each (diet vs stool) pair

```{r procrustes}
procrustes_ss <- function(dfn, sfn) {
  d <- read_pcoa_matrix(dfn)
  s <- read_pcoa_matrix(sfn)
  # Align to common samples in the same order (Procrustes needs matching rows)
  common <- intersect(rownames(d), rownames(s))
  d <- d[common, , drop = FALSE]
  s <- s[common, , drop = FALSE]
  procrustes(X = d, Y = s)$ss
}

# Food code vs genus
procrustes_food <- fns %>%
  mutate(ss = map2_dbl(food, stool, procrustes_ss)) %>%
  mutate(score = min(ss) - ss) %>%
  select(pNd = ID, W, ss, score)

# Macronutrient vs genus
procrustes_macro <- fns %>%
  mutate(ss = map2_dbl(macro, stool, procrustes_ss)) %>%
  mutate(score = min(ss) - ss) %>%
  select(pNd = ID, W, ss, score)

procrustes_food
procrustes_macro
```

> `score = min(ss) - ss` makes the W with the smallest residual SS score 0 (best),
> and every other W a negative number further from 0. If you want the best W to be
> the *highest* positive number instead, flip the sign: `score = ss - min(ss)` and
> negate, or use `score = max(ss) - ss` for "distance from worst".

## 4. one combined long table

```{r combined}
procrustes_all <- bind_rows(
  procrustes_food  %>% mutate(pair = "Food group based"),
  procrustes_macro %>% mutate(pair = "Macronutrient based")
)
procrustes_all
```

## 5. f2a

```{r plot, fig.width=6, fig.height=4}
f2a <- procrustes_all %>%  
  mutate(pNd = str_extract(pNd, '\\d')) %>% 
  arrange(desc(pair)) %>% 
  mutate(pNd =  factor(pNd)) %>% 
  ggplot(aes(x = pNd, y = score, group = pair, linetype = pair)) +
  geom_line(size = 0.3) +
  geom_point(size = 1.3) +
  labs(x = 'diet exposure days',y = 'Procrustes score') +
  scale_linetype_manual(values = c('solid','dashed')) +
  theme_classic()  +
  theme(aspect.ratio=1.2,
        legend.position = 'right',
        axis.text=element_text(size=10), 
        axis.title=element_text(size=10),
        legend.title = element_blank(), 
        axis.text.x = element_text(angle=0, hjust=0.5)) 
f2a 

ggsave(
  '../data/F2a_006_procrustes.pdf', 
  plot = f2a, 
  width = 150, 
  height = 50, 
  units = "mm", 
  device = 'pdf'
)
```

