---
title: "The fig1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Now this is the script for making the updated Figure 1 with the combined cohort 173 patients.

# Libraries

```{r setting}
library(tidyverse)
library(ggpubr)
library(cowplot)
library(ggimage)
library(ggrastr)
library(lemon)
library(viridis)
```

# Params

```{r setting}
axis_text_size <- 10
axis_title_size <- 10
stip_txt_size <- 8
alpha_val <- 0.05
point_size <- 0.2
scatter_col <- 'gray14'
pick_point_size <- 1
pick_line_size <- 0.5
example_pt_line_size <- 0.3
diet_line_color <- '#E41A1C'
stool_line_color <- 'blue'
strip_color <- 'gray91'
umap_pt_size <- 0.2
key <- read_csv('../data/food_group_color_key_final.csv', col_types = 'ccccc')
transplant_day <- 0
dayzeroline_size <- 1
scatter_transparency <- 0.1
day0_line_color <- 'gray40'

# the colors for each food group specified and saved here
fpalette <- tribble(
  ~fg1_name, ~color, ~fgrp1,
  #--|--|
  'fg_milk' , '#3498DB', '1',
  'fg_meat' , '#591605', '2',
  'fg_egg' , '#F1C40F', '3',
  'fg_legume' , '#E67E22', '4',
  'fg_grain' , '#D35400', '5',
  'fg_fruit' , '#7D3C98', '6',
  'fg_veggie' , '#229954', '7',
  'fg_oils' , '#707B7C',  '8',
  'fg_sweets' , '#db2589' , '9'
) 
```

# Data summaries

The first 2 need to be redo, since the data has been updated.

```{r load_table}
dtb  <- read_csv('../data/152_combined_DTB.csv')
meta <- read_csv('../data/153_combined_META.csv')

# how many stool samples per patient
meta %>% count(pid) %>% summarise(med = median(n), min_ = min(n), max_ = max(n))
```


```{r load_table}
dtb %>% distinct(Food_code)

# how many days of data among all of the patients?
dtb %>% count(pid,  fdrt) %>% 
  count(pid) %>% 
  summarise(total_days = sum(n))

dtb %>% distinct(Food_NSC)

dtb %>% count(pid, fdrt) %>% 
  distinct(pid, fdrt) 

range(dtb$fdrt)
  

# summary statistics 
# median N samples per patient
n_per_patient <- meta %>% 
  count(pid) 
summary(n_per_patient$n)

# max number of meals across all patients 
# calculate how many meals each patient had
tallyres <- dtb %>% 
  distinct(pid, Meal, fdrt) %>% 
  count(pid)
range(tallyres$n)

# “1 day prior to this fecal sample, the patient consumed 50 g of chicken, find a patient for this example
# find the food code that is this chicken 24122160
daily_chicken <- dtb %>% 
  filter(Food_code == 24122160) %>% 
  group_by(pid, fdrt) %>% 
  summarise(daily_ch = sum(total_weight))

# find the prior one day for every stool sample
p1d <- meta %>% 
  select(pid, sdrt) %>% 
  mutate(fdrt = sdrt - 1) %>% 
  inner_join(daily_chicken, by = join_by(pid, fdrt)) %>% 
  filter(sdrt < 0) %>% 
  arrange(desc(daily_ch)) %>% 
  slice(1)
```

```{r spectral_legend}
library(RColorBrewer)
pdf(file = "../intermediate_data/072_F1_spectral_legend.pdf",   # The directory you want to save the file in
    width = 4, # The width of the plot in inches
    height = 3)
display.brewer.pal(9, "Spectral")
dev.off()
```

```{r}
# summarize how many diet data days we have for the patients?  Median min and max.  
dtb %>% 
  distinct(pid, fdrt) %>% 
  #count(pid) %>% 
  summary

# -how common was it? how many of 97 patients consumed at least 1?
#-what fraction of meals (or fraction of days) included a smoothie?
#-for how many meals  (or days) was smoothie the only thing consumed?

# find the food code that has the highest per meal average consumption
smoothie <- dtb %>% 
  group_by(Food_code, description) %>% 
  summarise(total = sum(dehydrated_weight))  %>% 
  arrange(desc(total)) %>% 
  ungroup() %>% 
  slice(1)  %>% 
  pull(Food_code)
```
# 1A: transplant schematic

```{r tree}
# import the tree from pdf to a ggplot kind of object
schematic <- '../intermediate_data/transplant_schematic.png'
schematic_plot <- ggdraw() +
  draw_image(schematic,   scale = 1.1)  
```


# 1B: diet_hist

```{r diet_hist}
# the histogram of the N meals over transplant time
diet_hist <- dtb %>% 
  distinct(pid, Meal, fdrt) %>% 
  gghistogram(x = 'fdrt', xlab = 'Transplant day', ylab = 'Meals recorded', 
              color = diet_line_color, fill = diet_line_color, alpha = 0.5)+
    scale_x_continuous( breaks = seq(0, 50, 20)) + 
  geom_vline(xintercept = transplant_day, linetype = 'dashed', size = dayzeroline_size, color = day0_line_color) +
  annotate("text", x = 30, y = 600, label = "173 patients,\n9,419 total meals\n40,702 food items", hjust = 0, size = 2.5) +
  theme_classic(base_size = 14) +
  theme(aspect.ratio=1/1.5,
         axis.title=element_text(size=axis_title_size))
    
diet_hist
```



# 1C: stool_hist 

```{r stool_hist}
# a small histogram of N samples and N meals recorded over transplant time. That would show both the depth of sampling and emphasize the temporal nature of the dataset 
# the histogram of the N stool samples over transplant time
stool_hist <- meta %>% 
  ggplot(aes(x = sdrt)) +
  geom_histogram(alpha = 0.5, fill = stool_line_color, color = stool_line_color) +
  geom_vline(xintercept = transplant_day, linetype = 'dashed', size = dayzeroline_size, color = day0_line_color) +
  labs(x = 'Transplant day', y = 'Stool samples') +
  scale_x_continuous( breaks = seq(0, 50, 20)) + 
  annotate("text", x = 30, y = 60, label = "1,009\nstool samples", hjust = 0, size = 2.5) +
  theme_classic(base_size = 14) +
  theme(aspect.ratio=1/1.5,   axis.title=element_text(size=axis_title_size))
stool_hist   
```

# 1D: food_tree

```{r tree}
# import the tree from pdf to a ggplot kind of object
p <- '../intermediate_data/080_food_tree_ring.jpg'
tree <- ggdraw() +
  draw_image(p,   scale = 1.1) 
```

# 1M: stool line 

```{r stool_alpha}
stool_alpha <- meta %>% 
     ggplot() +
  rasterise(geom_point(aes(x = sdrt, y = simpson_reciprocal, shape = 'triangle'), alpha = 0.3,size = point_size , shape = 16), dpi = 300) +
  geom_smooth(aes(x = sdrt, y = simpson_reciprocal), method = "loess", formula = 'y ~ x', colour = stool_line_color ,  size = 1, fill = 'darkblue') +
  geom_vline(xintercept = transplant_day, linetype = 'dashed', size = dayzeroline_size, color = day0_line_color) +
  labs( x = '',title = 'Microbiome\nalpha diversity', y = '') +
  scale_y_sqrt() +
  theme_classic(base_size = 14) +
   scale_x_continuous( breaks = seq(0, 50, 20)) + 
  theme( axis.title=element_text(size=axis_title_size),
        #plot.title = element_blank(),
        aspect.ratio=1/1.5)
stool_alpha
```


```{r stool_alpha}
# to calculate a p value for the dropping trend of the line of the stool alpha diversity

library(geepack)
# Gee model
# Using simple linear model to easily describe increasing/decreasing trends
mod_stool_alpha <- geeglm(log(simpson_reciprocal) ~ sdrt,
                          family=gaussian,
                          corstr = "ar1",
                          id=pid,
                          data=meta %>% mutate(pid = as.factor(pid)))

# Wald p-value
# Alternative hypothesis: slope is less than zero
# Z-score calculation
zscore = coef(mod_stool_alpha)[2]/sqrt(vcov(mod_stool_alpha)[2,2])
# One-sided p-value
pval <- 1-pnorm(abs(zscore))
# the p value for the 1K
pval
```

# 1I: cal line

```{r }
day_calori <- dtb %>% 
  group_by(pid, fdrt) %>% 
  summarise(daycal = sum(Calories_kcal))

cal_line <- day_calori %>% 
  mutate(daycal = daycal/1000) %>%
       ggplot() +
  rasterise(geom_point(aes(x = fdrt, y = daycal), alpha = 0.3,size = point_size , shape = 16), dpi = 300) +
  geom_smooth(aes(x = fdrt, y = daycal), method = "loess", formula = 'y ~ x', colour = diet_line_color ,  size = 1, fill = 'hotpink') +
  geom_vline(xintercept = transplant_day, linetype = 'dashed', size = dayzeroline_size, color = day0_line_color) +
  labs( x = '',title = 'Caloric intake', y = '*1000 Kcal') +
  scale_x_continuous( breaks = seq(0, 50, 20)) +
  scale_y_sqrt() +
  theme_classic(base_size = 14) +
  theme(
        #axis.text.x=element_blank(),
        #axis.title=element_text(size=8),
        strip.background = element_blank(),
        strip.text.x = element_blank(), 
        plot.title = element_text(size=axis_title_size),
        aspect.ratio=1/1.5)
cal_line
```
# 1J: diet alpha line

```{r}
# this faith pd is all the combined data of the patients
faith <- read_csv('../data/162_all_faith_pid.csv')

faith_line <- faith %>% 
  mutate(faith_pd = faith_pd/1000) %>%
       ggplot() +
  rasterise(geom_point(aes(x = fdrt, y = faith_pd), alpha = 0.3,size = point_size , shape = 16), dpi = 300) +
  geom_smooth(aes(x = fdrt, y = faith_pd), method = "loess", formula = 'y ~ x', colour = diet_line_color ,  size = 1, fill = 'hotpink') +
  geom_vline(xintercept = transplant_day, linetype = 'dashed', size = dayzeroline_size, color = day0_line_color) +
  labs( x = 'Transplant day',title = 'Diet a diversity', y = '*1000') +
  scale_x_continuous( breaks = seq(0, 50, 20)) +
  scale_y_sqrt() +
  theme_classic(base_size = 14) +
  theme(
        #axis.text.x=element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(), 
        plot.title = element_text(size=axis_title_size),
        aspect.ratio=1/1.5)
faith_line
```



```{r}
# the p value from GEE about the G and H
#1G caloric intake
mod_cal <- geeglm(log(daycal+1) ~ fdrt,
                          family=gaussian,
                          corstr = "ar1",
                          id=pid,
                          data=day_calori %>% mutate(pid = as.factor(pid)))

# Wald p-value
# Alternative hypothesis: slope is less than zero
# Z-score calculation
zscore = coef(mod_cal)[2]/sqrt(vcov(mod_cal)[2,2])
# One-sided p-value
pval <- 1-pnorm(abs(zscore))
# the p value for the 1G
pval
```
```{r}
# 1H diet diversity
mod_faith_pd <- geeglm(log(faith_pd) ~ fdrt,
                          family = gaussian,
                          corstr = "ar1",
                          id = pid,
                          data= faith %>%  mutate(pid = as.factor(pid)))

# Wald p-value
# Alternative hypothesis: slope is less than zero
# Z-score calculation
zscore = coef(mod_faith_pd)[2]/sqrt(vcov(mod_faith_pd)[2,2])
# One-sided p-value
pval <- 1-pnorm(abs(zscore))
# the p value for the 1H
pval
```

# 1E-H: UMAPs

```{r}
# calculate the dominant food group of that day
key <- read_csv('../data/food_group_color_key_final.csv') %>% mutate(fgrp1 = as.character(fgrp1))

fg_daily <- dtb %>% 
  group_by(pid, fdrt ) %>% 
  summarise(daily_total = sum(dehydrated_weight))

fg_sum <- dtb %>% 
  mutate(fgrp1 = str_sub(Food_code, 1, 1)) %>% 
   group_by(pid, fdrt , fgrp1) %>% 
  summarise(daily_fg = sum(dehydrated_weight)) %>% 
  left_join(fg_daily) %>% 
  mutate(fg_relab = daily_fg/daily_total) %>% 
  ungroup() %>% 
  left_join(key) %>% 
  group_by(pid, fdrt) %>% 
  arrange(desc(fg_relab)) %>% 
  slice(1)

# find out why the few dots are on that far island
OUT <- read_csv('../intermediate_data/072_taxumap_embedding_cleaned_pid.csv')
  
color_key <- OUT %>% 
  distinct( color, shortname) %>% select(shortname, color) %>% 
  deframe()
```

## 1G: dominant food group

```{r}
gray_bg_color <- 'gray98'
out2_umap <- OUT %>% 
  #arrange(desc(n)) %>% 
  ggplot() +
  rasterise(geom_point(aes(x = taxumap1, y = taxumap2, color = shortname), alpha = 1,size = umap_pt_size , shape = 16), dpi = 300) +
  labs(x = 'TaxUMAP1', y = 'TaxUMAP2', title = 'Most consumed\nfood group') +
  scale_color_manual(values = color_key) +
  theme_classic() +
  #lemon::coord_capped_cart(bottom = 'both', left = 'both') +
  theme(line = element_blank(),
        legend.position = "none",
        legend.title = element_blank(),
        axis.title=element_blank(),
        axis.text= element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill=gray_bg_color), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent') , aspect.ratio=1) +
  guides(colour = guide_legend(override.aes = list(alpha=1)))
out2_umap
```

## 1E: daily cal intake

```{r} 
# calculate daily caloric intake   
cal <- dtb %>% 
  group_by(pid, fdrt) %>% 
  summarise(total_cal = sum(Calories_kcal)) 
out2_cal <- OUT %>% full_join(cal)

out2_caloric <- out2_cal %>% 
  mutate(transform_cal = sqrt(total_cal)) %>% 
    ggplot() +
  rasterise(geom_point(aes(x = taxumap1, y = taxumap2, color = transform_cal), alpha = 1,size = umap_pt_size , shape = 16), dpi = 300) +
  paletteer::scale_color_paletteer_c("viridis::plasma") +
  labs(title = 'Caloric intake\n(*1000 Kcal)') +
  theme_classic() +
  #lemon::coord_capped_cart(bottom = 'both', left = 'both') +
  theme(line = element_blank(),
        legend.position = "none",
        legend.title = element_blank(),
        axis.title=element_blank(),
        axis.text= element_blank(),
        axis.ticks = element_blank(),
         panel.background = element_rect(fill=gray_bg_color), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent') ,
        aspect.ratio=1)  
```

## 1F: diet alpha diversity

```{r}


faith <- read_csv('../data/162_all_faith_pid.csv')

out_faith <- OUT %>% 
  full_join(faith)


umap_faith <- out_faith %>% 
  #arrange(desc(n)) %>% 
  mutate(log10faith = log10(faith_pd + 1)) %>% 
    ggplot() +
  rasterise(geom_point(aes(x = taxumap1, y = taxumap2, color = faith_pd), alpha = 1,size = umap_pt_size , shape = 16), dpi = 300) +
  scale_color_viridis() +
  labs(title = 'Diet a diversity\n(*1000)') +
  theme_classic() +
  theme(line = element_blank(),
        legend.title = element_blank(),
        axis.title=element_blank(),
        axis.text= element_blank(),
        axis.ticks = element_blank(),
        legend.position = 'none', 
        aspect.ratio=1,
                 panel.background = element_rect(fill=gray_bg_color), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent') 
        )
```


## 1H: time relative to transplant

```{r}
# cuztomizing the times bins of the drt
cutoff = 0 # Day 0 is conveneint but this can be varied
n_groups_pre = 3;  # all HCT days <0 will be split into 3 bins; can vary this number to play
n_groups_post = 6; # all HCT days >0 will be split into 6 bins; can vary this number to play
# split the fdrt into groups:
umap_time <- OUT %>% 
  select(taxumap1, taxumap2, fdrt)  
times <- umap_time %>% select( fdrt)  
splits <- times %>% 
  split(.$fdrt <= 0)
pre <- splits %>% 
  pluck('TRUE') %>% 
  mutate(bin = cut_number(fdrt, n_groups_pre))
post <- splits %>% 
  pluck('FALSE') %>% 
  mutate(bin = cut_number(fdrt, n_groups_post)) 
both <- bind_rows(pre, post) %>% 
  distinct(fdrt, .keep_all = T)

umap_time_df <- umap_time %>% 
  left_join(both)

umap_drt_high_to_low_bin <- umap_time_df %>% 
  arrange(desc(fdrt)) %>% 
    ggplot() +
  rasterise(geom_point(aes(x = taxumap1, y = taxumap2, color = bin), alpha = 1,size = umap_pt_size , shape = 16), dpi = 300) +
   scale_color_manual(values = RColorBrewer::brewer.pal(9,"Spectral"))  +
  labs(title = 'Day relative to\ntransplant') +
   theme_classic() +
   theme(
         line = element_blank(),
         legend.position = "none",
         legend.title = element_blank(),
         axis.title=element_blank(),
         axis.text= element_blank(),
         axis.ticks = element_blank(),
              panel.background = element_rect(fill=gray_bg_color), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent') ,
        aspect.ratio=1)
```


```{r}
# the days that we know the patient didn't eat anythinng

zerodaysall <-  read_csv('../data/072_total_patients_zero_eating_days_pid.csv')
```

```{r 3X2}
# the macronutrients scatter plot
# good I've checked there are no zeros that I need to be put in
m_all <- dtb %>%  
  select(pid, fdrt,Protein_g:Sugars_g ) %>% 
  gather('grp','gram', Protein_g:Sugars_g) %>% 
  mutate(grp = str_replace(grp, '_g$','')) %>% 
  group_by(pid, fdrt, grp) %>% 
  summarise(eachsum = sum(gram)) %>% 
  spread('grp', 'eachsum')

# are there any NAs
m_all %>% 
  gather('key','value', Carbohydrates:Sugars) %>% 
  filter(is.na(value)) %>% nrow

dailymacro <- bind_rows(
  m_all,
  zerodaysall %>% 
    mutate(Carbohydrates = 0, Fat= 0, Fibers= 0, Protein= 0, Sugars= 0) %>% 
    select(colnames(m_all))
) %>% gather('grp','eachsum', Carbohydrates:Sugars)
  
# dailymacro %>% 
#   filter(grp %in% c('Carbohydrates','Sugars')) %>% 
#   distinct(pid, fdrt, grp, .keep_all = T) %>% 
#   spread('grp','eachsum') %>% 
#   ggscatter( x= 'Carbohydrates',y ='Sugars') +
#   stat_cor(method = "pearson") +
#   geom_abline(slope = 1, intercept = 0, color = 'red')

dailymacro %>% ungroup() %>% 
  filter(grp == 'Fibers') %>% 
  summarise(med = median(eachsum))
```
```{r}
# the p value from GEE about the panels in each of these macronutrients
macro5 <- dailymacro %>% 
  filter(fdrt <= 12) %>% 
  split(.$grp) %>% 
  map(function(df){
    mod_macro = geeglm(log(eachsum +0.5) ~ fdrt,
                          family=gaussian,
                          corstr = "ar1",
                          id=pid,
                          data=df %>% mutate(pid = as.factor(pid)))
    
    zscore = coef(mod_macro)[2]/sqrt(vcov(mod_macro)[2,2])

      pval = 1-pnorm(abs(zscore))
      return(pval)
  }) %>% 
  bind_rows(.id = 'grp') %>% 
  mutate(pval_lt_001 = if_else(fdrt < 0.01, T, F))%>% 
  mutate(pval_lt_005 = if_else(fdrt < 0.05, T, F))%>% 
  mutate(FDR = p.adjust(fdrt, method = 'BH')) %>% 
  mutate(q_lt_005 = if_else(FDR < 0.05, T, F))
 
macro5
# how to show these as p values 
```
# 1K: the macronutrients 3 by 2 grid

```{r 3X2}
m_panel <- dailymacro %>% 
  mutate(grp = if_else(str_detect(grp, 'Carbohydrates'), str_replace(grp, 'Carbohydrates','Carbs'), grp)) %>% 
   ggplot() +
  rasterise(geom_point(aes(x = fdrt, y = eachsum), alpha = scatter_transparency,size = point_size , shape = 16), dpi = 300) +
  geom_smooth(aes(x = fdrt, y = eachsum), method = "loess", formula = 'y ~ x', colour = diet_line_color ,  size = 1, fill = 'hotpink') +
  geom_vline(xintercept = transplant_day, linetype = 'dashed', size = dayzeroline_size, color = day0_line_color) +
  labs( x = 'Transplant day',title = 'Macronutrients', y = 'Grams') +
  facet_wrap(~ grp, nrow = 3, scales = 'free_y') +
  scale_x_continuous( breaks = seq(0, 50, 20)) + 
  scale_y_sqrt() +
  theme_classic() +
  theme(axis.text=element_text(size=6),
        strip.background = element_rect(color="white", fill=strip_color, size=1.5,linetype="solid"),
        strip.text.x = element_text(size = stip_txt_size), 
        axis.title=element_text(size=axis_title_size),
        plot.title = element_text(size=axis_title_size),
        aspect.ratio=1)
m_panel
```

# 1L: the food group  3 by 3 grid 

```{r 3X3}
fg_all <- dtb %>%  
  mutate(fgrp1 = str_sub(Food_code, 1, 1)) %>% 
  mutate(fgrp1 = as.numeric(fgrp1)) %>% 
  group_by(pid, fdrt, fgrp1) %>% 
  summarise(eachsum = sum(dehydrated_weight)) %>% 
  mutate(fgrp1 = as.character(fgrp1)) %>% 
  left_join(key %>% select(fgrp1, shortname)) %>% 
  select(-fgrp1) %>% 
  spread('shortname','eachsum', fill = 0)

fg_total <- bind_rows(
  fg_all,
  zerodaysall %>% 
    mutate(Eggs = 0, Fruits= 0, Grains= 0, Legumes= 0, Meats= 0, Milk= 0,Oils= 0,Sweets= 0,Vegetables= 0) %>% 
    select(colnames(fg_all))
) %>%  gather('grp','eachsum', Eggs:Vegetables)
  

fg_panel <- fg_total %>% 
     ggplot() +
  rasterise(geom_point(aes(x = fdrt, y = eachsum), alpha = scatter_transparency,size = point_size , shape = 16), dpi = 300) +
  geom_smooth(aes(x = fdrt, y = eachsum), method = "loess", formula = 'y ~ x', colour = diet_line_color ,  size = 1, fill = 'hotpink') +
  geom_vline(xintercept = transplant_day, linetype = 'dashed', size = dayzeroline_size, color = day0_line_color) +
  labs( x = 'Transplant day',title = 'Food groups', y = 'Grams') +
  facet_wrap(~ grp, 
             scales = 'free_y',
             nrow = 3 ) +
  scale_x_continuous( breaks = seq(0, 50, 20)) + 
  scale_y_sqrt() +
  theme_classic() +
  theme(axis.text=element_text(size=6),
       strip.background = element_rect(color="white", fill=strip_color, size=1.5,linetype="solid"),
        strip.text.x = element_text(size = stip_txt_size), 
        axis.title=element_text(size=axis_title_size),
        plot.title = element_text(size=axis_title_size),
        aspect.ratio=1)
fg_panel
```
 
 
```{r}
# the p value from the GEE
fg9 <- fg_total %>% 
  filter(fdrt <= 12) %>% 
  split(.$grp) %>% 
  map(function(df){
    mod_fg = geeglm(log(eachsum +0.5) ~ fdrt,
                          family=gaussian,
                          corstr = "ar1",
                          id=pid,
                          data=df %>% mutate(pid = as.factor(pid)))
    
    zscore = coef(mod_fg)[2]/sqrt(vcov(mod_fg)[2,2])

      pval = 1-pnorm(abs(zscore))
      return(pval)
  }) %>% 
  bind_rows(.id = 'grp') %>% 
  mutate(pval_lt_001 = if_else(fdrt < 0.01, T, F))%>% 
  mutate(pval_lt_005 = if_else(fdrt < 0.05, T, F)) %>% 
  mutate(FDR = p.adjust(fdrt, method = 'BH')) %>% 
  mutate(q_lt_005 = if_else(FDR < 0.05, T, F))
 
fg9
```
 
# Final F1

```{r}
title <- ggdraw() + 
  draw_label("Fig. 1",fontface = 'bold',x = 0,hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 5))


# assemble the two hists
two_hist <- plot_grid(diet_hist,  stool_hist, labels = c('B', 'C'),
          ncol = 2, hjust = 0.5,
          align = 'hv',
          axis = 'tblr') 


# top row 
top_left <- plot_grid(schematic_plot, two_hist,
          nrow = 2, 
          labels = c('A', NA))

top_row <- plot_grid(top_left, tree,
          nrow = 1, 
          labels = c(NA, 'D')) 


# assemble the 4 UMAPs
middle_maps <- plot_grid( out2_caloric, umap_faith, out2_umap, umap_drt_high_to_low_bin,
          nrow = 1, hjust = 0,
          align = 'hv',
          labels = c( 'E', "F", 'G','H'),
          axis = 'tblr')

ggsave('../results/F1_B_C_0072.pdf', plot = two_hist)
```

# bottom_left 

```{r}
bottom_left <- plot_grid(cal_line, NA, faith_line, NA, stool_alpha,
                      rel_heights = c( 1,-0.4, 1, -0.4, 1) ,
                      labels = c('I',NA,'J',NA,'M'), 
          ncol = 1, hjust = 0.5,
          align = 'hv',
          axis = 'tblr')


bottom_row <- plot_grid(bottom_left,m_panel, fg_panel, 
          nrow = 1, 
          #align = 'h',
          labels = c(NA, 'K', 'L'),
          #axis = 'tblr',
          rel_widths =  c(.7,1.4,2))

# assemble all macro_facets
f1 <-  plot_grid(top_row, middle_maps, bottom_row, 
                 align = 'hv',axis = 'tblr',
                 #labels = c(NA, NA, NA, NA, 'L'),
                 rel_heights = c(1,0.7, 1.5),
                 ncol = 1)
```


```{r}
combined <- plot_grid(
  title, f1,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.1, 1)
) +theme(plot.margin = unit(c(1,1,1,1), "cm"))

ggsave('../results/F1_overview_raw_072_new_test.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', 
      dpi = 300)    
```

# NEW fig 1

## generate the bmt schematic using ggplot

```{r}
# 1. Generate sample data to create the curve
# A cosine function is a good way to model this dip and recovery.
# We adjust parameters to get the trough around day 6 and reasonable start/end points.
day_seq <- seq(-4, 18.5, by = 0.1) # Extended sequence
plot_data <- data.frame(day = day_seq) |>
  mutate(
    # This cosine function:
    # - Has trough at day 6.5
    # - Has amplitude of 0.5, so trough (y_min) is (0.5 - 0.5) = 0
    # - Has period of 22 days, so half-period (dip) is 11 days
    # - Crosses y=0.5 at day = 6.5 - 5.5 = 1
    # - Crosses y=0.5 at day = 6.5 + 5.5 = 12
    # - Peaks at y=1.0 at day = 6.5 + 11 = 17.5
    n = -0.5 * cos((pi / 11) * (day - 6.5)) + 0.5
  ) |>
  # Cap the value at 1.0 to create the flat tail
  mutate(neutrophils = ifelse(day > 17.5, 1.0, n)) |>
  # Ensure count doesn't go below zero (though min is 0)
  mutate(neutrophils = ifelse(neutrophils < 0, 0, neutrophils)) |>
  select(-n) # Clean up intermediate column

# 2. Create a separate dataset for the "neutropenia" ribbon
# This ribbon shades the area where neutrophils are below 0.5
ribbon_data <- plot_data |>
  filter(neutrophils < 0.5) |> # This filter correctly finds the dip
  mutate(
    ymin = neutrophils, # Bottom of ribbon is the curve
    ymax = 0.5          # Top of ribbon is the horizontal line
  )

# 2b. Create data for the gradient "nausea" bar
gradient_days <- seq(-2, 14, by = 0.1) # Fade from -2 to 0, and 12 to 14
gradient_data <- data.frame(day = gradient_days) |>
  mutate(
    y_pos = 0.85, # Y-position for the center of the bar
    # Create the alpha gradient:
    # 0 -> 0.8 from day -2 to 0
    # 0.8 from day 0 to 12
    # 0.8 -> 0 from day 12 to 14
    alpha_val = case_when(
      day < 0  ~ pmax(0, 0.8 * (1 - (day / -2))), # Ramps up
      day > 12 ~ pmax(0, 0.8 * (1 - ((day - 12) / 2))), # Ramps down
      TRUE     ~ 0.8 # Solid center
    )
  )

# 3. Build the ggplot
bmt_plot <- ggplot(plot_data, aes(x = day, y = neutrophils)) +

  # --- Annotations (Bars & Shading) ---
  
  # "conditioning" bar (using a thick segment)
  annotate(
    "segment",
    x = -4, xend = 0, y = 1.05, yend = 1.05,
    color = "#6a3d9a", # A deep purple
    linewidth = 6,
    lineend = "butt"
  ) +
  
  # "nausea, mucositis" gradient bar (using tiles)
  geom_tile(
    data = gradient_data,
    aes(
      x = day,
      y = y_pos,
      alpha = alpha_val,
      fill = "#b3cde3", # Set fill *inside* aes
      height = 0.1
    ),
    color = NA # No border on the tiles
  ) +
  
  # "neutropenia" shaded ribbon
  geom_ribbon(
    data = ribbon_data,
    # Override main 'y' aesthetic, map ymin and ymax
    aes(ymin = ymin, ymax = ymax, y = NULL),
    fill = "#FBBFBF", # Changed to pink
    alpha = 0.7       # Made slightly more opaque
  ) +

  # --- Guidelines ---
  
  # Horizontal line at 0.5
  geom_hline(
    yintercept = 0.5,
    linetype = "dashed",
    color = "grey60"
  ) +
  
  # Vertical lines at day 0 and day 12
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey70"
  ) +
  geom_vline(
    xintercept = 12,
    linetype = "dashed",
    color = "grey70"
  ) +

  # --- Main Neutrophil Line ---
  geom_line(
    color = "#555555", # Matching red
    linewidth = .8
  ) +

  # --- Text Labels ---
  
  annotate(
    "text",
    x = 6, y = 0.85, # Centered between 0 and 12
    label = "nausea, mucositis",
    hjust = 0.5,
    size = 4.5
  ) +
  annotate(
    "text",
    x = 6.5, y = 0.15, # Centered at the new trough
    label = "neutropenia",
    hjust = 0.5,
    size = 4.5,
    color = "#111111"
  ) +

  # --- Scales and Theme ---
  
  # Set axis limits, breaks, and labels
  scale_x_continuous(
    breaks = c(0, 12),
    labels = c("0", "~12"),
    limits = c(-4.5, 19), # Extended to show flat tail
    expand = c(0, 0) # Remove padding
  ) +
  scale_y_continuous(
    breaks = c(0, 0.5),
    limits = c(0, 2),
    expand = c(0, 0) # Remove padding
  ) +
  
  # Add this scale to use the 'alpha_val' column directly
  scale_alpha_identity() +
  
  # Add this scale to use the fill color directly
  scale_fill_identity() +

  # Use unicode for the micron (µ) symbol
  labs(
    x = "Day",
    y = "neutrophils\n[k/\u00b5l blood]"
  ) +
  
  # Clean up the theme
  theme_classic(base_size = 14) +
  theme(
    # Remove all gridlines
    panel.grid = element_blank(),
    # Style axis titles
    axis.title.y = element_text(
      angle = 90,
      size = 12,
      margin = margin(r = 0) # Add space
    ),
    axis.title.x = element_text(
      size = 12,
      margin = margin(t = 0) # Add space
    ),
    # Style axis text
    axis.text = element_text(size = 11, color = "black"),
    # Style axis lines
    axis.line = element_line(linewidth = 0.5, color = "black"),
    aspect.ratio = 1/1.5
    # Add a plot margin
    #plot.margin = margin(20, 20, 20, 20)
  )

# Print the plot
print(bmt_plot)
```

## the final figure 1

```{r}
title <- ggdraw() + 
  draw_label("Fig. 1",fontface = 'bold',x = 0,hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 5))

f1 <-  plot_grid(bmt_plot, diet_hist, stool_hist, 
                 cal_line, faith_line, stool_alpha,
                 align = 'hv',
                 axis = 'tblr',
                 labels = 'auto',
                 #rel_heights = c(1,0.7, 1.5),
                 nrow = 2)

combined <- plot_grid(
  title, f1,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.1, 1)
) +theme(plot.margin = unit(c(1,1,11,1), "cm"))

ggsave('../data/072_F1_parelle_collapse.pdf',
      width = 8.5, height = 11, units = "in", device = 'pdf', plot = combined,
      dpi = 300)  
```

