---
title: "taxUMAP of the combined cohort (Figure 1)"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(patchwork)
library(RColorBrewer)
library(ggrastr)
library(cowplot)
```

# Paths


```{r paths}
# ---- edit this one variable if your data lives elsewhere -----------------
DATA_DIR <- "../data/"   
# --------------------------------------------------------------------------

# Inputs
dtb_file       <- file.path(DATA_DIR, "152_combined_DTB.csv")
color_key_file <- file.path(DATA_DIR, "food_group_color_key_final.csv")
taxonomy_file  <- file.path(DATA_DIR, "final_table_for_writing_out_to_newick.csv")
## the below results come from the snakemake pipeline 
faith <- read_tsv("../data/005_faith_pd/alpha-diversity.tsv") %>%
  rename(sampleid = 1)   # first column is unnamed

# Outputs (written into the same folder as the inputs)
food_relabd_out <- file.path(DATA_DIR, "162_food_code_relative.csv")
food_taxa_out   <- file.path(DATA_DIR, "162_food_taxa.csv")

# Verify everything is in place before we start.
required_inputs <- c(dtb_file, color_key_file, taxonomy_file)
missing <- required_inputs[!file.exists(required_inputs)]
if (length(missing) > 0) {
  stop(
    "Missing required input file(s):\n  ",
    paste(missing, collapse = "\n  "),
    "\nUpdate DATA_DIR in this chunk to point to the folder containing them."
  )
}
```


# Plot defaults

Shared aesthetics used by the downstream Figure 1 panels.

```{r plot-defaults}


umap_pt_size    <- 0.5
axis_text_size  <- 10
axis_title_size <- 10
gray_bg_color <- 'gray98'

# Food-group color key (used by downstream plotting).
color_key <- read_csv(color_key_file, col_types = "ccccc")

dtb <- read_csv(dtb_file)
```

# Stage 1 — Build the per-day relative-abundance matrix

taxUMAP expects rows of compositional (relative) abundances. The
diet-tracker records absolute dehydrated weight per meal item, so for
each `(patient, day)` we sum to a daily total per food code, then divide
by the daily grand total.

```{r build-relabd-matrix}


# 1. Sum dehydrated weight per (patient, day, food code).
fc_table <- dtb %>%
  select(pid, fdrt, Food_code, dehydrated_weight) %>%
  group_by(pid, fdrt, Food_code) %>%
  summarise(total = sum(dehydrated_weight), .groups = "drop") %>%
  mutate(index_column = str_glue("P{pid}_{fdrt}")) %>%
  select(index_column, Food_code, total)

# 2. Daily total weight per (patient, day) — denominator for relative abundance.
fc_total <- fc_table %>%
  group_by(index_column) %>%
  summarise(daily_total = sum(total), .groups = "drop")

# 3. Pivot to wide form: rows = patient-day, columns = food code, values = rel. abd.
fc_df <- fc_table %>%
  left_join(fc_total, by = "index_column") %>%
  mutate(daily_relative = total / daily_total) %>%
  select(index_column, Food_code, daily_relative) %>%
  pivot_wider(
    names_from  = Food_code,
    values_from = daily_relative,
    values_fill = 0
  )

# Sanity check: every row should sum to ~1.
stopifnot(all(abs(rowSums(fc_df[ , -1]) - 1) < 1e-8))

write_csv(fc_df, food_relabd_out)
glue::glue("Wrote {nrow(fc_df)} patient-days x {ncol(fc_df) - 1} food codes -> {food_relabd_out}")
```

# Stage 2 — Build the taxonomy table

> The taxonomy table is expected to resolve higher taxonomic groups for
> each ASV/OTU (here, food code) in the abundance table. The index
> should be the ASV/OTU labels, and columns should be higher taxonomic
> categories ordered from left to right in **decreasing** taxonomic
> hierarchy.

We restrict to food codes that actually appear in the abundance matrix,
and keep four taxonomic levels (L1–L4 → Kingdom / Phylum / Class /
Order).

```{r build-taxonomy}
# Food codes present in the abundance matrix (everything except the index col).
actual_foodids <- setdiff(colnames(fc_df), "index_column")

taxonomy_full <- read_csv(taxonomy_file) %>%
  mutate(FoodID = as.character(FoodID))

food_taxa <- taxonomy_full %>%
  filter(FoodID %in% actual_foodids) %>%
  select(
    ASV     = FoodID,
    Kingdom = L1,
    Phylum  = L2,
    Class   = L3,
    Order   = L4
  )

# Quick coverage check: every food code in the matrix should have taxonomy.
missing_taxa <- setdiff(actual_foodids, food_taxa$ASV)
if (length(missing_taxa) > 0) {
  warning(
    "No taxonomy found for ", length(missing_taxa),
    " food code(s); they will be dropped by taxUMAP. First few: ",
    paste(head(missing_taxa), collapse = ", ")
  )
}

write_csv(food_taxa, food_taxa_out)

# Top-two-level summary (handy reference for figure legends).
top2level <- food_taxa %>% distinct(Kingdom, Phylum)
```

# Stage 3 — Run taxUMAP

```{bash}
# run the below command in terminal where the conda env is activated in this dir
run_taxumap.py -t ../data/162_food_taxa.csv -m ../data/162_food_code_relative.csv --agg_levels Kingdom/Phylum/Class/Order -n 173 --weights 0.01/2/10/10  -o ../data
```

# Stage 4 — Colorcode the taxUMAP results 

## summary table

```{r}
# 1. load the taxUMAP results that were just generated
taxumap <- read_csv("../data/taxumap_embedding.csv")

# ---- 2. Per (pid, fdrt): top food group by dehydrated weight + total calories ----
# Extract the first digit of Food_code => fgrp1
dtb_fg <- dtb %>%
  mutate(fgrp1 = str_sub(as.character(Food_code), 1, 1))

# Sum dehydrated weight per food group within each (pid, fdrt)
fg_weights <- dtb_fg %>%
  group_by(pid, fdrt, fgrp1) %>%
  summarise(fg_dehydrated = sum(dehydrated_weight, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(pid, fdrt) %>%
  mutate(fg_proportion = fg_dehydrated / sum(fg_dehydrated)) %>%
  ungroup()

# Pick the dominant food group per (pid, fdrt)
top_fg <- fg_weights %>%
  group_by(pid, fdrt) %>%
  slice_max(fg_proportion, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(color_key, by = "fgrp1") %>%
  select(pid, fdrt,
         top_fgrp1   = fgrp1,
         top_fg_name = shortname,
         top_fg_desc = fdesc,
         top_fg_color = color,
         top_fg_proportion = fg_proportion)

# Total calories per (pid, fdrt)
cal_totals <- dtb %>%
  group_by(pid, fdrt) %>%
  summarise(total_calories = sum(Calories_kcal, na.rm = TRUE),
            .groups = "drop")

# Combine
summary_tbl <- top_fg %>%
  left_join(cal_totals, by = c("pid", "fdrt"))

# ---- 3. Join with faith PD ----
# Build the matching sample id: "{pid}d{fdrt}" (e.g. P100 + d + -1 -> P100d-1)
summary_tbl <- summary_tbl %>%
  mutate(sampleid = paste0(pid, "d", fdrt)) %>%
  left_join(faith, by = "sampleid") %>%
  relocate(sampleid, .after = fdrt)

# ---- 4. Join TaxUMAP coords ----
# index_column looks like "PP1_-6"  -->  "P" + pid + "_" + fdrt
plot_tbl <- summary_tbl %>%
  mutate(index_column = paste0("P", pid, "_", fdrt)) %>%
  inner_join(taxumap, by = "index_column")
```
## old code 

## e: day relative to transplant

```{r}
# cuztomizing the times bins of the drt
cutoff = 0 # Day 0 is conveneint but this can be varied
n_groups_pre = 3;  # all HCT days <0 will be split into 3 bins; can vary this number to play
n_groups_post = 6; # all HCT days >0 will be split into 6 bins; can vary this number to play

# split the fdrt into groups:
umap_time <- plot_tbl %>% 
  select(taxumap1, taxumap2, fdrt)  

times <- umap_time %>% select( fdrt)  

splits <- times %>% 
  split(.$fdrt <= 0)
pre <- splits %>% 
  pluck('TRUE') %>% 
  mutate(bin = cut_number(fdrt, n_groups_pre))
post <- splits %>% 
  pluck('FALSE') %>% 
  mutate(bin = cut_number(fdrt, n_groups_post)) 
both <- bind_rows(pre, post) %>% 
  distinct(fdrt, .keep_all = T)

umap_time_df <- umap_time %>% 
  left_join(both)

umap_drt_high_to_low_bin <- umap_time_df %>% 
  arrange(desc(fdrt)) %>% 
    ggplot() +
  rasterise(geom_point(aes(x = taxumap1, y = taxumap2, color = bin), alpha = 1,size = umap_pt_size , shape = 16), dpi = 300) +
   scale_color_manual(values = RColorBrewer::brewer.pal(9,"Spectral"))  +
  labs(title = 'Day relative to\ntransplant') +
   theme_classic() +
   theme(line = element_blank(),
         legend.position = "none",
         legend.title = element_blank(),
         axis.title=element_blank(),
         axis.text= element_blank(),
         axis.ticks = element_blank(),
              panel.background = element_rect(fill=gray_bg_color), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent') ,
        aspect.ratio=1)

umap_drt_high_to_low_bin
```

## f: caloric intake

```{r} 
out2_caloric <- plot_tbl %>% 
  mutate(transform_cal = sqrt(total_calories)) %>% 
    ggplot() +
  rasterise(geom_point(aes(x = taxumap1, y = taxumap2, color = transform_cal), alpha = 1,size = umap_pt_size , shape = 16), dpi = 300) +
  paletteer::scale_color_paletteer_c("viridis::plasma") +
  labs(title = 'Caloric intake\n(*1000 Kcal)') +
  theme_classic() +
  #lemon::coord_capped_cart(bottom = 'both', left = 'both') +
  theme(line = element_blank(),
        legend.position = "none",
        legend.title = element_blank(),
        axis.title=element_blank(),
        axis.text= element_blank(),
        axis.ticks = element_blank(),
         panel.background = element_rect(fill=gray_bg_color), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent') ,
        aspect.ratio=1)  
```

## g: dominant food group

```{r}
# Build the color lookup from the data itself
color_key <- plot_tbl |> 
  distinct(top_fg_name, top_fg_color) |> 
  deframe()  # turns two-col tibble into a named vector

out2_dominant <- plot_tbl %>% 
  ggplot() +
  rasterise(geom_point(aes(x = taxumap1, y = taxumap2, color = top_fg_name), alpha = 1,size = umap_pt_size , shape = 16), dpi = 300) +
  labs(x = 'TaxUMAP1', y = 'TaxUMAP2', title = 'Most consumed food group') +
  scale_color_manual(values = color_key) +
  theme_classic() +
  #lemon::coord_capped_cart(bottom = 'both', left = 'both') +
  theme(line = element_blank(),
        legend.position = "none",
        legend.title = element_blank(),
        axis.title=element_blank(),
        axis.text= element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_rect(fill=gray_bg_color), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent') , aspect.ratio=1) +
  guides(colour = guide_legend(override.aes = list(alpha=1)))
out2_dominant
```

## h: faith alpha diversity

```{r}
umap_faith <- plot_tbl %>% 
  mutate(log10faith = log10(faith_pd + 1)) %>% 
    ggplot() +
  rasterise(geom_point(aes(x = taxumap1, y = taxumap2, color = faith_pd), alpha = 1,size = umap_pt_size , shape = 16), dpi = 300) +
  viridis::scale_color_viridis() +
  labs(title = 'Diet a diversity\n(*1000)') +
  theme_classic() +
  theme(line = element_blank(),
        legend.title = element_blank(),
        axis.title=element_blank(),
        axis.text= element_blank(),
        axis.ticks = element_blank(),
        legend.position = 'none', 
        aspect.ratio=1,
                 panel.background = element_rect(fill=gray_bg_color), #transparent panel bg
    plot.background = element_rect(fill='transparent', color=NA), #transparent plot bg
    panel.grid.major = element_blank(), #remove major gridlines
    panel.grid.minor = element_blank(), #remove minor gridlines
    legend.background = element_rect(fill='transparent'), #transparent legend bg
    legend.box.background = element_rect(fill='transparent') 
        )
```


## assemble taxUMAPs

```{r}
middle_maps <- plot_grid(umap_drt_high_to_low_bin,  out2_caloric, out2_dominant, umap_faith, 
          nrow = 1, hjust = 0,
          align = 'hv',
          labels = c( 'e', "f", 'g','h'),
          axis = 'tblr')

ggsave('../data/F1efgh_162_four_taxUMAPs.pdf', plot = middle_maps)
```


```{r}
# # ---- Panel-e binning parameters ----
# cutoff         <- 0
# n_groups_pre   <- 3
# n_groups_post  <- 6
# 
# pre_breaks  <- quantile(plot_tbl$fdrt[plot_tbl$fdrt <  cutoff],
#                         probs = seq(0, 1, length.out = n_groups_pre  + 1),
#                         na.rm = TRUE)
# post_breaks <- quantile(plot_tbl$fdrt[plot_tbl$fdrt >= cutoff],
#                         probs = seq(0, 1, length.out = n_groups_post + 1),
#                         na.rm = TRUE)
# 
# # Force the boundary at the cutoff so pre/post share an edge -> no orphan bin
# pre_breaks[length(pre_breaks)] <- cutoff
# fdrt_breaks <- unique(c(pre_breaks, post_breaks))
# 
# plot_tbl <- plot_tbl %>%
#   mutate(fdrt_bin = cut(fdrt, breaks = fdrt_breaks,
#                         include.lowest = TRUE, dig.lab = 3))
# 
# # Derive bin count from the actual factor — not assumed
# n_bins <- nlevels(plot_tbl$fdrt_bin)
# spectral_cols <- rev(colorRampPalette(brewer.pal(11, "Spectral"))(n_bins))
# 
# # ---- Common theme: legend OUTSIDE each panel ----
# theme_umap <- theme_bw(base_size = 11) +
#   theme(
#     plot.title       = element_text(hjust = 0.5, face = "bold"),
#     panel.grid       = element_blank(),
#     legend.position  = "right",
#     legend.key.size  = unit(0.4, "cm"),
#     legend.title     = element_text(size = 9),
#     legend.text      = element_text(size = 8),
#     axis.title       = element_text(size = 9),
#     axis.text        = element_blank(),
#     axis.ticks       = element_blank()
#   )
# 
# base_aes <- aes(x = taxumap1, y = taxumap2)
# 
# # ---- e. Day relative to transplant (binned Spectral) ----
# p_e <- ggplot(plot_tbl, base_aes) +
#   geom_point(aes(color = fdrt_bin), size = 0.8, alpha = 0.85) +
#   scale_color_manual(values = spectral_cols, name = NULL, drop = FALSE) +
#   labs(title = "Day relative\nto transplant",
#        x = "TaxUMAP1", y = "TaxUMAP2") +
#   theme_umap
# 
# # ---- f. Caloric intake ----
# p_f <- ggplot(plot_tbl, base_aes) +
#   geom_point(aes(color = total_calories / 1000), size = 0.8, alpha = 0.85) +
#   scale_color_viridis_c(option = "plasma") +
#   labs(title = "Caloric intake",
#        x = "TaxUMAP1", y = "TaxUMAP2",
#        color = expression(x~10^3~Kcal)) +
#   theme_umap
# 
# # ---- g. Most consumed food group ----
# fg_colors <- setNames(color_key$color, color_key$shortname)
# 
# p_g <- ggplot(plot_tbl, base_aes) +
#   geom_point(aes(color = top_fg_name), size = 0.8, alpha = 0.9) +
#   scale_color_manual(values = fg_colors, name = NULL) +
#   labs(title = "Most consumed\nfood group",
#        x = "TaxUMAP1", y = "TaxUMAP2") +
#   theme_umap
# 
# # ---- h. Diet alpha-diversity ----
# p_h <- ggplot(plot_tbl, base_aes) +
#   geom_point(aes(color = faith_pd / 1000), size = 0.8, alpha = 0.85) +
#   scale_color_viridis_c(option = "viridis") +
#   labs(title = expression(Diet~alpha*"-diversity"),
#        x = "TaxUMAP1", y = "TaxUMAP2",
#        color = expression(""%*%10^3)) +
#   theme_umap
# 
# # ---- Combine ----
# fig <- (p_e | p_f | p_g | p_h) +
#   plot_annotation(tag_levels = list(c("e", "f", "g", "h"))) &
  # theme(plot.tag = element_text(face = "bold", size = 14))
```



