---
title: "taxUMAP of the combined cohort (Figure 1)"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Paths


```{r paths}
# ---- edit this one variable if your data lives elsewhere -----------------
DATA_DIR <- "../data/"   
# --------------------------------------------------------------------------

# Inputs
dtb_file       <- file.path(DATA_DIR, "152_combined_DTB.csv")
color_key_file <- file.path(DATA_DIR, "food_group_color_key_final.csv")
taxonomy_file  <- file.path(DATA_DIR, "final_table_for_writing_out_to_newick.csv")
## the below results come from the snakemake pipeline 
faith <- read_tsv("../data/005_faith_pd/alpha-diversity.tsv") %>%
  rename(sampleid = 1)   # first column is unnamed

# Outputs (written into the same folder as the inputs)
food_relabd_out <- file.path(DATA_DIR, "162_food_code_relative.csv")
food_taxa_out   <- file.path(DATA_DIR, "162_food_taxa.csv")

# Verify everything is in place before we start.
required_inputs <- c(dtb_file, color_key_file, taxonomy_file)
missing <- required_inputs[!file.exists(required_inputs)]
if (length(missing) > 0) {
  stop(
    "Missing required input file(s):\n  ",
    paste(missing, collapse = "\n  "),
    "\nUpdate DATA_DIR in this chunk to point to the folder containing them."
  )
}
```


# Plot defaults

Shared aesthetics used by the downstream Figure 1 panels.

```{r plot-defaults}
library(tidyverse)
library(patchwork)

umap_pt_size    <- 0.5
axis_text_size  <- 10
axis_title_size <- 10

# Food-group color key (used by downstream plotting).
color_key <- read_csv(color_key_file, col_types = "ccccc")
```

# Stage 1 — Build the per-day relative-abundance matrix

taxUMAP expects rows of compositional (relative) abundances. The
diet-tracker records absolute dehydrated weight per meal item, so for
each `(patient, day)` we sum to a daily total per food code, then divide
by the daily grand total.

```{r build-relabd-matrix}
dtb <- read_csv(dtb_file)

# 1. Sum dehydrated weight per (patient, day, food code).
fc_table <- dtb %>%
  select(pid, fdrt, Food_code, dehydrated_weight) %>%
  group_by(pid, fdrt, Food_code) %>%
  summarise(total = sum(dehydrated_weight), .groups = "drop") %>%
  mutate(index_column = str_glue("P{pid}_{fdrt}")) %>%
  select(index_column, Food_code, total)

# 2. Daily total weight per (patient, day) — denominator for relative abundance.
fc_total <- fc_table %>%
  group_by(index_column) %>%
  summarise(daily_total = sum(total), .groups = "drop")

# 3. Pivot to wide form: rows = patient-day, columns = food code, values = rel. abd.
fc_df <- fc_table %>%
  left_join(fc_total, by = "index_column") %>%
  mutate(daily_relative = total / daily_total) %>%
  select(index_column, Food_code, daily_relative) %>%
  pivot_wider(
    names_from  = Food_code,
    values_from = daily_relative,
    values_fill = 0
  )

# Sanity check: every row should sum to ~1.
stopifnot(all(abs(rowSums(fc_df[ , -1]) - 1) < 1e-8))

write_csv(fc_df, food_relabd_out)
glue::glue("Wrote {nrow(fc_df)} patient-days x {ncol(fc_df) - 1} food codes -> {food_relabd_out}")
```

# Stage 2 — Build the taxonomy table

> The taxonomy table is expected to resolve higher taxonomic groups for
> each ASV/OTU (here, food code) in the abundance table. The index
> should be the ASV/OTU labels, and columns should be higher taxonomic
> categories ordered from left to right in **decreasing** taxonomic
> hierarchy.

We restrict to food codes that actually appear in the abundance matrix,
and keep four taxonomic levels (L1–L4 → Kingdom / Phylum / Class /
Order).

```{r build-taxonomy}
# Food codes present in the abundance matrix (everything except the index col).
actual_foodids <- setdiff(colnames(fc_df), "index_column")

taxonomy_full <- read_csv(taxonomy_file) %>%
  mutate(FoodID = as.character(FoodID))

food_taxa <- taxonomy_full %>%
  filter(FoodID %in% actual_foodids) %>%
  select(
    ASV     = FoodID,
    Kingdom = L1,
    Phylum  = L2,
    Class   = L3,
    Order   = L4
  )

# Quick coverage check: every food code in the matrix should have taxonomy.
missing_taxa <- setdiff(actual_foodids, food_taxa$ASV)
if (length(missing_taxa) > 0) {
  warning(
    "No taxonomy found for ", length(missing_taxa),
    " food code(s); they will be dropped by taxUMAP. First few: ",
    paste(head(missing_taxa), collapse = ", ")
  )
}

write_csv(food_taxa, food_taxa_out)

# Top-two-level summary (handy reference for figure legends).
top2level <- food_taxa %>% distinct(Kingdom, Phylum)
```

# Stage 3 — Run taxUMAP

```{bash}
# run the below command in terminal where the conda env is activated in this dir
run_taxumap.py -t ../data/162_food_taxa.csv -m ../data/162_food_code_relative.csv --agg_levels Kingdom/Phylum/Class/Order -n 173 --weights 0.01/2/10/10  -o ../data
```

# Stage 4 — Colorcode the taxUMAP results 

## summary table

```{r}
# 1. load the taxUMAP results that were just generated
taxumap <- read_csv("../data/taxumap_embedding.csv")

# ---- 2. Per (pid, fdrt): top food group by dehydrated weight + total calories ----
# Extract the first digit of Food_code => fgrp1
dtb_fg <- dtb %>%
  mutate(fgrp1 = str_sub(as.character(Food_code), 1, 1))

# Sum dehydrated weight per food group within each (pid, fdrt)
fg_weights <- dtb_fg %>%
  group_by(pid, fdrt, fgrp1) %>%
  summarise(fg_dehydrated = sum(dehydrated_weight, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(pid, fdrt) %>%
  mutate(fg_proportion = fg_dehydrated / sum(fg_dehydrated)) %>%
  ungroup()

# Pick the dominant food group per (pid, fdrt)
top_fg <- fg_weights %>%
  group_by(pid, fdrt) %>%
  slice_max(fg_proportion, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(color_key, by = "fgrp1") %>%
  select(pid, fdrt,
         top_fgrp1   = fgrp1,
         top_fg_name = shortname,
         top_fg_desc = fdesc,
         top_fg_color = color,
         top_fg_proportion = fg_proportion)

# Total calories per (pid, fdrt)
cal_totals <- dtb %>%
  group_by(pid, fdrt) %>%
  summarise(total_calories = sum(Calories_kcal, na.rm = TRUE),
            .groups = "drop")

# Combine
summary_tbl <- top_fg %>%
  left_join(cal_totals, by = c("pid", "fdrt"))

# ---- 3. Join with faith PD ----
# Build the matching sample id: "{pid}d{fdrt}" (e.g. P100 + d + -1 -> P100d-1)
summary_tbl <- summary_tbl %>%
  mutate(sampleid = paste0(pid, "d", fdrt)) %>%
  left_join(faith, by = "sampleid") %>%
  relocate(sampleid, .after = fdrt)
```


```{r}
# ---- 4. Join TaxUMAP coords ----
# index_column looks like "PP1_-6"  -->  "P" + pid + "_" + fdrt
plot_tbl <- summary_tbl %>%
  mutate(index_column = paste0("P", pid, "_", fdrt)) %>%
  inner_join(taxumap, by = "index_column")

# Named vector of food-group colors for ggplot scale
fg_colors <- setNames(color_key$color, color_key$shortname)

# ---- 5. Common theme ----
theme_umap <- theme_void(base_size = 11) +
  theme(plot.title    = element_text(hjust = 0.5, face = "bold"),
        legend.position = c(0.85, 0.25),
        legend.key.size = unit(0.35, "cm"),
        legend.title  = element_text(size = 8),
        legend.text   = element_text(size = 7),
        axis.title    = element_text(size = 8),
        axis.title.y  = element_text(angle = 90))

base_aes <- aes(x = taxumap1, y = taxumap2)

# ---- 6. Panel e: day relative to transplant ----
p_e <- ggplot(plot_tbl, base_aes) +
  geom_point(aes(color = fdrt), size = 0.8, alpha = 0.85) +
  scale_color_distiller(palette = "Spectral",
                        breaks = c(-10, 0, 7, 15, 49)) +
  labs(title = "Day relative\nto transplant",
       x = "TaxUMAP1", y = "TaxUMAP2", color = NULL) +
  theme_umap

# ---- 7. Panel f: caloric intake (x 10^3 kcal) ----
p_f <- ggplot(plot_tbl, base_aes) +
  geom_point(aes(color = total_calories / 1000), size = 0.8, alpha = 0.85) +
  scale_color_viridis_c(option = "plasma") +
  labs(title = "Caloric intake",
       x = "TaxUMAP1", y = "TaxUMAP2",
       color = expression(x~10^3~Kcal)) +
  theme_umap

# ---- 8. Panel g: most consumed food group ----
p_g <- ggplot(plot_tbl, base_aes) +
  geom_point(aes(color = top_fg_name), size = 0.8, alpha = 0.9) +
  scale_color_manual(values = fg_colors, name = NULL) +
  labs(title = "Most consumed\nfood group",
       x = "TaxUMAP1", y = "TaxUMAP2") +
  theme_umap +
  theme(legend.position = "right",
        legend.key.size = unit(0.4, "cm"))

# ---- 9. Panel h: diet alpha-diversity (x 10^3) ----
p_h <- ggplot(plot_tbl, base_aes) +
  geom_point(aes(color = faith_pd / 1000), size = 0.8, alpha = 0.85) +
  scale_color_viridis_c(option = "viridis") +
  labs(title = expression(Diet~alpha*"-diversity"),
       x = "TaxUMAP1", y = "TaxUMAP2",
       color = expression(""%*%10^3)) +
  theme_umap

# ---- 10. Combine ----
fig <- (p_e | p_f | p_g | p_h) +
  plot_annotation(tag_levels = list(c("e", "f", "g", "h"))) &
  theme(plot.tag = element_text(face = "bold", size = 14))

ggsave("../data/162_taxumap_4panel.pdf", fig, width = 10, height = 4)

fig
```
