---
title: "taxUMAP of the combined cohort (Figure 1)"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Paths


```{r paths}
# ---- edit this one variable if your data lives elsewhere -----------------
DATA_DIR <- "../data/"   
# --------------------------------------------------------------------------

# Inputs
dtb_file       <- file.path(DATA_DIR, "152_combined_DTB.csv")
color_key_file <- file.path(DATA_DIR, "food_group_color_key_final.csv")
taxonomy_file  <- file.path(DATA_DIR, "final_table_for_writing_out_to_newick.csv")

# Outputs (written into the same folder as the inputs)
food_relabd_out <- file.path(DATA_DIR, "162_food_code_relative.csv")
food_taxa_out   <- file.path(DATA_DIR, "162_food_taxa.csv")

# Verify everything is in place before we start.
required_inputs <- c(dtb_file, color_key_file, taxonomy_file)
missing <- required_inputs[!file.exists(required_inputs)]
if (length(missing) > 0) {
  stop(
    "Missing required input file(s):\n  ",
    paste(missing, collapse = "\n  "),
    "\nUpdate DATA_DIR in this chunk to point to the folder containing them."
  )
}
```


# Plot defaults

Shared aesthetics used by the downstream Figure 1 panels.

```{r plot-defaults}
library(tidyverse)

umap_pt_size    <- 0.5
axis_text_size  <- 10
axis_title_size <- 10

# Food-group color key (used by downstream plotting).
color_key <- read_csv(color_key_file, col_types = "ccccc")
```

# Stage 1 — Build the per-day relative-abundance matrix

taxUMAP expects rows of compositional (relative) abundances. The
diet-tracker records absolute dehydrated weight per meal item, so for
each `(patient, day)` we sum to a daily total per food code, then divide
by the daily grand total.

```{r build-relabd-matrix}
dtb <- read_csv(dtb_file)

# 1. Sum dehydrated weight per (patient, day, food code).
fc_table <- dtb %>%
  select(pid, fdrt, Food_code, dehydrated_weight) %>%
  group_by(pid, fdrt, Food_code) %>%
  summarise(total = sum(dehydrated_weight), .groups = "drop") %>%
  mutate(index_column = str_glue("P{pid}_{fdrt}")) %>%
  select(index_column, Food_code, total)

# 2. Daily total weight per (patient, day) — denominator for relative abundance.
fc_total <- fc_table %>%
  group_by(index_column) %>%
  summarise(daily_total = sum(total), .groups = "drop")

# 3. Pivot to wide form: rows = patient-day, columns = food code, values = rel. abd.
fc_df <- fc_table %>%
  left_join(fc_total, by = "index_column") %>%
  mutate(daily_relative = total / daily_total) %>%
  select(index_column, Food_code, daily_relative) %>%
  pivot_wider(
    names_from  = Food_code,
    values_from = daily_relative,
    values_fill = 0
  )

# Sanity check: every row should sum to ~1.
stopifnot(all(abs(rowSums(fc_df[ , -1]) - 1) < 1e-8))

write_csv(fc_df, food_relabd_out)
glue::glue("Wrote {nrow(fc_df)} patient-days x {ncol(fc_df) - 1} food codes -> {food_relabd_out}")
```

# Stage 2 — Build the taxonomy table

> The taxonomy table is expected to resolve higher taxonomic groups for
> each ASV/OTU (here, food code) in the abundance table. The index
> should be the ASV/OTU labels, and columns should be higher taxonomic
> categories ordered from left to right in **decreasing** taxonomic
> hierarchy.

We restrict to food codes that actually appear in the abundance matrix,
and keep four taxonomic levels (L1–L4 → Kingdom / Phylum / Class /
Order).

```{r build-taxonomy}
# Food codes present in the abundance matrix (everything except the index col).
actual_foodids <- setdiff(colnames(fc_df), "index_column")

taxonomy_full <- read_csv(taxonomy_file) %>%
  mutate(FoodID = as.character(FoodID))

food_taxa <- taxonomy_full %>%
  filter(FoodID %in% actual_foodids) %>%
  select(
    ASV     = FoodID,
    Kingdom = L1,
    Phylum  = L2,
    Class   = L3,
    Order   = L4
  )

# Quick coverage check: every food code in the matrix should have taxonomy.
missing_taxa <- setdiff(actual_foodids, food_taxa$ASV)
if (length(missing_taxa) > 0) {
  warning(
    "No taxonomy found for ", length(missing_taxa),
    " food code(s); they will be dropped by taxUMAP. First few: ",
    paste(head(missing_taxa), collapse = ", ")
  )
}

write_csv(food_taxa, food_taxa_out)

# Top-two-level summary (handy reference for figure legends).
top2level <- food_taxa %>% distinct(Kingdom, Phylum)
```

# Stage 3 — Run taxUMAP

```{bash}
# run the below command in terminal where the conda env is activated in this dir
run_taxumap.py -t ../data/162_food_taxa.csv -m ../data/162_food_code_relative.csv --agg_levels Kingdom/Phylum/Class/Order -n 173 --weights 0.01/2/10/10  -o ../data
```



