---
title: "Supplementary for the patient intercept figure"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(cowplot)
library(tidybayes)
```

```{r }
# make 
post_all <- read_csv('../data/171_div_model_fg_post_interaction.csv')

# make  a forest  plot of the pid intercepts
pid_forest <- post_all %>% 
  gather() %>% 
  filter(str_detect(key, 'r_pid')) %>% 
  separate(key, into = c('rm','more'), sep = 'r_pid\\[') %>% 
  separate(more, into = c('pid','rm2'), sep = ',Intercept\\]') 


# find the median value of the coeff and sort by that in an asc order
pid_order <- pid_forest %>% 
  group_by(pid) %>% 
  summarise(medianvalue = median(value)) %>% 
  arrange(medianvalue) %>% pull(pid)
```


```{r }
pid_intercept <- pid_forest %>% 
  mutate(pid = factor(pid, levels = pid_order)) %>% 
  ggplot(aes(x = value, y = pid)) +
  stat_pointinterval(.width =.95, point_size = 1) +
  geom_vline(xintercept = 0, col = 'blue') +
  labs(x = 'Patient-Specific Deviation from Average Microbiome log(Diversity)',
       y = '', 
       title = '') +
  theme_classic() +
  theme(legend.position = 'none') +
  theme(axis.text=element_text(size=10), 
        panel.background = element_rect(fill = "gray96",colour = "gray96",size = 0.5, linetype = "solid"),
        axis.text.y=element_text(size=6))


title <- ggdraw() + 
  draw_label(
    "Fig. S5",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )

combined <- plot_grid(
  title, pid_intercept,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.1, 6)
)+theme(plot.margin = unit(c(1,0,0,1), "cm"))
  
ggsave('../results//S05_pid_intercepts_forest_179.pdf',  
     width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)
```

Interpretation:

Deviation from the Average: The value of r_pid[P1,Intercept] tells you how much the baseline outcome for patient P1 deviates from the overall average outcome predicted by the fixed effects in your model.
Positive/Negative Value:
A positive value means that patient P1 tends to have a higher outcome value than the average, even after accounting for other predictors in the model.
A negative value means that patient P1 tends to have a lower outcome value than the average.
Magnitude: The absolute value of the random intercept indicates the magnitude of the deviation from the average. A larger absolute value means a greater difference from the average.