---
title: "The mice 16s sequencing data from diet experiments"
output: html_document
date: "2024-07-18"
---

```{r}
library(tidyverse)
library(ggpubr)
library(janitor)
library(kableExtra)
library(knitr)
library(Maaslin2)
library(vegan)
library(ggrepel)
library(scales)
library(readxl)
library(cowplot)
library(patchwork)
```

```{r}
healthyall <- read_csv('../data/204_mice_diet_healthy_all.csv')

meta <- healthyall %>%
  select(-Taxon, -relab) %>% 
  distinct() 

meta %>% 
  count(experiment_no, abx_treatment, diet_treatment, day) %>% 
  summarise(min_ = min(n), max_ = max(n))
```

# alpha diversity 

## d3 VS d0

```{r}
# what I really care is the change from day 0 to day 3 
alpha_change <-  meta %>% 
  mutate(day = factor(day), experiment_no = factor(experiment_no)) %>% 
  arrange(abx_treatment,desc(diet_treatment),  day) %>% 
  mutate(xvar = str_glue('{abx_treatment}__{diet_treatment}__{day}')) %>% 
  mutate(grp = str_glue('{abx_treatment}__{diet_treatment}')) %>% 
  mutate(grp = factor(grp, levels = c('PBS__vehicle','PBS__sucrose','biapenem__vehicle','biapenem__sucrose'))) %>%    
  ggboxplot(x = 'xvar', y = 'simpson_reciprocal',add = 'jitter', 
             xlab = '', ylab = 'alpha diversity(simpson reciprocal)', add.params = list(shape ='experiment_no',alpha = 0.8), width = 0.6,color  = 'grp')+
   scale_color_manual(values = c('gray76','#ffbcdc','gray32','deeppink2')) +
  stat_compare_means(comparisons = list(
    c('biapenem__vehicle__0','biapenem__vehicle__3'), 
    c('biapenem__sucrose__0','biapenem__sucrose__3'),
    c('PBS__vehicle__0','PBS__vehicle__3'),
    c('PBS__sucrose__0','PBS__sucrose__3')
    ),paired = T,
    label= "p.signif", method= 'wilcox.test',tip.length = 0.04,exact=T,correct=T) +
  scale_x_discrete(labels=rep(c(0,3,6),4)) + 
  scale_y_sqrt() +
  theme_light() +
  theme(axis.text =  element_text(size = 10, angle = 0, hjust = 1),axis.title=element_text(size=10),legend.position = 'none',aspect.ratio=1/1.3)
alpha_change      
```

```{r}
# it would be good to have a metric of the effect size of the alpha reduction because of the abx treatment
  # the difference of the median in the two groups
median_d3 <- meta %>% filter(day == 3) %>% 
  group_by(abx_treatment, diet_treatment) %>% 
  summarise(median_trt_d3 = median(simpson_reciprocal)) %>% 
  ungroup()

median_pbs_d3 <- median_d3 %>% 
  filter(abx_treatment == 'PBS' & diet_treatment == 'vehicle') %>% 
  pull(median_trt_d3)

median_abx_d3 <- median_d3 %>% 
  filter(abx_treatment == 'biapenem' & diet_treatment == 'vehicle') %>% 
  pull(median_trt_d3)

# the difference 
median_pbs_d3 - median_abx_d3
```


sucrose-only doesn’t reduce diversity on day 3, compared with vehicle, but biapenem-only does.  However, sucrose doesn’t further reduce diversity beyond the biapenem effect on day 3. 


```{r}
# what we actually care is the difference between day 3 and day 0 in the two groups
alpha_wide <- meta %>% 
  select(experiment_no:day, mouse_no,abx_treatment:simpson_reciprocal) %>% 
  mutate(day = str_glue('d{day}')) %>% 
  spread('day','simpson_reciprocal') %>% 
  # mutate(delta_0_3 = d3 - d0,
  #        delta_0_6 = d6 - d0) %>% 
  # gather('grp','deltas', delta_0_3:delta_0_6)
  mutate(fc_0_3 = d3 / d0,
         fc_0_6 = d6 / d0) %>% 
  gather('grp','fc', fc_0_3:fc_0_6)
```
sucrose_only doesn't have much effects. abx_only has a much bigger reduction in the alpha diversity after abx trt


# beta diversity 

```{r}
# the enterococccus relab
g_entero <- healthyall %>% filter(str_detect(Taxon, 'Enterococc')) %>% 
  group_by(sampleid) %>% 
  summarise(enterococcus = sum(relab))

# calculate my own bc distance using the subsetted samples

relab_df <- healthyall %>% 
  select(Taxon, sampleid, relab) %>% 
  spread(key = 'Taxon', value = 'relab', fill = 0) %>% 
  column_to_rownames('sampleid')

dist_ <- vegdist(relab_df, method = 'bray')
eigen <- cmdscale(dist_, eig = T)$eig
percent_var <- signif(eigen/sum(eigen), 3)*100
bc <- cmdscale(dist_, k = 2)
beta_all <- bc %>%
  as.data.frame() %>%
  rownames_to_column('sampleid')  %>% 
  full_join(meta) %>% 
  inner_join(g_entero) %>% 
  mutate(treatment = str_glue('{abx_treatment}_{diet_treatment}')) %>% 
  mutate(treatment = factor(treatment, levels = c('PBS_vehicle','PBS_sucrose','biapenem_vehicle','biapenem_sucrose')))
```

## PERMONOVA for global difference in composition

```{r}
#lump together two experiments
#and simply compare biapenem_vehicle (dark grey) vs biapenem_sucrose(dark red), because that is the key comparison.
#and I would do a separate permanova test on day 3 from the test on day 6
# day 3
relab_df <- healthyall %>% 
  select(Taxon, sampleid, relab, abx_treatment, diet_treatment, experiment_no, day) %>%
  spread(key = 'Taxon', value = 'relab', fill = 0) %>% 
  filter(abx_treatment == 'biapenem' & day != 0)  %>% 
  column_to_rownames('sampleid') %>% 
  split(.$day) %>% 
  map(function(df){
    dist_ <- vegdist(df %>% 
  select(-c(abx_treatment, diet_treatment, experiment_no, day) ), method = 'bray')
    stool_dist_df <- dist_ %>%
          as.matrix() %>%
          as.data.frame() %>%
          rownames_to_column('sampleid') %>%
          inner_join(healthyall %>% distinct(sampleid, abx_treatment, diet_treatment, experiment_no, day))
    
        stool_matrix <- stool_dist_df %>%
          select(-c(abx_treatment, diet_treatment, day)) %>%
          column_to_rownames('sampleid') %>% as.matrix()
    
    # Run PERMANOVA
    adonis2(stool_matrix ~ stool_dist_df$diet_treatment , permutations = 999, parallel = 10 )
  })


relab_df

```

```{r}
relab_df_2 <- healthyall %>% 
  select(Taxon, sampleid, relab, abx_treatment, diet_treatment, experiment_no, day) %>%
  spread(key = 'Taxon', value = 'relab', fill = 0) %>% 
  filter(abx_treatment == 'biapenem' & day != 0)  %>% 
  column_to_rownames('sampleid') %>% 
  split(.$day) %>% 
  map(function(df){
    dist_ <- vegdist(df %>% 
  select(-c(abx_treatment, diet_treatment, experiment_no, day) ), method = 'bray')
    stool_dist_df <- dist_ %>%
          as.matrix() %>%
          as.data.frame() %>%
          rownames_to_column('sampleid') %>%
          inner_join(healthyall %>% distinct(sampleid, abx_treatment, diet_treatment, experiment_no, day))
    
        stool_matrix <- stool_dist_df %>%
          select(-c(abx_treatment, diet_treatment, day)) %>%
          column_to_rownames('sampleid') %>% as.matrix()
    
    # Run PERMANOVA
    adonis2(stool_matrix ~ stool_dist_df$diet_treatment + stool_dist_df$experiment_no, permutations = 999, parallel = 10 )
  })


relab_df_2
```


## PCoA plot

```{r}
# color by abx        
beta_4 <- beta_all %>% 
  mutate(day = factor(day), experiment_no = factor(experiment_no)) %>% 
  arrange(abx_treatment,desc(diet_treatment),  day) %>% 
  mutate(xvar = str_glue('{abx_treatment}__{diet_treatment}__{day}')) %>% 
  mutate(grp = str_glue('{abx_treatment}__{diet_treatment}')) %>% 
  mutate(grp = factor(grp, levels = c('PBS__vehicle','PBS__sucrose','biapenem__vehicle','biapenem__sucrose'))) %>%
  mutate(experiment_no = str_glue('Expe. {experiment_no}'),
         day  = str_glue('D{day}')) %>% 
  ggscatter(x = 'V1', y = 'V2', color = 'grp', shape = 'experiment_no', alpha = 0.8, ) +
  scale_color_manual(values = c('gray76','#ffbcdc','gray32','deeppink2'))  +
  facet_grid( .~  day) +
  theme_light() +
  xlab(paste0("PC 1 [",percent_var[1],"%]")) +
  ylab(paste0("PC 2 [",percent_var[2],"%]")) +
  theme_light() +
  theme(aspect.ratio=1, legend.position = 'bottom',
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 
beta_4  
beta_4_plot <- beta_4 +theme(legend.position = 'none')
```


# Maaslin2 asv

## @abx effect

```{r}
# use the PBS_vehicle as the reference level
df_target <- meta  %>%
  select(-group, -tube_no, -investigator) %>%
  mutate(experiment_no = as.character(experiment_no),
         day = as.character(day)) %>% 
  mutate(treatment = str_glue('{abx_treatment}_{diet_treatment}'),
       treatment = factor(treatment, levels = c( 'PBS_vehicle','PBS_sucrose', 'biapenem_vehicle','biapenem_sucrose'))) 

asv_target <- healthyall %>%
  inner_join(df_target %>% select(sampleid, treatment)) %>% 
  select(Taxon, sampleid, relab, treatment, day, experiment_no)

asv_target_sample_n <- asv_target %>% distinct(sampleid) %>% nrow

kept_asv <- asv_target %>%
  group_by(Taxon) %>%
  count(relab > 0.0001) %>%
  rename(criteria = names(.)[2]) %>%
  filter(criteria == 'TRUE') %>%
  mutate(perc = round(n/asv_target_sample_n*100, 0)) %>%
  filter(perc > 10) %>%
  pull(Taxon)

asv_relab_target_fil <-asv_target %>%
  filter( Taxon %in% kept_asv) %>%
  spread('Taxon','relab', fill = 0) %>% 
  mutate(day = factor(day),
         experiment_no = factor(experiment_no))
```

```{r}
# the setup for the Maaslin2
metadata <- asv_relab_target_fil %>% select(sampleid:experiment_no) %>% column_to_rownames('sampleid')
formula <- ~ treatment * day + experiment_no
design_matrix <- model.matrix(formula, data = metadata)
metadata <- cbind(metadata, design_matrix)

paste(colnames(metadata) , collapse = ", ")
```

```{r}
# Create a MaAsLin2 input feature data file (ASV table)
feature_data <- asv_relab_target_fil %>% 
  select(sampleid, starts_with('seq')) %>% 
  column_to_rownames('sampleid')

fit_res <- Maaslin2(
  input_data = feature_data, 
  input_metadata = metadata,
  output = str_glue("../data/R01_maaslin2_output_ALL_grp4_pbs_reference_abx_effects"),  # Output directory
  fixed_effects = c('treatmentPBS_sucrose', 'treatmentbiapenem_vehicle', 'treatmentbiapenem_sucrose', 'day3', 'day6', 'experiment_no2', 'treatmentPBS_sucrose:day3', 'treatmentbiapenem_vehicle:day3', 'treatmentbiapenem_sucrose:day3', 'treatmentPBS_sucrose:day6', 'treatmentbiapenem_vehicle:day6', 'treatmentbiapenem_sucrose:day6'), 
  random_effects = NULL,  # No random effects in this case
  normalization = "TSS",  # Total sum scaling normalization
  transform = "LOG",  # Log transformation actually log2
  analysis_method = "LM",  # Linear model
  max_significance = 0.05,
  min_abundance = 0.0001,  # Minimum abundance threshold
  min_prevalence = 0.10,  # Minimum prevalence threshold,
  cores = 10
)
sig_res = read_tsv(str_glue('../data/R01_maaslin2_output_ALL_grp4_pbs_reference_abx_effects/significant_results.tsv'))

```

### RUN volcano d3

```{r}
set.seed(1)
coeff_upper <- 1
coeff_lower <- -1
# to use the volcano plot to highlight the results on day 3
all_res_abx <- read_tsv('../data/R01_maaslin2_output_ALL_grp4_pbs_reference_abx_effects/all_results.tsv') %>% 
  mutate(genus = str_extract(feature, 'g__.+s__'),
         species = str_extract(feature, 's__.+$'),
         genus = str_replace(genus, 'g__',''),
         species = str_replace(species, 's__',''),
         genus = str_replace(genus, '\\.s__',''),
         cleaned_name = str_glue('{genus} {species}')) %>% 
  separate(feature, into = c('seq_id','taxa'), sep = '\\.', remove = F) %>% 
  # create a column with the feature name have seq id new line and then the actual taxa
  mutate(feature_formattted = str_replace(feature, '\\.','\n')) %>% 
  relocate(feature_formattted, .after = feature)


abx_vol_d3 <- all_res_abx %>%
  filter(metadata == 'treatmentbiapenem_vehicle:day3') %>% 
  mutate(sig_ = if_else(qval < 0.05 & coef > 1, 'pos', if_else(qval < 0.05 & coef < -1, 'neg', 'not'))) %>% 
  mutate(neglog10_q = -log10(qval)) %>%
  ggscatter(x = 'coef', y = 'neglog10_q', color = 'sig_', alpha = 0.5, shape = 16, 
            xlab = 'Maaslin2 model coefficient', ylab = '- log10(q value)',
            title = 'Between day 0 and day 3') +
  scale_color_manual(values = c('blue', 'black', 'red')) +
  geom_vline(xintercept =  0, color = 'black') + 
  geom_vline(xintercept =  c(coeff_upper, coeff_lower), color = 'gray', linetype = 'dashed') + 
  geom_hline(yintercept =  -log10(0.05), color = 'gray', linetype = 'dashed') + 
  geom_text_repel(aes(label = ifelse(qval < 0.05 & coef > coeff_upper, seq_id, "")), color = 'red') +
  geom_text_repel(aes(label = ifelse(qval < 0.05 & coef < coeff_lower, seq_id, "")), color = 'blue') +
  theme(aspect.ratio = 1/2, legend.position = 'none')
```


### volcano d6  

```{r}
abx_vol_d6 <- all_res_abx %>% 
  filter(metadata == 'treatmentbiapenem_vehicle:day6') %>% 
  mutate(sig_ = if_else(qval < 0.05 & coef > 1, 'pos', if_else(qval < 0.05 & coef < -1, 'neg', 'not'))) %>% 
  mutate(neglog10_q = -log10(qval)) %>%
  ggscatter(x = 'coef', y = 'neglog10_q', 
            color = 'sig_', 
            alpha = 0.5, shape = 16, 
            xlab = 'Maaslin2 model coefficient', ylab = '- log10(q value)',
            title = 'Between day 0 and day 6') +
  scale_color_manual(values = c( 'black', 'red')) +
  geom_vline(xintercept =  0, color = 'black') + 
  geom_vline(xintercept =  c(coeff_upper, coeff_lower), color = 'gray', linetype = 'dashed') + 
  geom_hline(yintercept =  -log10(0.05), color = 'gray', linetype = 'dashed') + 
  geom_text_repel(aes(label = ifelse(qval < 0.05 & coef > coeff_upper, seq_id, "")), color = 'red') +
  geom_text_repel(aes(label = ifelse(qval < 0.05 & coef < coeff_lower, seq_id, "")), color = 'blue') +
  theme(aspect.ratio = 1/2, legend.position = 'none')
```

### a table for all q < 0.05 results

```{r}
abx_effects_table <- all_res_abx %>% 
  filter(metadata == 'treatmentbiapenem_vehicle:day6' | metadata == 'treatmentbiapenem_vehicle:day3') %>% 
  filter(qval < 0.05) %>% 
  select(-value) %>% 
  arrange(metadata) %>% 
  mutate(coef = round(coef, 2))
```

### relab plot for these few highlighed asv

```{r}
abx_asvs <- abx_effects_table %>% 
  filter(coef > 1 | coef < -1) %>% 
  distinct(feature)

abx_asvs_relab <- healthyall %>% 
  mutate(feature = str_replace(Taxon, '\\:','.'),
         feature = str_replace_all(feature, '\\;','.')) %>% 
  filter(feature %in% abx_asvs$feature)%>% 
  # create a column with the feature name have seq id new line and then the actual taxa
  mutate(feature_formattted = str_replace(feature, '\\.','\n')) 

abx_asvs_relab_plots <- abx_asvs_relab %>% 
  split(.$feature_formattted) %>% 
  imap(function(df, name_){
    df %>% 
      filter(diet_treatment == 'vehicle') %>% 
      mutate(abx_treatment = factor(abx_treatment , levels = c('PBS','biapenem')),
             u_mouse_id = str_glue('{experiment_no}_{mouse_no}')) %>% 
  ggboxplot(x = 'day', y = 'relab',outlier.shape = NA,
             xlab = 'Day', ylab = 'Relative abundance', add.params = list(alpha = 0.5, shape = 16),
            title = str_glue('{name_}'),
            #color  = 'grp',
            width = 0.6)+
       geom_point(aes(group = u_mouse_id),alpha=0.3, shape = 16) +
      geom_line(aes(group = u_mouse_id),alpha=0.3) +
   #scale_color_manual(values = c('gray76','#ffbcdc','gray32','deeppink2')) +
  #scale_x_discrete(labels=rep(c(0,3,6),4)) + 
      facet_grid(. ~ abx_treatment) +
  theme(axis.text =  element_text(size = 10, hjust = 1),plot.title = element_text(size = 10), axis.title=element_text(size=10),legend.position = 'none',aspect.ratio=1.5)
  })
abx_asvs_relab_plots
```

##  @sugar effect on top of abx effect

```{r}
# to put all the healthy mice data in one single model
df_target <- meta  %>%
  select(-group, -tube_no, -investigator) %>%
  mutate(experiment_no = as.character(experiment_no),
         day = as.character(day)) %>% 
  mutate(treatment = str_glue('{abx_treatment}_{diet_treatment}'),
       treatment = factor(treatment, levels = c('biapenem_vehicle','biapenem_sucrose', 'PBS_vehicle','PBS_sucrose'))) 
```

```{r}
asv_target <- healthyall %>%
  inner_join(df_target %>% select(sampleid, treatment)) %>% 
  select(Taxon, sampleid, relab, treatment, day, experiment_no)

asv_target_sample_n <- asv_target %>% distinct(sampleid) %>% nrow

kept_asv <- asv_target %>%
  group_by(Taxon) %>%
  count(relab > 0.0001) %>%
  rename(criteria = names(.)[2]) %>%
  filter(criteria == 'TRUE') %>%
  mutate(perc = round(n/asv_target_sample_n*100, 0)) %>%
  filter(perc > 10) %>%
  pull(Taxon)

asv_relab_target_fil <-asv_target %>%
  filter( Taxon %in% kept_asv) %>%
  spread('Taxon','relab', fill = 0) %>% 
  mutate(day = factor(day),
         experiment_no = factor(experiment_no))
```

```{r}
# the setup for the Maaslin2
metadata <- asv_relab_target_fil %>% select(sampleid:experiment_no) %>% column_to_rownames('sampleid')
formula <- ~ treatment * day + experiment_no
design_matrix <- model.matrix(formula, data = metadata)
metadata <- cbind(metadata, design_matrix)

paste(colnames(metadata) , collapse = ", ")
```

```{r}
# Create a MaAsLin2 input feature data file (ASV table)
feature_data <- asv_relab_target_fil %>% 
  select(sampleid, starts_with('seq')) %>% 
  column_to_rownames('sampleid')

fit_res <- Maaslin2(
  input_data = feature_data, 
  input_metadata = metadata,
  output = str_glue("../data/R01_maaslin2_output_ALL_grp4"),  # Output directory
  fixed_effects = c('treatmentbiapenem_sucrose', 'treatmentPBS_vehicle', 'treatmentPBS_sucrose', 'day3', 'day6', 'experiment_no2', 'treatmentbiapenem_sucrose:day3', 'treatmentPBS_vehicle:day3', 'treatmentPBS_sucrose:day3', 'treatmentbiapenem_sucrose:day6', 'treatmentPBS_vehicle:day6', 'treatmentPBS_sucrose:day6'), 
  random_effects = NULL,  # No random effects in this case
  normalization = "TSS",  # Total sum scaling normalization
  transform = "LOG",  # Log transformation actually log2
  analysis_method = "LM",  # Linear model
  max_significance = 0.05,
  min_abundance = 0.0001,  # Minimum abundance threshold
  min_prevalence = 0.10,  # Minimum prevalence threshold,
  cores = 10
)
sig_res = read_tsv(str_glue('../data/R01_maaslin2_output_ALL_grp4/significant_results.tsv'))

```

### RUN volcano d3

```{r}

# to use the volcano plot to highlight the results on day 3
all_res_sugar <- read_tsv('../data/R01_maaslin2_output_ALL_grp4/all_results.tsv') %>% 
  mutate(genus = str_extract(feature, 'g__.+s__'),
         species = str_extract(feature, 's__.+$'),
         genus = str_replace(genus, 'g__',''),
         species = str_replace(species, 's__',''),
         genus = str_replace(genus, '\\.s__',''),
         cleaned_name = str_glue('{genus} {species}')) %>% 
  separate(feature, into = c('seq_id','taxa'), sep = '\\.', remove = F) %>% 
  # create a column with the feature name have seq id new line and then the actual taxa
  mutate(feature_formattted = str_replace(feature, '\\.','\n')) %>% 
  relocate(feature_formattted, .after = feature)

# A coefficient of 2 means an 2²-fold change (approximately a 4-fold increase) in relative abundance.
# A coefficient of 3 means an 2³-fold change (approximately a 8-fold increase) in relative abundance.
 
sugar_vol_d3 <- all_res_sugar %>%
  filter(metadata == 'treatmentbiapenem_sucrose:day3') %>% 
  mutate(sig_ = if_else(qval < 0.05 & coef > 1, 'pos', if_else(qval < 0.05 & coef < -1, 'neg', 'not'))) %>% 
  mutate(neglog10_q = -log10(qval)) %>%
  ggscatter(x = 'coef', y = 'neglog10_q', color = 'sig_', alpha = 0.5, shape = 16, 
            xlab = 'Maaslin2 model coefficient', ylab = '- log10(q value)',
            title = 'Between day 0 and day 3') +
  scale_color_manual(values = c('blue', 'black', 'red')) +
  geom_vline(xintercept =  0, color = 'black') + 
  geom_vline(xintercept =  c(coeff_upper, coeff_lower), color = 'gray', linetype = 'dashed') + 
  geom_hline(yintercept =  -log10(0.05), color = 'gray', linetype = 'dashed') + 
  geom_text_repel(aes(label = ifelse(qval < 0.05 & coef > coeff_upper, seq_id, "")), color = 'red') +
  geom_text_repel(aes(label = ifelse(qval < 0.05 & coef < coeff_lower, seq_id, "")), color = 'blue') +
  theme(aspect.ratio = 1/2, legend.position = 'none')
sugar_vol_d3
```

### volcano d6

```{r}
# asv at day 6
sugar_vol_d6 <-all_res_sugar %>%
  filter(metadata == 'treatmentbiapenem_sucrose:day6') %>% 
  mutate(sig_ = if_else(qval < 0.05 & coef > 1, 'pos', if_else(qval < 0.05 & coef < -1, 'neg', 'not'))) %>% 
  mutate(neglog10_q = -log10(qval)) %>%
  ggscatter(x = 'coef', y = 'neglog10_q', color = 'sig_', alpha = 0.5, shape = 16, 
            xlab = 'Maaslin2 model coefficient', ylab = '- log10(q value)',
            title = 'Between day 0 and day 6') +
  scale_color_manual(values = c('blue', 'black', 'red')) +
  geom_vline(xintercept =  0, color = 'black') + 
  geom_vline(xintercept =  c(coeff_upper, coeff_lower), color = 'gray', linetype = 'dashed') + 
  geom_hline(yintercept =  -log10(0.05), color = 'gray', linetype = 'dashed') + 
  geom_text_repel(aes(label = ifelse(qval < 0.05 & coef > coeff_upper, seq_id, "")), color = 'red') +
  geom_text_repel(aes(label = ifelse(qval < 0.05 & coef < coeff_lower, seq_id, "")), color = 'blue') +
  theme(aspect.ratio = 1/2, legend.position = 'none')

```

### a table for all q < 0.05 results

```{r}
sugar_effects_table <- all_res_sugar %>% 
  filter(metadata == 'treatmentbiapenem_sucrose:day6' | metadata == 'treatmentbiapenem_sucrose:day3') %>% 
  filter(qval < 0.05) %>% 
  select(-value) %>% 
  arrange(metadata)

```

### relab plot for these few highlighed asv

```{r}
sugar_asvs <- sugar_effects_table %>% 
  filter(coef > 1 | coef < -1) %>% 
  distinct(feature)

sugar_asvs_relab <- healthyall %>% 
  mutate(feature = str_replace(Taxon, '\\:','.'),
         feature = str_replace(feature, '\\;','.')) %>% 
  filter(feature %in% sugar_asvs$feature) %>% 
  # create a column with the feature name have seq id new line and then the actual taxa
  mutate(feature_formattted = str_replace(feature, '\\.','\n')) 

sugar_asvs_relab_plots <- sugar_asvs_relab %>% 
  split(.$feature_formattted) %>% 
  imap(function(df, name_){
    df %>% 
      filter(abx_treatment == 'biapenem') %>% 
      mutate(diet_treatment = factor(diet_treatment , levels = c('vehicle','sucrose')),
             u_mouse_id = str_glue('{experiment_no}_{mouse_no}')) %>% 
  ggboxplot(x = 'day', y = 'relab',outlier.shape = NA,
             xlab = 'Day', ylab = 'Relative abundance',, add.params = list(alpha = 0.5, shape = 16),
            title = str_glue('{name_}'),
            #color  = 'grp',
            width = 0.6)+
       geom_point(aes(group = u_mouse_id),alpha=0.3, shape = 16) +
      geom_line(aes(group = u_mouse_id),alpha=0.3) +
      facet_grid(. ~ diet_treatment) +
  theme(axis.text =  element_text(size = 10, hjust = 1),plot.title = element_text(size = 10), axis.title=element_text(size=10),legend.position = 'none',aspect.ratio=1.5)
  })
sugar_asvs_relab_plots
```


# Final graphs 

```{r}
# the legend for the pcoa
pcoa_leg <-  plot_grid(ggpubr::get_legend(beta_4))
ggsave('../data/01_pcoa_leg.pdf')
```


## page1

```{r}
title <- ggdraw() + 
  draw_label(
    "Fig. S14",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )

# the top row of the alpha and beta diversity
top_2 <- plot_grid(alpha_change, beta_4_plot, nrow = 1, align = 'vh', axis = 'lrtb', rel_widths = c(1,1.5), labels = c('A','B'))

abx_top <- abx_vol_d3 + abx_vol_d6 + 
  plot_annotation(title = "Differential abundance in DPBS + vehicle VS abx + vehicle", theme = theme(plot.title = element_text(hjust = 0.5)))

abx_relabs <- plot_grid(plotlist = abx_asvs_relab_plots, nrow = 1, align = 'vh', axis = 'lrtb')

p1 <- plot_grid(top_2, abx_top, abx_relabs, nrow = 3, labels = c(NA, 'C', 'D'))
 
combined <- plot_grid(
  title, p1,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.05, 1)
)+theme(plot.margin = unit(c(1,1,1,1), "cm"))

ggsave('../results//S14_mice_16s_p1_R01.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)  
```

## page2

```{r} 
title <- ggdraw() + 
  draw_label(
    "Fig. S14 (Continued)",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )


sugar_top <- plot_grid(sugar_vol_d3, sugar_vol_d6, nrow = 1, align = 'vh', axis = 'lrtb')
# Combine plots with shared title
sugar_top_plot <- sugar_vol_d3 + sugar_vol_d6 + 
  plot_annotation(title = "Differential abundance in abx + sucrose VS abx + vehicle", theme = theme(plot.title = element_text(hjust = 0.5)))

sugar_relabs <- plot_grid(plotlist = sugar_asvs_relab_plots, nrow = 2, align = 'vh', axis = 'lrtb')
sugars <- plot_grid(sugar_top_plot,sugar_relabs, nrow = 2 , align = 'vh', axis = 'lrtb', rel_heights = c(1,2), labels = c('E','F'))

 
combined <- plot_grid(
  title, sugars,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.05, 1)
)+theme(plot.margin = unit(c(1,1,1,1), "cm"))

ggsave('../results/S14_mice_16s_p2_R01.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)  
```

# meta data for the SRA

```{r}
accessions <- read_tsv('../data/mice_SRA.tsv') %>% 
  dplyr::select(accession, sampleid = library_ID) %>% 
  inner_join(meta %>% dplyr::select(sampleid, experiment_no,day,  mouse_no,abx_treatment,  diet_treatment))

accessions %>% write_csv('../data/01_mice_accession_info.csv')
```

