---
title: "beta diversity of microbiome and diet"
output: html_document
date: "2024-07-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(vegan)
library(cowplot)
```

Use the prior two day summary of the data, meaning count the average of the prior two day and their distance to the earliest 

# to summarize the food code based average of the prior two days 

```{r}
dtb <- read_csv('../data/152_combined_DTB.csv')
meta <- read_csv('../data/153_combined_META.csv')
```

```{r}
stb_pair <- meta %>%
  select(pid, sdrt) %>%
  transmute(pid = pid,
            p1d = sdrt-1,
            p2d = sdrt-2)

food_code_daily <- dtb %>%
  group_by(pid, fdrt, Food_code) %>%
  summarise(fc_daily = sum(dehydrated_weight))
 
mean_p2d_fc <-  function(pid_, p1d_, p2d_){
  df = food_code_daily %>%
    filter(pid == pid_) %>%
    filter(fdrt %in% c(p1d_, p2d_  )) %>%
    group_by(Food_code) %>%
    summarise(ave_fc = sum(fc_daily)/2)
  return(df)
} 

mean_p2d_fc_df <- pmap(stb_pair, function(pid, p1d, p2d){
    mean_p2d_fc(pid, p1d, p2d)
  }) %>%
  set_names(meta %>% pull(sampleid)) %>%
  bind_rows(.id = 'sampleid')
 
```
```{r}
# transform the shape so that I can calculate the unweighted unifrac distance 
p2d_fc_df <- mean_p2d_fc_df %>% 
  spread('sampleid','ave_fc', fill = 0) 

p2d_fc_df %>% write_tsv('../data/R06_p2d_fc_df.tsv')

```

```{bash}
biom convert -i  R06_p2d_fc_df.tsv -o R06_p2d_fc_df.biom --to-hdf5 --table-type="Table"
```


```{bash}
qiime tools import --input-path R06_p2d_fc_df.biom --output-path  R06_p2d_fc_df.qza --type "FeatureTable[Frequency]"
```

```{bash}
 qiime diversity beta-phylogenetic \
            --i-table R06_p2d_fc_df.qza \
            --i-phylogeny output_food_tree_datatree.qza \
            --p-metric unweighted_unifrac \
            --o-distance-matrix R06_p2d_fc_unweighted_unifrac_distance_matrix.qza
```

```{bash}
qiime diversity pcoa \
            --i-distance-matrix R06_p2d_fc_unweighted_unifrac_distance_matrix.qza \
            --o-pcoa R06_p2d_fc_unweighted_unifrac_distance_PCOA.qza

```

```{bash}
qiime tools export --input-path R06_p2d_fc_unweighted_unifrac_distance_PCOA.qza --output-path R06_p2d_fc_unweighted_unifrac_distance_PCOA
```

```{bash}
qiime tools export \
    --input-path  R06_p2d_fc_unweighted_unifrac_distance_matrix.qza \
    --output-path R06_p2d_fc_unweighted_unifrac_distance_matrix
```


```{r}
# import the ordination results
dfn <- '../data/R06_p2d_fc_unweighted_unifrac_distance_PCOA/ordination.txt'
dpcoa <- read_tsv(dfn, skip = 9, col_names = F)  %>% 
    filter(! X1 %in% c('Biplot','Site constraints')) %>%
    select(sampleid = X1, PC1 = X2, PC2 = X3) %>% 
  inner_join(meta %>% select(sampleid, pid, sdrt))

dpcoa_dist_earliest <- dpcoa %>% 
  group_by(pid) %>%
  arrange(sdrt) %>%
  mutate(distance_to_earliest = sqrt((PC1 - first(PC1))^2 + (PC2 - first(PC2))^2))

# some patient follows a very clear trajectory of the sdrt up, distance also up.
# dpcoa_dist_earliest %>% 
#   ggscatter(x = 'sdrt', y = 'distance_to_earliest', facet.by = 'pid')
```

# similary do this for the stool samples relative abundance at genus level

```{r}
relab_genus <- read_csv('../data/022_ALL173_stool_samples_genus_counts.csv') %>% 
  spread(key = 'genus', value = 'relab', fill = 0) %>% 
  column_to_rownames('sampleid')

dist_ <- vegdist(relab_genus, method = 'bray')
eigen <- cmdscale(dist_, eig = T)$eig
percent_var <- signif(eigen/sum(eigen), 3)*100
bc <- cmdscale(dist_, k = 2)
beta_ <- bc %>%
  as.data.frame() %>%
  rownames_to_column('sampleid')  %>% 
  inner_join(meta %>% select(sampleid, pid, sdrt)) %>% 
  rename(PC1 = V1, PC2 = V2)

beta_dist_earliest <- beta_ %>% 
  group_by(pid) %>%
  arrange(sdrt) %>%
  mutate(distance_to_earliest = sqrt((PC1 - first(PC1))^2 + (PC2 - first(PC2))^2))
```
# now see if there is some association

```{r}
joined <- dpcoa_dist_earliest %>% 
  select(-PC1, -PC2) %>% 
  rename(diet_dist = distance_to_earliest) %>% 
  inner_join(beta_dist_earliest %>% select(-PC1, -PC2) %>% rename(stool_dist = distance_to_earliest)) %>% 
  # filter out the baseline compared to baseline (distance is zero)
  filter(diet_dist > 0)

# it's important to account for the repeated measurements here 
# joined %>% 
#   ggscatter(x = 'diet_dist', y = 'stool_dist', facet.by = 'pid')

```


```{r}
# use the linear mixed effects model  
library(lmerTest)
model = lmer(stool_dist ~ diet_dist + (1 | pid), data = joined)
ret = summary(model)
ret
coeff = ret$coefficients[2,1]
pval = ret$coefficients[2,5]

# residuals <- residuals(model)
# hist(residuals, main = "Distribution of Residuals", xlab = "Residuals")
# plot(density(residuals), main = "Density Plot of Residuals", xlab = "Residuals")
# 
# qqnorm(residuals)
# qqline(residuals) # Adds a reference line for the normal distribution
# 
# ggplot(data.frame(residuals), aes(x = residuals)) +
#   geom_histogram(aes(y = ..density..), binwidth = 0.1, color = "black", fill = "lightblue") +
#   geom_density(alpha = 0.2, fill = "red") +
#   labs(title = "Distribution of Residuals", x = "Residuals", y = "Density")
```
```{r}
library(ggpmisc)
Partial_residual <- joined %>% 
  ungroup() %>% 
  mutate(fixed_effects = predict(model, re.form=NA),
         residuals = residuals(model),
         partial_residuals = residuals + fixed_effects) %>% 
  ggplot( aes(x = diet_dist, y = partial_residuals)) +
  geom_point(alpha = 0.5, shape = 16) +
  stat_poly_eq(formula = y ~ x, 
               aes(label = paste(..eq.label.., ..rr.label.., sep = "\n"), colour = 'red'), 
               parse = TRUE) +  # Add the equation and R-squared
  geom_smooth(method = "lm", se = FALSE) +  
  labs(x = "Diet Distance", y = "Partial Residuals (Stool Distance)") +
  ggtitle("Partial Residual Plot") +
  theme_pubr() +
  theme(aspect.ratio = 1)

Partial_residual
  
# In a linear regression model, a partial residual is a way to isolate and visualize the relationship between the response variable and a specific predictor variable, while accounting for the effects of all the other predictor variables in the model.

# Formula:

# The partial residual for a predictor X_i is calculated as:
# 
# Partial Residual_i = Residual + β_i * X_i 

# Check Linearity Assumption: Partial residual plots are a valuable diagnostic tool to check if the assumption of linearity holds for each predictor in a multiple regression model.

# Direction of Association: The positive slope of the line confirms that there is a positive association between diet distance and stool distance. This means that, on average, as the dissimilarity between a patient's subsequent and initial food intake increases, so does the dissimilarity between their subsequent and initial stool microbiome composition.
# 
# Average Effect Size: The slope of the line quantifies the average change in stool distance for each unit increase in diet distance. For example, if the slope is 0.3, it would mean that a 1-unit increase in diet distance is associated with a 0.3-unit increase in stool distance, on average, after controlling for individual patient effects. 
```
```{r}
# was I plotting the original data in the above plot??
# joined %>% 
#   ungroup() %>% 
#   mutate(fixed_effects = predict(model, re.form=NA),
#          residuals = residuals(model),
#          partial_residuals = residuals + fixed_effects) %>% 
# ggplot( aes(x = diet_dist, y = stool_dist)) +
#   geom_point(alpha = 0.8, shape = 16) +
#   stat_poly_eq(formula = y ~ x, 
#                aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")), 
#                parse = TRUE) +  # Add the equation and R-squared
#   geom_smooth(method = "lm", se = FALSE) +  
#   labs(x = "Diet Distance", y = " (Stool Distance)") +
#   ggtitle("regression Plot") +
#   theme_pubr() +
#   theme(aspect.ratio = 1)
# No I was not! this one is not accounting for patient individual differences. 
```


```{r}
# would be good to do a visualization of the beta diversity of the food and stool 
# too hard to do the whole thing in ggplot, so start with a scatter plot and then fininsh in illustrator 
sbeta_example <-  beta_dist_earliest %>% 
  mutate(highlightpt = if_else(pid == 'P104', T, F),
         highlight_time = if_else(pid == 'P104' & sdrt == -2, T, F)) %>% 
  arrange(highlightpt)

sbeta_example_pcoa <- sbeta_example %>% 
  ggscatter(x = 'PC1', y = 'PC2', color = 'highlightpt', shape = 'highlight_time', alpha = 0.8,
            title = 'Microbiome Bray curtis distance PCoA') +
  scale_color_manual(name = "", values = c('gray','red'),
                       labels = c("Other patients", "Patient 104")) +
  scale_shape_manual(name = "", values = c(16,17),
                       labels = c("Later time points", "Earliest timepoint")) +
  theme(aspect.ratio =  1)
```
```{r}
dbeta_example <-  dpcoa_dist_earliest %>% 
  mutate(highlightpt = if_else(pid == 'P104', T, F),
         highlight_time = if_else(pid == 'P104' & sdrt == -2, T, F)) %>% 
  arrange(highlightpt)

dbeta_example_pcoa <- dbeta_example %>% 
  ggscatter(x = 'PC1', y = 'PC2', color = 'highlightpt', shape = 'highlight_time', alpha = 0.8,
            title = 'Diet unweighted unifrac distance PCoA') +
  scale_color_manual(name = "", values = c('gray','red'),
                       labels = c("Other patients", "Patient 104")) +
  scale_shape_manual(name = "", values = c(16,17),
                       labels = c("Later time points", "Earliest timepoint")) +
  theme(aspect.ratio =  1)
```
```{r}
# are there correlation within each pt own diet
library(vegan)

diet_dist_matrix <- read_tsv('../data/R06_p2d_fc_unweighted_unifrac_distance_matrix/distance-matrix.tsv') %>% 
  rename(sampleid = ...1) %>% 
  full_join(meta %>% select(pid, sampleid))

diet_matrix <- diet_dist_matrix %>% 
  select(-pid) %>% 
  column_to_rownames('sampleid') %>% as.matrix()

# Run PERMANOVA
adonis2(diet_matrix ~ diet_dist_matrix$pid, permutations = 999, parallel = 10) 
```
The results of the PERMANOVA strongly suggest that there is significant within-patient correlation in dietary intake. In other words, individual patients tend to have consistent dietary patterns over time, and these patterns are distinct from those of other patients.

```{r}
# I wanna repeat the above step for microbiome to look at within patient correlation
stool_dist_df <- dist_ %>%
  as.matrix() %>%
  as.data.frame() %>%
  rownames_to_column('sampleid') %>%
  inner_join(meta %>% select(sampleid, timebin, pid, empirical))

stool_matrix <- stool_dist_df %>%
  select(-pid, -timebin, empirical) %>%
  column_to_rownames('sampleid') %>% as.matrix()

# Run PERMANOVA
adonis2(stool_matrix ~ stool_dist_df$pid, permutations = 999, parallel = 10)
adonis2(stool_matrix ~ stool_dist_df$timebin, permutations = 999, parallel = 10)

both <- adonis2(stool_matrix ~ stool_dist_df$pid + stool_dist_df$timebin, permutations = 999, parallel = 10)
three <- adonis2(stool_matrix ~ stool_dist_df$pid + stool_dist_df$timebin + stool_dist_df$empirical, permutations = 999, parallel = 10)
```
 
Yes there is strong within individual correlation as well. 

```{r}
# to combine them together as one single supp figure
 title <- ggdraw() + 
  draw_label(
    "Fig. S_X",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 5)
  )


S_X <- plot_grid(dbeta_example_pcoa, sbeta_example_pcoa, Partial_residual, 
                 label_size = 12, ncol = 1,labels = c('A','B','C'),
                align = 'vh', axis = 'lrtb')

 
combined <- plot_grid(
  title, S_X,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.05, 1)
)+theme(plot.margin = unit(c(1,1,1,1), "cm"))

ggsave('../data/06_distance_to_earliest.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)   
```

