---
title: "Clinical outcome"
output: html_document
date: "2024-07-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(gtsummary)
library(ggsurvfit)
library(tidycmprsk)
library(survival)
library(ggpubr)
library(gt)
library(cowplot)
library(survminer)
library(patchwork)
library(geepack)
library(broom)
library(openxlsx)
```

# load the cleaned data

```{r}
#load("../data/df_main_Dec10.Rdata") 
# link <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/152_pid_match.csv')
# 
# df_main <- df_main |>
#   mutate(gvhdppx_CNI = as.numeric(gvhd_ppx == "CNI-based"),
#          gvhdppx_PTCy = as.numeric(gvhd_ppx == "PTCy-based"))
# 
# 
# colnames(df_main)
# df_main <- df_main %>%
#   select(source:gvhdppx_PTCy) %>%
#   mutate(mrn = as.numeric(mrn_str)) %>%
#   full_join(link) %>%
#   select(-d100_a_gvhd_onset, -mrn_str, -agvhd_onset,  -mrn) %>%
#   relocate(pid, .before = 'source')
# 
# df_main %>% write_rds('../data/R02_cleaned_clinical_outcome2.rds')

# the below table has the pid
df_main <- read_rds('../data/R02_cleaned_clinical_outcome.rds') %>%
  mutate(modal_diet = as.character(modal_diet),
         modal_diet = if_else(modal_diet == 'High intake', 'Higher-intake', 'Lower-intake/sugar-enriched'),
         modal_diet = factor(modal_diet, levels = c( 'Lower-intake/sugar-enriched', 'Higher-intake'))) %>% 
  mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>% 
  mutate(source_and_gvhdppx = str_glue('{source}+{gvhd_ppx}'),
         source_and_gvhdppx = factor(source_and_gvhdppx, levels = c('TCD+TCD','Cord+CNI-based','Unmodified+CNI-based','Unmodified+PTCy-based')),
         day_exposed_cat = factor(day_exposed_cat, levels = c('Short exposure (< 8 days)','Long exposure (>= 8 days)'))) %>% 
  mutate(day_exposed_cat = str_replace(day_exposed_cat, ' \\(.+\\)','')) %>% 
  mutate(abx_diet_cat = str_glue('{modal_diet}, {day_exposed_cat}'))
df_main %>% count(abx_diet_cat)
 
```
 
```{r}
colnames(df_main)
df_main %>% count(type_gvhd_simple)
df_main %>% count(abx_diet_cat)
```
 
# OS model

## survival 

```{r survival}
# survival between two clusters
OS_clusers <- survfit2(Surv(OStime_30,OSevent)~ modal_diet  , data = df_main |> filter(OStime_30 > 0 )) %>%
  ggsurvfit(linewidth = 0.8) +
  add_pvalue("annotation",size=5) +
  add_risktable()+
  ylab("Survival probability") +
  xlab("Days relative to HCT, landmarked at day 12")+
   scale_color_manual(values = c( "darkslateblue",  "darkgoldenrod2"), name = '') +
  ylim(c(0,1)) +
  ggtitle("")+
  theme(aspect.ratio = 1/1.5, legend.position = c(0.5, 0.3))
OS_clusers

survival_plot <- list(
  ggsurvfit::ggsurvfit_build(OS_clusers)
) |> 
  wrap_plots()

```

```{r}
# Kaplan-Meier survival curves for 4 groups
plot_colors <- c(
  "Higher-intake, Short exposure" = "#8B4513", # SaddleBrown (or a similar brown)
  "Higher-intake, Long exposure"  = "#FFD700", # Gold (Yellow)
  "Lower-intake/sugar-enriched, Short exposure"  = "#ADD8E6", # LightBlue
  "Lower-intake/sugar-enriched, Long exposure"   = "#00008B"  # DarkBlue
)

OS_clusers_4 <- survfit2(Surv(OStime_30,OSevent)~ abx_diet_cat  , data = df_main |> filter(OStime_30 > 0 )) %>%
  ggsurvfit(linewidth = 0.8) +
  scale_colour_manual(values = plot_colors) +
  add_pvalue("annotation",size=5) +
  add_risktable()+
  ylab("Survival probability") +
  xlab("Days relative to HCT, landmarked at day 12")+
   #scale_color_manual(values = c( "darkslateblue",  "darkgoldenrod2"), name = '') +
  ylim(c(0,1)) +
  ggtitle("")+
  theme(aspect.ratio = 1/1.5, legend.position = c(0.5, 0.3))
OS_clusers_4

survival_plot <- list(
  ggsurvfit::ggsurvfit_build(OS_clusers_4)
) |> 
  wrap_plots()     

# "At Risk" Table : For each group, this table tells you the number of patients who were still alive and being followed (i.e., had not yet experienced the event and had not been censored) at specific time points shown on the X-axis (0, 100, 200, 300 days landmarked). When the number "at risk" becomes very small (e.g., single digits), the survival curve estimates for that group at later time points become less reliable because they are based on fewer patients.
# Understanding Censoring: Drops in the "at risk" numbers that don't correspond to steps down in the curve can indicate censoring (patients lost to follow-up, or the study ended before they had an event).
# "Events" : For each group, this table typically shows the cumulative number of events (e.g., deaths) that have occurred up to the specific time points on the X-axis.
# "landmarked at day 12": This means the analysis effectively starts at HCT day 12. Only patients who were alive at day 12 are included, and their follow-up time begins from this point. This is often done to avoid "immortal time bias," where patients who survive longer have more opportunity to, for example, receive a longer duration of treatment, which could skew results if not accounted for.
```

## Table S3 model fit

```{r OS_model}
# rename the columns so that they look nice in the output table!
input_df <- df_main |> 
  filter(OStime_30 > 0) %>% 
  rename(`Diet-pattern cluster` = modal_diet,
         `Abx exposure(days)` =day_exposed,
         `Conditioning intensity` = intensity,
         Graft_type_and_GvHD_prophylaxis = source_and_gvhdppx) %>% 
  mutate(Graft_type_and_GvHD_prophylaxis = case_when(
    Graft_type_and_GvHD_prophylaxis == 'TCD+TCD' ~ 'T-cell depleted PBSC',
    Graft_type_and_GvHD_prophylaxis == 'Cord+CNI-based' ~ 'Cord blood + CNI-based',
    Graft_type_and_GvHD_prophylaxis == 'Unmodified+CNI-based' ~ 'Unmodified graft + CNI-based',
    Graft_type_and_GvHD_prophylaxis == 'Unmodified+PTCy-based' ~ 'Unmodified graft + PTCy-based'
  )) %>% 
  mutate(Graft_type_and_GvHD_prophylaxis = factor(Graft_type_and_GvHD_prophylaxis, levels = c('T-cell depleted PBSC','Cord blood + CNI-based','Unmodified graft + CNI-based','Unmodified graft + PTCy-based'))) %>% 
    rename(`Graft source and GvHD prophylaxis` = Graft_type_and_GvHD_prophylaxis) %>% 
   mutate(`Diet-pattern cluster` = case_when(
    `Diet-pattern cluster` == 'Lower-intake/sugar-enriched' ~ 'Cluster 1 (Lower-intake/sugar-enriched)',
    `Diet-pattern cluster` == 'Higher-intake' ~ 'Cluster 2 (Higher-intake)' 
  )) |> 
  mutate(exposure_in_week = `Abx exposure(days)`/7)



fit <- coxph(Surv(OStime_30,OSevent)~modal_diet*day_exposed  + intensity + source_and_gvhdppx,
             data= df_main |> filter(OStime_30 > 0) |>  mutate(modal_diet = case_when(
    modal_diet == 'Lower-intake/sugar-enriched' ~ 'Cluster 1 (Lower-intake/sugar-enriched)',
    modal_diet == 'Higher-intake' ~ 'Cluster 2 (Higher-intake)' 
  )))




# just print the multivar table to the correct size

title <- ggdraw() + 
  draw_label(
    "Table S3",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 5)
  )


table_OS <- ggdraw() + draw_image("../data/09_OS_model_multi.png", scale = .8)
    
combined <- plot_grid(
  title, table_OS,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.05, 1)
)+theme(plot.margin = unit(c(1,1,1,1), "cm"))

ggsave('../results//Table_S3_OS_multivariate_R09.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)  
```
```{r}
# make the above table publication ready with newer code
publication_table_OS_output_model_cluster <- tbl_regression(
  fit,
  # Just list the main variables you want to see.
  # gtsummary is smart enough to keep the interaction term 
  # associated with these variables automatically.
  include = c(modal_diet, day_exposed, "modal_diet:day_exposed"),
  exponentiate = T, # This converts log(HR) to Hazard Ratios (HR), which are more interpretable
  # The 'label' argument maps variable names to publication-ready labels.
  # The formula structure is list(variable_name ~ "Desired Label")
  label = list(
    modal_diet ~ "Diet-pattern cluster",
    day_exposed ~ "Abx exposure(days)"
    # source_and_gvhdppx ~ "Graft source and GvHD prophylaxis",
    # intensity ~ "Conditioning intensity"
    # Note: The interaction term label is automatically created from the main variable labels.
  )
  ) %>%
  # Improve the table's appearance and clarity
  bold_p(t = 0.05) %>% # Make p-values < 0.05 bold
  bold_labels() %>%
  # Add a clear title and a source note for context
  as_gt() %>% # Convert to a gt object for better customization
  gt::tab_header(
    title = "Multivariable Cox Model for Overall Survival"
  ) 


# --- Step 3: Display the Table ---
# In RStudio or a Quarto/RMarkdown document, this will render a beautiful HTML table.
publication_table_OS_output_model_cluster %>% 
  gt::gtsave(filename = "../data/R09_cluster_model_multivariate_output.png") 

# to produce the pdf version of the table in a pdf 
table_S5 <- ggdraw() + draw_image("../data/R09_cluster_model_multivariate_output.png", scale = 0.7)

 title <- ggdraw() + 
  draw_label(
    "Extended Table E5",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 5)
  )

combined <- plot_grid(
  title, table_S5,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.1, 1)
)+theme(plot.margin = unit(c(1,1,10,1), "cm"))

ggsave('../data/Table_S5_cluster_model_multivariate_output_R09.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)  

```
 
 
```{r}
# what if I change the exposure to be in weeks
fit_wk <- coxph(Surv(OStime_30,OSevent)~`Diet-pattern cluster`*exposure_in_week  + `Conditioning intensity` + `Graft source and GvHD prophylaxis`,
             data= input_df)

fit_wk |> 
  tbl_regression(exponentiate = T) |> 
  bold_p() |> 
  as_gt() |> 
  gtsave(filename="../data/09_OS_model_multi_wk.png")
```
 
## F3C  

```{r}
## The following lines construct a Wald test on the association between abx exposure and OS for high intake patients.
diff <- sum(coef(fit)[c(2,9)])
p <- aod::wald.test(Sigma=vcov(fit)[c(2,9),c(2,9)],
                    b=coef(fit)[c(2,9)],
                    L=t(c(1,1)))

## Visualizing associations between abx and OS for low and high intake patient groups:
loghr <- data.frame(exposure = 0:20,
                    hr = exp(c(coef(fit)[2]*0:20,
                               diff*0:20)),
                    group = c(rep("Lower intake",21),
                              rep("Higher intake",21)),
                    ub = exp(c((coef(fit)[2]+sqrt(vcov(fit)[2,2])*1.96)*0:20,
                               (diff+1.96*sqrt(sum(vcov(fit)[c(2,9),c(2,9)])))*0:20)),
                    lb = exp(c((coef(fit)[2]-sqrt(vcov(fit)[2,2])*1.96)*0:20,
                               (diff-1.96*sqrt(sum(vcov(fit)[c(2,9),c(2,9)])))*0:20))
)
```


```{r abx_VS_diet_logHR}
abx_diet_logHR <- loghr |> 
  ggplot(aes(x=exposure,y=hr)) + 
  geom_line(aes(color=group), size = 1.5)+
  geom_ribbon(aes(ymin=lb,ymax=ub),alpha=0.2) +
  geom_hline(yintercept = 1,color="black",linetype=2)+
  scale_y_continuous(transform = "log",labels=function(x) sprintf("%.0f", x))+
  #facet_grid(.~group, labeller = labeller(facet_var = c("Higher intake" = "Lower intake", "Lower intake" = "Higher intake"))) +
  facet_grid(.~group) +
  ylab("log(Hazard ratio) compared to\nnon-exposure patients") + 
  xlab("Number of days exposed to broad-spectrum\nantibiotics between day -7 to 12\nrelative to transplant") + theme_light()+
  scale_color_manual(values = c( "darkgoldenrod2", "darkslateblue"), name = '') +
  theme(aspect.ratio = 1,  legend.position = 'none') 

abx_diet_logHR
```

# me  
 
```{r}
fit_original <- coxph(Surv(OStime_30,OSevent)~modal_diet*day_exposed  +intensity + source_and_gvhdppx,
             data= df_main)
colnames(df_main)
```


```{r}
# --- 1. Extract Coefficients and Variance-Covariance Matrix ---
model_coefs <- coef(fit_original)
model_vcov <- vcov(fit_original)

# --- 2. Identify the EXACT Coefficient Names ---
# Run this line to see the names R assigned:
names(model_coefs)
```


```{r}
abx_coef_name <- "day_exposed" 
interaction_coef_name <- "modal_dietHigher-intake:day_exposed" 

# Check if the names exist in the model output
if (!abx_coef_name %in% names(model_coefs)) {
  stop("The specified name for the antibiotic exposure coefficient was not found in the model. Check names(coef(fit)).")
}
if (!interaction_coef_name %in% names(model_coefs)) {
  stop("The specified name for the interaction coefficient was not found in the model. Check names(coef(fit)).")
}
```


```{r}
# --- 3. Calculate Log(HR) and SE for Cluster 1 (Reference) ---

# Log(HR) is just the coefficient for the main effect
loghr_c1 <- model_coefs[abx_coef_name]

# Standard Error (SE) is the square root of the variance (diagonal element of vcov matrix)
se_c1 <- sqrt(model_vcov[abx_coef_name, abx_coef_name])

# --- 4. Calculate Log(HR) and SE for Cluster 2 ---

# Log(HR) is the sum of the main effect and the interaction term
loghr_c2 <- model_coefs[abx_coef_name] + model_coefs[interaction_coef_name]

# Standard Error (SE) for the sum requires variances and covariance:
# SE_sum = sqrt(Var(A) + Var(B) + 2*Cov(A, B))
var_abx <- model_vcov[abx_coef_name, abx_coef_name]
var_int <- model_vcov[interaction_coef_name, interaction_coef_name]
cov_abx_int <- model_vcov[abx_coef_name, interaction_coef_name]

se_c2 <- sqrt(var_abx + var_int + 2 * cov_abx_int)

# --- 5. Calculate 95% Confidence Intervals ---
# CI = Estimate +/- Z * SE (where Z is approx 1.96 for 95% CI)
z_crit <- qnorm(0.975) # More precise Z-value for 95%

ci_c1_low <- loghr_c1 - z_crit * se_c1
ci_c1_high <- loghr_c1 + z_crit * se_c1

ci_c2_low <- loghr_c2 - z_crit * se_c2
ci_c2_high <- loghr_c2 + z_crit * se_c2

# --- 6. Display Results ---
results <- data.frame(
  Cluster = c("Cluster 1:Lower-intake/sugar-enriched", "Cluster 2:Higher-intake"),
  LogHR = c(loghr_c1, loghr_c2),
  SE = c(se_c1, se_c2),
  CI_Lower = c(ci_c1_low, ci_c2_low),
  CI_Upper = c(ci_c1_high, ci_c2_high)
) %>% 
  mutate(Cluster = factor(Cluster, levels = c('Cluster 2:Higher-intake','Cluster 1:Lower-intake/sugar-enriched')))

```


```{r}
ggplot(results, aes(x = LogHR, y = Cluster, color = Cluster)) +
  # Add horizontal error bars using the calculated CIs
  geom_errorbarh(aes(xmin = CI_Lower, xmax = CI_Upper), height = 0.1, linewidth = 0.8) +
  # Add points for the log HR estimates
  geom_point(shape = 15, size = 4) + # Shape 15 is a solid square
  # Add vertical line at log HR = 0 (no effect)
  geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 0.5) +
  # Apply the custom colors
  scale_color_manual(values = c( "darkgoldenrod2", "darkslateblue")) +
  # Set axis labels
  labs(
    x = "Log HR per day\nof antibiotic exposure",
    y = NULL # Remove y-axis title
  ) +
  # Customize the theme
  theme_classic() + # A clean theme
  theme(
    #axis.line.y = element_blank(), # Remove y-axis line
    axis.ticks.y = element_blank(), # Remove y-axis ticks
    axis.text.y = element_text(hjust = 1, size = 10), # Align y-axis labels, adjust size
    axis.title.x = element_text(size = 11), # Adjust x-axis title size
    legend.position = "none" # Hide the legend
  ) +
  # Set x-axis limits - adjust these based on your actual CI ranges for a good view
  coord_cartesian(xlim = c(-0.15, 0.25))  
```
## cluster model + ave caloric intake 

```{r}
df_main <- read_rds('../data/R09_df_main_for_nutrition_clinical_outcome.RDS') |> 
  inner_join(read_rds('../data/R02_cleaned_clinical_outcome.rds') |> select(pid, modal_diet))

fit_cal_cluster <- coxph(Surv(OStime_30,OSevent)~ modal_diet*day_exposed  + intensity + source_and_gvhdppx + avg_total_caloric_intake,
             data= df_main |> filter(OStime_30 > 0)) 
```
### plotting 

```{r}
# Pull tidy coefficients from the fitted coxph object and keep just the two
# antibiotic-exposure effects we want to display (one per cluster).
plot_df <- fit_cal_cluster |>
  tidy(conf.int = TRUE) |>
  filter(term %in% c("day_exposed", "modal_dietHigh intake:day_exposed")) |>
  # Cluster 1 effect is day_exposed alone; Cluster 2 is day_exposed + interaction.
  # Easier path: refit and use emmeans, but here we just reconstruct manually.
  mutate(
    cluster = if_else(term == "day_exposed", "Cluster 1", "Cluster 2"),
    # For Cluster 2, the displayed effect should be the sum of main + interaction.
    # Pull both rows, then combine below.
  )

# Cleaner: grab the main effect and the sum (main + interaction) with their SEs
# from the variance-covariance matrix so the CIs are correct.
vc <- vcov(fit_cal_cluster)
b  <- coef(fit_cal_cluster)

main_est  <- b["day_exposed"]
main_se   <- sqrt(vc["day_exposed", "day_exposed"])

sum_est <- b["day_exposed"] + b["modal_dietHigh intake:day_exposed"]
sum_se  <- sqrt(
  vc["day_exposed", "day_exposed"] +
  vc["modal_dietHigh intake:day_exposed", "modal_dietHigh intake:day_exposed"] +
  2 * vc["day_exposed", "modal_dietHigh intake:day_exposed"]
)

forest_df <- tibble(
  cluster  = c("Cluster 1", "Cluster 2"),
  estimate = c(main_est, sum_est),
  se       = c(main_se,  sum_se)
) |>
  mutate(
    conf.low  = estimate - 1.96 * se,
    conf.high = estimate + 1.96 * se,
    # Reverse the factor so Cluster 1 appears on top, matching the published figure.
    cluster   = fct_rev(factor(cluster))
  )

# Plot on the log-HR scale to match Figure 3b exactly.
ggplot(forest_df, aes(x = estimate, y = cluster)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, linewidth = 0.5) +
  geom_point(shape = 22, fill = "black", color = "black", size = 2) +
  scale_x_continuous(limits = c(-0.1, 0.2), breaks = seq(-0.1, 0.2, 0.1)) +
  labs(x = "Log HR per day\nof antibiotic exposure", y = NULL) +
  theme_classic(base_size = 12) 

ggsave('../data/R09_forest_fit_cal_cluster.png', width = 3, height = 2)
```


#  now F3

```{r}
# the legend for the stacked bar graph
bar_leg <-  plot_grid(ggpubr::get_legend(Average_intake))
ggsave('../data/09_bar_leg.pdf')
```

```{r}
title <- ggdraw() + 
  draw_label(
    "Fig. 3",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 5)
  )

top <- plot_grid(pdiet2, pdiet2_comp, Average_intake_plot, Average_composition,
                 label_size = 12, ncol = 2,labels = c('A','B','C','D'),
                align = 'vh', axis = 'lrtb')




bottom <- plot_grid(survival_plot, abx_diet_logHR,
                 label_size = 12, ncol = 2,labels = c('E','F'),rel_widths  = c(1,1),
                align = 'vh', axis = 'lrtb')

all <- plot_grid(top, bottom, align = 'vh', axis = 'lrtb', nrow = 2, rel_heights = c(2,1.5))
 
combined <- plot_grid(
  title, all,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.05, 1)
)+theme(plot.margin = unit(c(1,1,1,1), "cm"))

ggsave('../data/09_clinical_outcome.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)  
```

# Sugar density and OS

```{r}
dtb <- read_csv('../data/152_combined_DTB.csv')

daily_intake <- dtb %>%
  # Group by patient ID (pid) and day relative to transplant (fdrt)
  group_by(pid, fdrt) %>%
  # Calculate the sum of calories and sugar for each day
  summarise(
    total_caloric_intake = sum(Calories_kcal, na.rm = TRUE),
    sugar_intake_grams = sum(Sugars_g, na.rm = TRUE)
  ) %>%
  ungroup()

# --- Step 2: Calculate Calorie-Adjusted Sugar Intake ---
# This pipe will now take the aggregated daily data, calculate the average 
# sugar density (grams per 1000 kcal), and prepare it for the model.
patient_summary <- daily_intake %>%
  # Filter for the specified time window of day -7 to day 12
  filter(fdrt >= -7 & fdrt <= 12) %>%

  # Calculate the daily sugar density (grams of sugar per 1000 kcal)
  mutate(
    daily_sugar_density = if_else(total_caloric_intake > 0,
                                  sugar_intake_grams / (total_caloric_intake / 1000),
                                  0) # If no calories, density is 0
  ) %>%

  # Group by patient to create a single summary row for each one
  group_by(pid) %>%
  summarise(
    # also summarize the absolute gram weight intake
    avg_sugar_daily_intake_gram = mean(sugar_intake_grams, na.rm = TRUE),
    # This is your diet COMPOSITION variable
    avg_sugar_density_per_1000kcal = mean(daily_sugar_density, na.rm = TRUE),
    # This is your diet QUANTITY variable
    avg_total_caloric_intake = mean(total_caloric_intake, na.rm = TRUE)
  ) %>%
  ungroup() %>% 
  inner_join(df_main %>% select(pid, OStime_30, OSevent, intensity, source_and_gvhdppx, day_exposed, source, gvhd_ppx, age, sex, Disease = disease.simple)) %>%
  mutate(
    # Create a factor "High" or "Low" based on the median of the cohort
    SugarCal_cat_high = if_else(avg_sugar_density_per_1000kcal > median(avg_sugar_density_per_1000kcal, na.rm = TRUE),
                                          "Above-median", "Below-median"),
    SugarCal_cat_high = factor(SugarCal_cat_high, levels = c('Above-median', 'Below-median'))
  )

patient_summary |> write_rds('../data/R09_df_main_for_nutrition_clinical_outcome.RDS')
```

## two scatter correlation

```{r}
# --- Plot 1: Raw Sugar Intake vs. Total Calories ---
# This plot shows the relationship between the total amount of food eaten
# and the absolute amount of sugar consumed. We expect a strong positive correlation,
# as people who eat more food in general will also eat more sugar.
plot_raw_sugar <- ggplot(patient_summary, aes(x = avg_total_caloric_intake / 1000, y = avg_sugar_daily_intake_gram)) +
  geom_point(alpha = 0.6, color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  # This function from ggpubr calculates and adds the Pearson correlation coefficient
  stat_cor(method = "spearman", label.x = 1, label.y = 170, p.accuracy = 0.001) +
  labs(
    title = "Raw Sugar Intake vs. Total Calories",
    x = "Average Daily Caloric Intake (in 1000s kcal)",
    y = "Average Daily Raw Sugar Intake (grams)"
  ) +
  theme_minimal()

# --- Plot 2: Calorie-Adjusted Sugar vs. Total Calories ---
# This plot checks if our calorie-adjustment worked. After normalizing sugar
# intake to a standard energy unit (grams per 1000 kcal), there should be
# little to no correlation with total calories. This confirms we have
# successfully removed total intake as a confounder.
plot_adjusted_sugar <- ggplot(patient_summary, aes(x = avg_total_caloric_intake / 1000, y = avg_sugar_density_per_1000kcal)) +
  geom_point(alpha = 0.6, color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  # Calculate and add the Pearson correlation for this plot
  stat_cor(method = "spearman", label.x = 1, label.y = 170, p.accuracy = 0.001) +
  labs(
    title = "Calorie-Adjusted Sugar vs. Total Calories",
    x = "Average Daily Caloric Intake (in 1000s kcal)",
    y = "Average Sugar Density (grams per 1000 kcal)"
  ) +
  theme_minimal()


# --- Arrange and display both plots side-by-side ---
# This makes it easy to compare the two relationships directly.
ggarrange(plot_raw_sugar, plot_adjusted_sugar, ncol = 2)

```

## model 

```{r}
fit_sugar_density_binarized <- coxph(Surv(OStime_30,OSevent)~ SugarCal_cat_high * day_exposed + source_and_gvhdppx + intensity ,
             data=patient_summary |> filter(OStime_30 > 0))
```
## forest plot

```{r}
# Pull per-group log HR for antibiotic exposure from the coef vector and vcov.
# Above-median sugar is the reference → its effect is `day_exposed` alone.
# Below-median sugar's effect is `day_exposed` + the interaction term.
vc <- vcov(fit_sugar_density_binarized)
b  <- coef(fit_sugar_density_binarized)

above_est <- b["day_exposed"]
above_se  <- sqrt(vc["day_exposed", "day_exposed"])

below_est <- b["day_exposed"] + b["SugarCal_cat_highBelow-median:day_exposed"]
below_se  <- sqrt(
  vc["day_exposed", "day_exposed"] +
  vc["SugarCal_cat_highBelow-median:day_exposed",
     "SugarCal_cat_highBelow-median:day_exposed"] +
  2 * vc["day_exposed", "SugarCal_cat_highBelow-median:day_exposed"]
)

forest_df <- tibble(
  group    = c("above-median\nsugar intake\ng/1000 kcal",
               "below-median\nsugar intake\ng/1000 kcal"),
  estimate = c(above_est, below_est),
  se       = c(above_se,  below_se)
) |>
  mutate(
    # Build CIs on the log-HR scale, then exponentiate everything to HR scale
    conf.low  = estimate - 1.96 * se,
    conf.high = estimate + 1.96 * se,
    HR        = exp(estimate),
    HR.low    = exp(conf.low),
    HR.high   = exp(conf.high),
    # Keep above-median on top to match the published Fig. 3a layout
    group     = fct_inorder(group) |> fct_rev()
  )

# Forest plot on the HR scale with a log-scaled x-axis so CIs look symmetric
forest_sugar_density_binarized <- ggplot(forest_df, aes(x = HR, y = group)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = HR.low, xmax = HR.high),
                 height = 0, linewidth = 0.5) +
  geom_point(shape = 22, fill = "black", color = "black", size = 2) +
  scale_x_log10() +
  labs(x = "HR per day\nof antibiotic exposure", y = NULL) +
  theme_classic(base_size = 12)
```

 
## model table output

```{r}
publication_table <- tbl_regression(
  fit_sugar_density_binarized,
  include = c(SugarCal_cat_high, day_exposed, "SugarCal_cat_high:day_exposed"),
  exponentiate = TRUE,
  label = list(
    SugarCal_cat_high ~ "Averaged Sugar Intake",
    day_exposed ~ "Abx exposure (days)"
  )
) |>
  bold_p(t = 0.05) |>
  bold_labels()

body <- publication_table$table_body
label_rows <- which(body$row_type == "label")
level_rows <- which(body$row_type == "level")
sig_rows   <- which(!is.na(body$p.value) & body$p.value < 0.05)

reg_df <- publication_table |> as_tibble()

# Strip markdown markers from EVERY cell, not just headers and column 1
# mutate(across(...)) walks every column; the regex catches ** and __
reg_df <- reg_df |>
  mutate(across(everything(),
                ~ str_remove_all(as.character(.x), "\\*\\*|__")))

names(reg_df) <- names(reg_df) |> str_remove_all("\\*\\*|__")

# Convert text that looks like a number into actual numbers, leaving
# things like "0.15, 2.32" (the CI) and "—" (the reference dash) alone
to_numeric_if_possible <- function(x) {
  num <- suppressWarnings(as.numeric(x))
  # Only swap in the number where conversion succeeded
  ifelse(is.na(num), x, num)
}

# The CI column has commas, so keep it as text; convert HR and p-value
# Column positions: 1 = Characteristic, 2 = HR, 3 = 95% CI, 4 = p-value
reg_df <- reg_df |>
  mutate(
    across(c(2, 4), ~ suppressWarnings(as.numeric(.x)))
  )

wb <- createWorkbook()
addWorksheet(wb, "Cox Model")

writeData(wb, "Cox Model",
          "Multivariable Cox Model for Overall Survival",
          startRow = 1)
addStyle(wb, "Cox Model",
         createStyle(textDecoration = "bold", fontSize = 14),
         rows = 1, cols = 1)

writeData(wb, "Cox Model", reg_df, startRow = 3)

# Header styling
addStyle(wb, "Cox Model",
         createStyle(textDecoration = "bold",
                     fgFill = "#D9D9D9",
                     border = "bottom",
                     halign = "left"),
         rows = 3, cols = 1:ncol(reg_df), gridExpand = TRUE)

# Bold variable labels, indent levels
addStyle(wb, "Cox Model",
         createStyle(textDecoration = "bold"),
         rows = label_rows + 3, cols = 1, gridExpand = TRUE)
addStyle(wb, "Cox Model",
         createStyle(indent = 1),
         rows = level_rows + 3, cols = 1, gridExpand = TRUE)

# Bold rows with significant p-values
if (length(sig_rows) > 0) {
  addStyle(wb, "Cox Model",
           createStyle(textDecoration = "bold"),
           rows = sig_rows + 3, cols = 1:ncol(reg_df),
           gridExpand = TRUE, stack = TRUE)
}

# Number formatting — two decimals for HR, three for p-value
addStyle(wb, "Cox Model",
         createStyle(numFmt = "0.00", halign = "center"),
         rows = 4:(nrow(reg_df) + 3), cols = 2,
         gridExpand = TRUE, stack = TRUE)
addStyle(wb, "Cox Model",
         createStyle(numFmt = "0.000", halign = "center"),
         rows = 4:(nrow(reg_df) + 3), cols = 4,
         gridExpand = TRUE, stack = TRUE)
addStyle(wb, "Cox Model",
         createStyle(halign = "center"),
         rows = 4:(nrow(reg_df) + 3), cols = 3,
         gridExpand = TRUE, stack = TRUE)

# Footnote
footnote_row <- nrow(reg_df) + 4
writeData(wb, "Cox Model",
          "Abbreviations: CI = Confidence Interval, HR = Hazard Ratio",
          startRow = footnote_row)
addStyle(wb, "Cox Model",
         createStyle(fontSize = 10, textDecoration = "italic"),
         rows = footnote_row, cols = 1)

setColWidths(wb, "Cox Model", cols = 1:ncol(reg_df), widths = "auto")
freezePane(wb, "Cox Model", firstActiveRow = 4)

saveWorkbook(wb, "../data/tableE4_R09_cox_model_binarized_sugar_density.xlsx",
             overwrite = TRUE)

```


## pt characteristics

```{r}
final <- patient_summary %>% 
  select(age, sex, Disease , source,  intensity, gvhd_ppx, SugarCal_cat_high , day_exposed) %>% 
  mutate(Disease = case_when(
    Disease == 'NHL' ~ "Non-Hodgkin's lymphoma",
    # Myelodysplastic/myeloproliferative neoplasms
    Disease == 'MDS/MPN' ~ 'MDS/MPN',
    Disease == 'AML' ~ 'Acute myeloid leukemia',
    Disease == 'ALL' ~ 'Acute lymphoid leukemia',
    Disease == 'CLL' ~ 'Chronic lymphocytic leukemia',
    Disease == 'CML' ~ 'Other',
    Disease == 'Hodgkins' ~ "Other",
    Disease == 'Myeloma' ~ 'Myeloma', 
    Disease == 'AA' ~ 'Other',
    Disease == 'other' ~ 'Other'
  )) %>% 
  mutate(source = case_when(
    source == 'Unmodified' ~ "Unmodified bone marrow or PBSC",
    source == 'Cord' ~ "Cord blood",
    source == 'TCD' ~ "T-cell depleted PBSC"
  )) %>% 
  mutate(sex = case_when(
    sex == 'F' ~ "Female",
    sex == 'M' ~ "Male"
  )) %>% 
  mutate(intensity = case_when(
    intensity == 'nonablative' ~ "Nonmyeloablative",
    intensity == 'ablative' ~ "Ablative",
    intensity == 'reduced' ~ "Reduced intensity"
  )) %>% 
  mutate(gvhd_ppx = case_when(
    gvhd_ppx == 'CNI-based' ~ "CNI-based",
    gvhd_ppx == 'PTCy-based' ~ "PTCy-based",
    gvhd_ppx == 'TCD' ~ "T-cell depleted PBSC"
  ))  %>% 
  mutate(intensity = fct_reorder(intensity, intensity, .fun=length, .desc = T),
         source = fct_reorder(source, source, .fun=length, .desc = T),
         Disease = fct_reorder(Disease, Disease, .fun=length, .desc = T),
         sex = fct_reorder(sex, sex, .fun=length, .desc = T)) %>% 
  rename(`Graft type` = source, 
         #`HCT-CI` = ci,
         `Intensity of conditioning regimen` = intensity ) %>% 
  rename(Sex = sex, Age = age, `GvHD prophylaxis` = gvhd_ppx,
         `Days exposed to broad-spectrum antibiotics` = day_exposed)
```

```{r}
# Build the summary as before
summary_tbl <- final |>
  tbl_summary(
    statistic = list(all_continuous() ~ "{median} ({IQR})",
                     all_categorical() ~ "{n} ({p}%)"),
    by = SugarCal_cat_high
  ) |>
  bold_labels()

# Pull out the data and figure out which rows are variable labels vs levels
# gtsummary marks label rows in the row_type column when you peek at $table_body
body <- summary_tbl$table_body
label_rows <- which(body$row_type == "label")  # these get bolded, no indent
level_rows <- which(body$row_type == "level")  # these get indented

# Convert to a tibble and strip the markdown asterisks/underscores from headers
summary_df <- summary_tbl |> as_tibble()
names(summary_df) <- names(summary_df) |>
  str_remove_all("\\*\\*") |>           # drop the ** bold markers
  str_replace_all("__", "")             # drop the __ italic markers

# Clean the label cells too — gtsummary wraps them in __ for italics
summary_df[[1]] <- summary_df[[1]] |> str_remove_all("__")

# Now build the workbook with real formatting
wb <- createWorkbook()
addWorksheet(wb, "Patient Summary")

# Title row
writeData(wb, "Patient Summary",
          "Patient Characteristics by Averaged Sugar Intake",
          startRow = 1)
addStyle(wb, "Patient Summary",
         createStyle(textDecoration = "bold", fontSize = 14),
         rows = 1, cols = 1)

# Write the table starting at row 3
writeData(wb, "Patient Summary", summary_df, startRow = 3)

# Header row styling — bold, light gray fill, bottom border
header_style <- createStyle(textDecoration = "bold",
                            fgFill = "#D9D9D9",
                            border = "bottom",
                            halign = "left")
addStyle(wb, "Patient Summary", header_style,
         rows = 3, cols = 1:ncol(summary_df), gridExpand = TRUE)

# Bold the variable label rows (offset by 3: title + header + 1-indexed)
bold_style <- createStyle(textDecoration = "bold")
addStyle(wb, "Patient Summary", bold_style,
         rows = label_rows + 3, cols = 1, gridExpand = TRUE)

# Indent the level rows so categories visually nest under their variable
indent_style <- createStyle(indent = 1)
addStyle(wb, "Patient Summary", indent_style,
         rows = level_rows + 3, cols = 1, gridExpand = TRUE)

# Auto-fit columns and freeze the header
setColWidths(wb, "Patient Summary",
             cols = 1:ncol(summary_df), widths = "auto")
freezePane(wb, "Patient Summary", firstActiveRow = 4)

saveWorkbook(wb, "../data/tableE3_R09_sugar_density_patient_summary_binarized.xlsx",
             overwrite = TRUE)
```



## add caloric intake to the above model 

```{r}
# caloric intake is moderate to highly correlated with the sugar denstiy, and now there are way too many vars for way too few events , so overfit is very likely
fit_cal <- coxph(Surv(OStime_30,OSevent)~ SugarCal_cat_high * day_exposed + 
                   avg_total_caloric_intake +
                   source_and_gvhdppx + intensity ,
             data=patient_summary |> filter(OStime_30 > 0))
# good the p for day_exposed is still significant! the results basically is the same!!!!

### forest 
#  pull the per-group log HR for antibiotic exposure
# directly from the coefficient vector and variance-covariance matrix.
# Above-median sugar is the reference, so its effect is `day_exposed` alone.
# Below-median sugar's effect is `day_exposed` + the interaction term.
vc <- vcov(fit_cal)  
b  <- coef(fit_cal)

above_est <- b["day_exposed"]
above_se  <- sqrt(vc["day_exposed", "day_exposed"])

below_est <- b["day_exposed"] + b["SugarCal_cat_highBelow-median:day_exposed"]
below_se  <- sqrt(
  vc["day_exposed", "day_exposed"] +
  vc["SugarCal_cat_highBelow-median:day_exposed",
     "SugarCal_cat_highBelow-median:day_exposed"] +
  2 * vc["day_exposed", "SugarCal_cat_highBelow-median:day_exposed"]
)

forest_df <- tibble(
  group    = c("above-median\nsugar intake\ng/1000 kcal",
               "below-median\nsugar intake\ngrams/1000Kcal"),
  estimate = c(above_est, below_est),
  se       = c(above_se,  below_se)
) |>
  mutate(
    conf.low  = estimate - 1.96 * se,
    conf.high = estimate + 1.96 * se,
    # Exponentiate to get HR and its CI bounds
    HR        = exp(estimate),
    HR.low    = exp(conf.low),
    HR.high   = exp(conf.high),
    group     = fct_inorder(group) |> fct_rev()
  )

ggplot(forest_df, aes(x = HR, y = group)) +
  # Null effect is HR = 1, not 0
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey60") +
  geom_errorbarh(aes(xmin = HR.low, xmax = HR.high),
                 height = 0, linewidth = 0.5) +
  geom_point(shape = 22, fill = "black", color = "black", size = 2) +
  # Log scale keeps the CIs visually symmetric around the point estimate
  scale_x_log10() +
  labs(x = "HR per day\nof antibiotic exposure", y = NULL) +
  theme_classic(base_size = 12)

ggsave('../data/R09_forest_fit_cal_sugar_density_HR.pdf', width = 3, height = 2)
```



## KM survival curve

```{r}
# # 1. Prepare the data for plotting
# # We need to bin the continuous antibiotic variable to visualize it.
# # We'll use the median as the cutoff (or you can set a hard number like 8 days).
# km_data <- patient_summary |> 
#   filter(OStime_30 > 0) |> 
#   mutate(
#     # Create binary Abx group (split by median)
#     abx_cat = ifelse(day_exposed > median(day_exposed, na.rm = TRUE), 
#                      "High Abx", "Low Abx"),
#     
#     # Ensure Sugar variable is nicely labeled for the legend
#     sugar_label = ifelse(SugarCal_cat_high == "High" , 
#                          "High Density", "Low Density"),
#     
#     # Create the 4-level Interaction Stratification
#     # This creates the distinct groups for the lines
#     interaction_group = paste(sugar_label, abx_cat, sep = " + ")
#   )
# 
# # 2. Fit the Kaplan-Meier Object
# km_fit <- survfit(Surv(OStime_30, OSevent) ~ interaction_group, data = km_data)
# 
# # 3. Generate the Plot
# ggsurvplot(
#   km_fit,
#   data = km_data,
#   size = 1.5,                 # Line thickness
#   palette = c("darkorange", "#E7B800", "blue", "#2E9FDF"), # Colors (Gold for High Sugar, Blue for Low)
#   linetype = c("solid", "dashed", "solid", "dashed"),      # Dashed for Low Abx, Solid for High Abx
#   
#   # Aesthetics
#   pval = TRUE,                # Add p-value (Log-rank)
#   conf.int = FALSE,           # Turn off confidence intervals for cleaner look (optional)
#   risk.table = TRUE,          # Add risk table at bottom
#   risk.table.col = "strata",  # Color risk table by groups
#   legend.labs = c("High Density + High Abx", "High Density + Low Abx",  "Low Density + High Abx",  "Low Density + Low Abx"),
#   
#   # Labels
#   title = "Overall Survival by Sugar Density & Antibiotic Exposure",
#   xlab = "Days after HCT",
#   ylab = "Survival Probability"
# )
# ggsave('../data/R09_KM_curve.png', width = 7, height = 5)
```



```{r}
# problems with the length of stay clinical outcome 
# Reverse Causality (The "Sick Behavior" Confounder): The strongest counter-argument is that patients who are clinically doing worse (more nausea, pain, slower recovery) naturally default to a "comfort food" diet (high sugar, low intake) because they feel terrible. In this scenario, the diet isn't causing the longer stay; the underlying severity of their condition is causing both the poor diet and the longer stay. While you adjusted for conditioning intensity, unmeasured factors (like mucositis severity beyond opioid use) could drive this.
# 
# 
# The "Low Intake" vs. "Sugar" problem: Cluster 1 is defined by both low intake and high sugar. A critic could argue that the delayed discharge is simply due to malnutrition (caloric deficit) rather than the specific presence of sugar or the microbiome bloom. The "sugar" signal is intertwined with the "starvation" signal.
# 
# Discharge is Subjective: Unlike death or engraftment, "discharge" is somewhat subjective. A cautious attending physician might keep a patient who eats poorly (Cluster 1) in the hospital longer purely for nutritional monitoring, regardless of their actual immune recovery or infection status.
```

# 2d contour map 

This code gives you a 2D contour of log-HR on a grid of sugar intake per 1000kcal and abx exposure. Note that I used TCD + non-ablative as the reference group, which you can easily change. Changing the reference group will shift the scale of the contours, while the shape of the contours should be the same. 

In the plot below, the red dashed curve represents 0 log-HR - this means with combinations of abx and sugar intake on the red line, the risk was the same as taking no sugar at all + no abx at all. This is more like a reference line defined by numerical values rather than the reality, because we know nobody took 0 sugar during the treatment. 

What we can say is, the combinations corresponding to the yellow-ish contours reflected a higher risk of death, while the blue-ish contours reflected a lower risk of death. The most intense yellow zone happened to be high abx exposure + reasonably high sugar intake per 1000kcal. I believe this shows a reasonable interaction in a continuous manner which is consistent with what we claimed previously.

You can also find some meshed areas, which represent areas where the 95% confidence intervals didn't cover the reference log-HR level of zero. You can use that to claim statistical significance for the high sugar + high abx region and the low sugar + low abx region. 

What you may add in addition is that the median value of sugar intake per 1000kcal happened to be the grey horizontal line. This probably helped us to identify the abx effect for the high sugar group in our previous analyses because the most contrastive contours happened to be above the grey line. (You may choose not to say this to avoid further complexity in explaining this because it also depends on sampling distribution)

Let me know if you want to have a call to discuss this. 

@Dai, Angel One thing you may try is to adjust the range of sugar intake per 1000kcal for this contour plot. We have very few patients who took more than 150 grams per 1000 kcal, so I set a range between 23 to 150. This may be further tuned to hide some dark blue areas in the top which are likely to be driven by the outliers.

```{r}
library(splines)

# Read the dataset. 
df_use <- patient_summary

# Fit the Cox Proportional Hazards model.
# Using a natural spline (ns) allows the relationship between sugar and mortality to be non-linear.
# Multiplying by day_exposed tests if the sugar effect changes based on antibiotic duration.
# Filtering out patients who died on day 0 to avoid immortal time bias.
fit <- coxph(
  Surv(OStime_30, OSevent) ~ 
    ns(avg_sugar_density_per_1000kcal, knots = quantile(df_use$avg_sugar_density_per_1000kcal, probs = c(0.1, 0.9))) * day_exposed + 
    source_and_gvhdppx + 
    intensity,
  data = df_use |> filter(OStime_30 > 0)
)

# Output a clean summary table of the model's hazard ratios
fit |> 
  tbl_regression(exponentiate = FALSE) |> 
  bold_p() |> 
  as_gt() %>%
  gt::gtsave(filename = "../data/R09_contour_model_res.png") 

# Create a grid of hypothetical scenarios to map out the risk landscape.
# This holds the clinical variables constant while varying sugar and antibiotics.
grid_df <- expand.grid(
  day_exposed = seq(0, 20, length.out = 100),
  avg_sugar_density_per_1000kcal = seq(23, 150, length.out = 100),
  source_and_gvhdppx = "TCD+TCD",
  intensity = "nonablative"
)

# Generate risk predictions (log-hazard) for every scenario in the grid
pred <- predict(fit, newdata = grid_df, type = "lp", se.fit = TRUE)

# Convert the log-hazards into real-world Hazard Ratios using exp()
# On the log scale, 0 is neutral. On the HR scale, 1 is neutral.
grid_df <- grid_df |> 
  mutate(
    lp = pred$fit,
    se = pred$se.fit,
    hr = exp(lp), # This is the crucial transformation
    
    # Calculate confidence bounds on the log scale first, then exponentiate them.
    # Doing it in this order ensures mathematical accuracy.
    lower_hr = exp(lp - 1.96 * se),
    upper_hr = exp(lp + 1.96 * se),
    
    # Check significance against 1 (the new neutral line)
    significant = case_when(
      lower_hr > 1 ~ "Significantly > 1 (Harmful)",
      upper_hr < 1 ~ "Significantly < 1 (Protective)",
      TRUE         ~ "Not significant"
    )
  )

# Draw the contour map using the transformed HR values
ggplot(grid_df, aes(x = day_exposed, y = avg_sugar_density_per_1000kcal)) +
  # Custom breaks prevent the color palette from being stretched by massive HR values.
  # Anything above 5 gets grouped into a single color.
  geom_contour_filled(
    aes(z = hr),
    breaks = c(0, 0.5, 0.8, 1, 1.5, 2, 3, 5, Inf)
  ) +
  geom_contour(aes(z = hr), breaks = 1, color = "red", linewidth = 1, linetype = 2) +
  geom_tile(aes(color = significant), fill = NA, alpha = 0.25) +
  geom_segment(
    y = median(df_use$avg_sugar_density_per_1000kcal, na.rm = TRUE),
    x = 0, xend = 20, color = "grey"
  ) + 
  # Using a cleaner viridis color palette for the background hazard ratios
  scale_fill_viridis_d(option = "mako", direction = -1, name = "Contours of\nHazard Ratio") +
  scale_color_manual(
    name = "Significance based on\n95% confidence interval",
    values = c("Not significant" = NA, "Significantly > 1 (Harmful)" = "maroon", "Significantly < 1 (Protective)" = "forestgreen")
  ) +
  theme_classic() + 
  xlab("Days of broad-spectrum antibiotics exposure") + 
  ylab("Average grams of sugar intake per 1000Kcal\nbetween day -7 and day 12") 
```

pspline(df=2) uses an algorithm to automatically find the best smooth curve by penalizing wild wiggles. ns() requires you to drop physical anchor points (knots) for the curve to bend around, and mathematically forces the extreme ends of the graph to be straight lines to prevent erratic predictions at the boundaries. Using ns() turns your data into standard, predictable math equations, allowing R's predict() function to easily map out risks and confidence intervals for hypothetical scenarios. pspline() creates a highly customized curve tied to the original data, making it notoriously difficult for R to calculate standard errors when generating a new 2D map.