---
title: "The PROPHYLACTIC abx and empirical abx"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(cowplot) 
library(ggtext)
```

bar plot by day. For each day what % of samples being exposed in the three categories

1. No antibacterial whatsoever
2. ciprofloxacin only (the two fluoroquinolones that we use for these patients are cipro and levo.  mostly cipro)
3. Any Broad-spectrum (PO vanco, metronidazole, pip-tazo, cefepime, or ‘penem drugs). If still on cipro but getting a broad spectrum, classify as broad-spectrum-exposed.
4. other anti-bacterial

```{r}
meta <- read_csv('../data/153_combined_META.csv')

# loading the cleaned med data
med <- read_csv('../data/191_medication_cleaned_all_with_type.csv')

# copy the above table to be supp table upload folder 
```

# plotting

```{r}
# properly annotate these samples into groups
# total number of samples per day 
total_by_day <- meta %>% 
  select(pid, sampleid, sdrt) %>% 
  group_by(sdrt) %>% 
  summarise(daytotal = n())

exposure_grps <- med %>% 
  distinct(sampleid, med_exposure_type, sdrt) %>% 
  count(sdrt, med_exposure_type) %>% 
  inner_join(total_by_day) %>% 
  mutate(perc = round(n/daytotal*100, 2)) %>% 
  mutate(med_exposure_type = factor(med_exposure_type, levels = c('broad_spectrum','fluoroquinolones','other_antibacterial','no_abx')))
  
exposure_grps %>% count(med_exposure_type) 
```

```{r}
# plot them in percentage as a stacked bar plot 
library(RColorBrewer)
perc_ <- ggplot(exposure_grps, aes(x = sdrt, y = perc, fill = med_exposure_type)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c( "tomato", "steelblue","goldenrod3","gray"))  +
  geom_vline( xintercept =  0 , color = 'black', linetype = 'dashed', size = 1) +
  labs(x = "Transplant day", y = "Percentage", fill = "Exposure group") +
  theme_minimal() 

# I see so if I plot stat = identity it will be percentage in the bar plot 
# also plot this in absolute numbers
count_ <- ggplot(exposure_grps, aes(x = sdrt, y = n, fill = med_exposure_type)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = c( "tomato", "steelblue","goldenrod3","gray"))  +
    geom_vline( xintercept =  0 , color = 'black', linetype = 'dashed', size = 1) +
  labs(x = "Transplant day", y = "Count", fill = "Exposure group") +
  theme_minimal()
```


```{r}
title <- ggdraw() + 
  draw_label(
    "Fig. S3",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )


plot_ <- plot_grid(count_, perc_, nrow = 2, align = 'vh', axis = 'lrtb', labels = c('A','B'))

 
combined <- plot_grid(
  title, plot_,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.1, 1)
)+theme(plot.margin = unit(c(1,1,5,1), "cm"))

ggsave('../data/14_group4_antibiotics.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)  

```


```{r}
# to just put the broad spectrum together with the fluoroquinolones
library(tidyverse)
library(brms)   
library(ggpubr)
library(tidybayes)
library(cowplot)
library(ggridges)
library(brmstools)
library(bayesplot)
options(mc.cores = parallel::detectCores())
rstan::rstan_options(auto_write = TRUE)
theme_set(theme_tidybayes() + panel_border())
ncores <- parallel::detectCores()
axis_text_size <- 10
axis_title_size <- 10
zero_line_size <- .8
```


```{r}
exposure_grps <- med %>% 
  distinct(sampleid, med_exposure_type, sdrt)

# how many % of the samples had the abx if to put the broad_spectrum, fluoroquinolones all considered as the abx exposure btween 0 and 12
two_time <-  exposure_grps %>% mutate(two = if_else(med_exposure_type %in% c('broad_spectrum','fluoroquinolones'), T, F)) %>% 
  filter(sdrt %in% 0:12)


two_time %>% 
  count(two) %>% 
  mutate(perc = round(n/sum(n)*100, 2))
```


```{r}
two <- exposure_grps %>% filter(med_exposure_type %in% c('broad_spectrum','fluoroquinolones')) %>% pull(sampleid)
two %>% write_csv('../data/14_two_abx.csv')

meta <- read_csv('../data/153_combined_META.csv') %>% 
  mutate(timebin = cut_width(sdrt, 7, boundary=0, closed = 'left')) %>% 
  mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>% 
  mutate(pid = factor(pid)) %>% 
  mutate(fg_egg = fg_egg/100,
         fg_fruit = fg_fruit/100,
         fg_grain = fg_grain/100,
         fg_legume = fg_legume/100,
         fg_meat = fg_meat/100,
         fg_milk = fg_milk/100,
         fg_oils = fg_oils/100,
         fg_sweets = fg_sweets/100,
         fg_veggie = fg_veggie/100) %>% 
  mutate(abx = if_else(empirical == 'TRUE', 1, 0),
         TPN = if_else(TPN == 'TRUE', 1, 0),
         EN = if_else(EN == 'TRUE', 1, 0)) %>% 
  mutate(      fg_fruit_e= fg_fruit*abx,  
                fg_meat_e=fg_meat*abx,
                fg_milk_e=fg_milk*abx,
                fg_oils_e=fg_oils*abx,
                fg_egg_e=fg_egg*abx,
                fg_grain_e=fg_grain*abx,
                fg_sweets_e=fg_sweets*abx,
                fg_legume_e=fg_legume*abx,
                fg_veggie_e = fg_veggie*abx) %>% 
  filter(sampleid %in% two) %>% 
  filter(!sampleid %in% empirical_and_fq)

meta %>% count(abx)
```
```{r}
div_model  <- log(simpson_reciprocal) ~ 0 +
                intensity+
               fg_fruit_e+
                fg_meat_e+
                fg_milk_e+
                fg_oils_e+
                fg_egg_e+
                fg_grain_e+
                fg_sweets_e+
                fg_legume_e+
                fg_veggie_e+
                abx+
               fg_fruit+
                fg_meat+
                fg_milk+
                fg_oils+
                fg_egg+
                fg_grain+
                fg_sweets+
                fg_legume+
                fg_veggie+
                TPN+
                EN+
               (1 | pid) +
                (1 | timebin)
get_prior(div_model,data = meta )

div_priors <- c(# for the food group variables
            prior(normal(0, 1), class = 'b', coef = "fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "fg_veggie"),
            # interaction terms
            prior(normal(0, 1), class = 'b', coef = "fg_egg_e"),
            prior(normal(0, 1), class = 'b', coef = "fg_fruit_e"),
            prior(normal(0, 1), class = 'b', coef = "fg_grain_e"),
            prior(normal(0, 1), class = 'b', coef = "fg_legume_e"),
            prior(normal(0, 1), class = 'b', coef = "fg_meat_e"),
            prior(normal(0, 1), class = 'b', coef = "fg_milk_e"),
            prior(normal(0, 1), class = 'b', coef = "fg_oils_e"),
            prior(normal(0, 1), class = 'b', coef = "fg_sweets_e"),
            prior(normal(0, 1), class = 'b', coef = "fg_veggie_e"),
            # for the TPN
            prior(normal(0, 0.1), class = 'b', coef = "TPN"),
            # for the EN
            prior(normal(0, 0.1), class = 'b', coef = "EN"),
            # for the empirical
            prior(normal(0, 0.5), class = 'b', coef = "abx"),
            # for the intensity
            prior(normal( 2, .1), class = 'b', coef = "intensityablative"),
            prior(normal( 2, .1), class = 'b', coef = "intensityreduced"),
            prior(normal( 2, .1), class = 'b', coef = "intensitynonablative")
            )

model_div <- brm( div_model,  
              data = meta, 
              warmup = 1000, iter = 3000, 
              prior = div_priors,
              cores = ncores, 
              chains = 2, 
               control = list(adapt_delta = 0.99),
              seed = 123, sample_prior = T) 

# actually draw samples from the priors 
samples_prior <- prior_draws(model_div)
# save it for future use
post_res <- suppressWarnings(posterior_samples(model_div)) 
post_res %>%  write_csv('../data/14_div_model_interaction_two_abx.csv')
```


```{r}
fg_order <- c( 'Vegetables','abx * Vegetables',
                 'Oils','abx * Oils',
                 'Fruits','abx * Fruits',
                 'Meats','abx * Meats',
                 'Legumes','abx * Legumes',
                 'Eggs','abx * Eggs',
                 'Milk','abx * Milk',
                 'Grains','abx * Grains',
                 'Sweets','abx * Sweets',
                 "TPN" ,"EN" , 'abx')
key <- read_csv('../data/food_group_color_key_final.csv', col_types = 'ccccc')

cleaned_fg <- post_res  %>% select(starts_with('b_')) %>% 
  gather('item', 'coeff') %>% 
  mutate(item = str_replace(item, 'b_fg_',''),
         item = str_replace(item, 'b_',''),
         item = str_replace(item, 'intensity','')) %>% 
   mutate(fgrp1 = case_when(
    item ==  'milk' ~ '1',
    item == 'meat' ~ '2',
    item ==  'egg' ~ '3',
    item ==  'legume' ~ '4',
    item == 'grain' ~ '5',
    item == 'fruit' ~ '6',
    item == 'veggie' ~ '7',
    item ==  'oils' ~ '8', 
    item ==  'sweets' ~ '9'
  ))  %>% 
  left_join(key %>% select(fgrp1, color, shortname)) %>% 
  mutate(shortname = case_when(
        item ==  'milk_e' ~ 'abx * Milk',
    item == 'meat_e' ~ 'abx * Meats',
    item ==  'egg_e' ~ 'abx * Eggs',
    item ==  'legume_e' ~ 'abx * Legumes',
    item == 'grain_e' ~ 'abx * Grains',
    item == 'fruit_e' ~ 'abx * Fruits',
    item == 'veggie_e' ~ 'abx * Vegetables',
    item ==  'oils_e' ~ 'abx * Oils', 
    item ==  'sweets_e' ~ 'abx * Sweets',
    item ==  'nonablative' ~ 'Nonablative',
    item ==  'reduced' ~ 'Reduced',
    item ==  'ablative' ~ 'Ablative',
    item ==  'TPN' ~ 'TPN',
    item ==  'EN' ~ 'EN',
    item ==  'abx' ~ 'abx',
    TRUE ~ `shortname`
  )) %>% 
  mutate(grp = if_else(item %in% c('nonablative','reduced','ablative'), 'patient', 'temporal')) 

# find the credible interval of the abx line
cleaned_fg %>% filter(str_detect(shortname, 'abx')) %>% 
  group_by(shortname) %>% 
  tidybayes::median_qi(coeff , .width = c( .95)) %>% 
  mutate(coeff = round(coeff, 2),
         .lower = round(.lower, 2),
         .upper = round(.upper, 2))

fg_label <- tibble(ord = fg_order) %>% 
  mutate(shortname = str_replace(ord, 'abx \\+ ','')) %>% 
  left_join(key %>% select(shortname, color)) %>% 
  mutate(color = if_else(is.na(color), 'black', color))

interaction_terms <- tibble(
  shortname = fg_label$ord
) %>% 
  mutate(ymin = seq(1, nrow(.)) - 0.5,
         ymax = seq(1, nrow(.)) + 0.5) %>% 
  filter(str_detect(shortname, '\\*')) %>% 
  mutate(xmin = -Inf, xmax = Inf,
         interaction_fill = 'interaction_fill')

# this leaves me no choice but I need to fill the other blue area with the blue rectangles 
blue_terms <- tibble(
  shortname = fg_label$ord
) %>% 
  mutate(ymin = seq(1, nrow(.)) - 0.5,
         ymax = seq(1, nrow(.)) + 0.5) %>% 
  filter(!str_detect(shortname, '\\*')) %>% 
  mutate(xmin = -Inf, xmax = Inf,
         blue_fill = 'blue_fill')

cross0 <- cleaned_fg %>%
   filter(grp == 'temporal') %>% 
  group_by(item) %>% 
  summarise(q2.5 = quantile(coeff, probs = 0.025),
            q97.5 = quantile(coeff, probs = 0.975)) %>% 
  mutate(Cross = if_else(q2.5 >= 0 | q97.5 <= 0, F, T))

fg_temporal <- cleaned_fg %>% 
  filter(grp == 'temporal') %>% 
  mutate(shortname = factor(shortname, levels = fg_label$ord)) %>% 
  left_join(cross0) %>% 
  ggplot() +
   stat_pointinterval(aes(x = coeff, y = shortname, col = Cross), .width = .95, fatten_point = 1.2) +
  geom_rect(data = interaction_terms, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = interaction_fill), alpha = .2) +
  geom_rect(data = blue_terms, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = blue_fill) , alpha = .25) +
  scale_fill_manual(values=c("interaction_fill" = "tan2", "blue_fill" = "#FAEFD1")) +
  scale_color_manual(values = c("#EC0000", "black")) +
  scale_y_discrete( labels = function(x) { ifelse(str_detect(x, '\\*'),  str_glue("<b style='color:royalblue'>{x}</b>"), as.character(x)) }) +
  geom_vline(xintercept = 0, col = 'blue', size = zero_line_size) +
  labs(x = 'ln(diversity) change', y = '') +
  theme_classic() +
  theme(legend.position = 'none') +
  theme(axis.text = element_text( size = 10),
        axis.text.y = element_markdown(),
        plot.background = element_rect(fill='transparent', color=NA), 
        axis.title=element_text(size=10), 
        aspect.ratio=1.5)      
fg_temporal   


```


```{r}
num_annot <- meta %>% 
  split(.$abx) %>% 
  imap_dfr(function(df, name_) {
    df %>% distinct(pid) %>% nrow
  }) %>% gather('abx','pt_num') %>% 
  full_join(meta %>% count(abx) %>% mutate(abx = as.character(abx))) %>% 
  rename(sample_num = n) %>% 
  mutate(antibiotics = if_else(abx == 0, 'not exposed', 'exposed')) %>% 
  mutate(lbl = str_glue('{sample_num} samples\nfrom {pt_num} patients'))


library(wesanderson) 

palette <- wes_palette("Royal1", 2)

# the raw data 
rawsweets <- meta %>% 
  mutate(log_div = log(simpson_reciprocal)) %>% 
  mutate(antibiotics = if_else(abx == 0, 'not exposed', 'exposed'),
         antibiotics = factor(antibiotics, levels = c('not exposed','exposed'))) %>% 
  ggscatter(x = 'fg_sweets', y = 'log_div', color = 'antibiotics',
            ylab = 'log(diversity)',xlab = 'Sweets consumed (grams)',
            alpha = 0.35, shape = 16, size = 1,
            add = "reg.line",  # Add regressin line
           add.params = list(color = 'antibiotics', fill = 'antibiotics', alpha = 0.3, size = 1.5), # Customize line
           conf.int = TRUE, # Add confidence interval
           cor.coef = TRUE, # Add correlation coefficient.
           cor.coeff.args = list(method = "spearman",  label.sep = "\n", cor.coef.name = c("rho"),p.accuracy = 0.01, r.accuracy = 0.01,label.x.npc = "left", label.y.npc = "bottom", size = 3.5)) +
  geom_text(data = num_annot, aes(label = lbl), 
              x = Inf, y = Inf, hjust = 0.6, vjust = 1.5) +
  scale_x_sqrt() +
  scale_y_sqrt() +
  scale_fill_manual('antibiotics', values = palette, labels=c('not exposed', 'exposed')) +
  scale_colour_manual('antibiotics',values = palette, labels=c('not exposed', 'exposed')) +
  facet_wrap(~ antibiotics, labeller = 'label_both', dir = 'h') +
  theme(aspect.ratio = 1/1.15, 
        #legend.position = 'none',
          strip.background = element_blank(),
        strip.text.x = element_blank(), 
     axis.text=element_text(size=9), axis.title=element_text(size=9)) 
rawsweets

ggsave('../data/14_two_abx.png', width = 10)
```


```{r}
# to filter from the empircial ones the only exposed to empirical ones
library(vdbR)
connect_database()
get_table_from_database('samples_castori_ag')

meta <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/153_combined_META.csv')
  
fluoroquinolones <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/169_all_medication.csv') %>% 
  # find the fq in this dataset
  filter(str_detect(together, 'ciprofloxacin|levofloxacin')) %>% 
  filter(!str_detect(together, 'otic')) %>% 
  mutate(med_int = interval(start_date, stop_date)) 


fluoro_exposed <- meta %>% filter(empirical == 'TRUE') %>%   
  select(sampleid, mrn) %>% 
  inner_join(samples_castori_ag %>% select(sampleid, datecollection)) %>% 
  mutate(p1d = datecollection -1,
         p2d = datecollection -2) %>% 
  mutate(p2d_int = interval(p2d, p1d)) %>% 
  select(sampleid, datecollection, mrn, p1d, p2d, p2d_int) %>% 
  inner_join(fluoroquinolones, by = "mrn") %>% 
  mutate(had_fluoroquinolones = int_overlaps(p2d_int, med_int)) %>% 
  filter(had_fluoroquinolones == 'TRUE') 
 
fluoro_exposed %>% distinct(sampleid)       

empirical_and_fq <- fluoro_exposed %>% distinct(sampleid)       %>% pull(sampleid)
# let's just remove these 89 from the empirical samples and run again 
```





