---
title: "The PROPHYLACTIC abx and empirical abx"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(cowplot) 
```

bar plot by day. For each day what % of samples being exposed in the three categories

1. No antibacterial whatsoever
2. ciprofloxacin only (the two fluoroquinolones that we use for these patients are cipro and levo.  mostly cipro)
3. Any Broad-spectrum (PO vanco, metronidazole, pip-tazo, cefepime, or ‘penem drugs). If still on cipro but getting a broad spectrum, classify as broad-spectrum-exposed.
4. other anti-bacterial

```{r}
meta <- read_csv('../data/153_combined_META.csv')

# loading the cleaned med data
med <- read_csv('../data/191_medication_cleaned_all_with_type.csv')
```

# plotting

```{r}
# properly annotate these samples into groups
# total number of samples per day 
total_by_day <- meta %>% 
  select(pid, sampleid, sdrt) %>% 
  group_by(sdrt) %>% 
  summarise(daytotal = n())

exposure_grps <- med %>% 
  distinct(sampleid, med_exposure_type, sdrt) %>% 
  count(sdrt, med_exposure_type) %>% 
  inner_join(total_by_day) %>% 
  mutate(perc = round(n/daytotal*100, 2)) %>% 
  mutate(med_exposure_type = factor(med_exposure_type, levels = c('broad_spectrum','fluoroquinolones','other_antibacterial','no_abx')))
  
exposure_grps
```

```{r}
# plot them in percentage as a stacked bar plot 
library(RColorBrewer)
perc_ <- ggplot(exposure_grps, aes(x = sdrt, y = perc, fill = med_exposure_type)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_manual(values = c( "tomato", "steelblue","goldenrod3","gray"))  +
  geom_vline( xintercept =  0 , color = 'black', linetype = 'dashed', size = 1) +
  labs(x = "Transplant day", y = "Percentage", fill = "Exposure group") +
  theme_minimal() 

# I see so if I plot stat = identity it will be percentage in the bar plot 
# also plot this in absolute numbers
count_ <- ggplot(exposure_grps, aes(x = sdrt, y = n, fill = med_exposure_type)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(values = c( "tomato", "steelblue","goldenrod3","gray"))  +
    geom_vline( xintercept =  0 , color = 'black', linetype = 'dashed', size = 1) +
  labs(x = "Transplant day", y = "Count", fill = "Exposure group") +
  theme_minimal()
```


```{r}
title <- ggdraw() + 
  draw_label(
    "Fig. S3",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )


plot_ <- plot_grid(count_, perc_, nrow = 2, align = 'vh', axis = 'lrtb', labels = c('A','B'))

 
combined <- plot_grid(
  title, plot_,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.1, 1)
)+theme(plot.margin = unit(c(1,1,5,1), "cm"))

ggsave('../data/14_group4_antibiotics.pdf',
      width = 210, height = 297, units = "mm", device = 'pdf', plot = combined, dpi = 300)  

```











