---
title: "FQ battel"
output: html_document
date: "2025-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(brms)   
library(ggpubr)
library(tidybayes)
library(cowplot)
library(brmstools)
library(bayesplot)
rstan::rstan_options(auto_write = TRUE)
theme_set(theme_tidybayes() + panel_border())
```


```{r}
library(vdbR)
connect_database()
list_table_from_database('mmf')
get_table_from_database('samples_mmf')
```

```{r}
med <- read_csv('../data/191_medication_cleaned_all_with_type.csv') %>% distinct(sampleid, med_exposure_type)

fq_exposed <- med %>% filter(med_exposure_type =='fluoroquinolones') %>% pull(sampleid)

fg_and_other <- med %>% filter(med_exposure_type =='fluoroquinolones' | med_exposure_type == 'other_antibacterial') %>% pull(sampleid)
```

# no interaction

## diversity

```{r}
# meta <- read_csv('../data/153_combined_META.csv') %>% 
#   mutate(timebin = cut_width(sdrt, 7, boundary=0, closed = 'left')) %>% 
#   mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>% 
#   mutate(pid = factor(pid)) %>% 
#   mutate(fg_egg = fg_egg/100,
#          fg_fruit = fg_fruit/100,
#          fg_grain = fg_grain/100,
#          fg_legume = fg_legume/100,
#          fg_meat = fg_meat/100,
#          fg_milk = fg_milk/100,
#          fg_oils = fg_oils/100,
#          fg_sweets = fg_sweets/100,
#          fg_veggie = fg_veggie/100) %>% 
#   mutate(abx = if_else(empirical == 'TRUE', 'broad', if_else(sampleid %in% fq_exposed, 'FQ','no'))) %>% 
#   mutate(abx = factor(abx, levels = c('no','FQ','broad')))

# div_model  <- log(simpson_reciprocal) ~ 0 +
#                 intensity+
#                 abx+
#                fg_fruit+
#                 fg_meat+
#                 fg_milk+
#                 fg_oils+
#                 fg_egg+
#                 fg_grain+
#                 fg_sweets+
#                 fg_legume+
#                 fg_veggie+
#                 TPN+
#                 EN+
#                (1 | pid) +
#                 (1 | timebin)
# 
# get_prior(div_model,  
#               data = meta)
```


```{r} 
# div_priors <- c(# for the food group variables
#             prior(normal(0, 1), class = 'b', coef = "fg_egg"),
#             prior(normal(0, 1), class = 'b', coef = "fg_fruit"),
#             prior(normal(0, 1), class = 'b', coef = "fg_grain"),
#             prior(normal(0, 1), class = 'b', coef = "fg_legume"),
#             prior(normal(0, 1), class = 'b', coef = "fg_meat"),
#             prior(normal(0, 1), class = 'b', coef = "fg_milk"),
#             prior(normal(0, 1), class = 'b', coef = "fg_oils"),
#             prior(normal(0, 1), class = 'b', coef = "fg_sweets"),
#             prior(normal(0, 1), class = 'b', coef = "fg_veggie"),
#        
#             # for the TPN
#             prior(normal(0, 0.1), class = 'b', coef = "TPNTRUE"),
#             # for the EN
#             prior(normal(0, 0.1), class = 'b', coef = "ENTRUE"),
#             # for the empirical
#             prior(normal(0, 0.5), class = 'b', coef = "abxbroad"),
#             prior(normal(0, 0.5), class = 'b', coef = "abxFQ"),
#             # for the intensity
#             prior(normal( 2, .1), class = 'b', coef = "intensityablative"),
#             prior(normal( 2, .1), class = 'b', coef = "intensityreduced"),
#             prior(normal( 2, .1), class = 'b', coef = "intensitynonablative")
#             )

# model_div <- brm( div_model,  
#               data = meta, 
#               warmup = 1000, iter = 3000, 
#               prior = div_priors,
#               cores = ncores, 
#               chains = 2, 
#                control = list(adapt_delta = 0.99),
#               seed = 123, sample_prior = T) 
# 
# # save it for future use
# post_res <- suppressWarnings(posterior_samples(model_div)) 
# post_res %>%  write_csv('../data/18_div_model_fg_post.csv')
```
```{r}
# post_res %>% 
#   select(starts_with('b')) %>% 
#   select(!contains('intensity')) %>% 
#   gather('shortname','coeff') %>% 
#   group_by(shortname) %>% # Group by shortname to calculate medians per group
#   mutate(median_coeff = median(coeff)) %>% # Calculate median for each shortname
#   ungroup() %>% # Important to ungroup after calculating the median
#   mutate(shortname = fct_reorder(shortname, median_coeff)) %>% 
#   ggplot() +
#    stat_pointinterval(aes(x = coeff, y = shortname), .width = .95, fatten_point = 1.2) +
#   geom_vline(xintercept = 0, col = 'blue', size = 0.5) 
```

## Enterococcus

```{r}
clr_res <- read_csv('../data/171_genus_CLR_res.csv')

clr_wide <- clr_res %>% 
  spread('genus','clr')

extra_full <- meta %>% 
  inner_join(clr_wide)  
# 
#   
# entero_mod <- Enterococcus ~ 0 +
#                 intensity+
#                 abx+
#                fg_fruit+
#                 fg_meat+
#                 fg_milk+
#                 fg_oils+
#                 fg_egg+
#                 fg_grain+
#                 fg_sweets+
#                 fg_legume+
#                 fg_veggie+
#                 TPN+
#                 EN+
#                (1 | pid) +
#                 (1 | timebin)
# get_prior(entero_mod,  
#               data = extra_full)
```


```{r}
priors <- c(# for the food group variables
            prior(normal(0, 1), class = 'b', coef = "fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "fg_veggie"),
            # for the TPN
            prior(normal(0, 0.1), class = 'b', coef = "TPNTRUE"),
            # for the EN
            prior(normal(0, 0.1), class = 'b', coef = "ENTRUE"),
            # for the empirical
            prior(normal(0, 0.5), class = 'b', coef = "abxbroad"),
            prior(normal(0, 0.5), class = 'b', coef = "abxFQ")
            )

# mod_entero =  brm(entero_mod,  
#                 data = extra_full, 
#               warmup = 1000, iter = 3000, 
#               prior = priors,
#               cores = ncores, 
#               chains = 2, 
#               control = list(adapt_delta = 0.99),
#               seed = 456)
# 
# # save it for future use
# post_res <- suppressWarnings(posterior_samples(mod_entero)) 
post_res %>%  write_csv('../data/18_entero_model_fg_post.csv')
```
     
```{r}
# post_res %>% 
#   select(starts_with('b')) %>% 
#   select(!contains('intensity')) %>% 
#   gather('shortname','coeff') %>% 
#   group_by(shortname) %>% # Group by shortname to calculate medians per group
#   mutate(median_coeff = median(coeff)) %>% # Calculate median for each shortname
#   ungroup() %>% # Important to ungroup after calculating the median
#   mutate(shortname = fct_reorder(shortname, median_coeff)) %>% 
#   ggplot() +
#    stat_pointinterval(aes(x = coeff, y = shortname), .width = .95, fatten_point = 1.2) +
#   geom_vline(xintercept = 0, col = 'blue', size = 0.5) 
```

# yes interaction

## diversity

```{r}
# to put the other antibacterial into the FQ group
      
meta <- read_csv('../data/153_combined_META.csv') %>% 
  mutate(timebin = cut_width(sdrt, 7, boundary=0, closed = 'left')) %>% 
  mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>% 
  mutate(pid = factor(pid)) %>% 
  mutate(fg_egg = fg_egg/100,
         fg_fruit = fg_fruit/100,
         fg_grain = fg_grain/100,
         fg_legume = fg_legume/100,
         fg_meat = fg_meat/100,
         fg_milk = fg_milk/100,
         fg_oils = fg_oils/100,
         fg_sweets = fg_sweets/100,
         fg_veggie = fg_veggie/100) %>% 
  full_join(exposures) %>% 
  filter(exposure_category %in% c('no_abx','fluoroquinolones')) %>% 
  mutate(abx = if_else(exposure_category == 'no_abx', '0_no_abx', '1_FQ'),abx = factor(abx)) %>% 
  #mutate(abx = if_else(empirical == 'TRUE', '2_broad', if_else(sampleid %in% fg_and_other, '1_fq_and_other','0_no_abx')),abx = factor(abx)) %>% 
  mutate(TPN = if_else(TPN == 'TRUE', 1, 0),
         EN = if_else(EN == 'TRUE', 1, 0)) 

num_annot <- meta %>% 
  split(.$abx) %>% 
  imap_dfr(function(df, name_) {
    df %>% distinct(pid) %>% nrow
  }) %>% gather('abx','pt_num') %>% 
  full_join(meta %>% count(abx) %>% mutate(abx = as.character(abx))) %>% 
  rename(sample_num = n) %>% 
  #mutate(antibiotics = if_else(empirical == 'FALSE', 'not exposed', 'exposed')) %>% 
  mutate(lbl = str_glue('{sample_num} samples\nfrom {pt_num} patients'))
```
 
```{r}
mod_original <- log(simpson_reciprocal) ~ 0 +        
                intensity+
                abx+
               fg_fruit:abx+
                fg_meat:abx+
                fg_milk:abx+
                fg_oils:abx+
                fg_egg:abx+
                fg_grain:abx+
                fg_sweets:abx+
                fg_legume:abx+
                fg_veggie:abx+
                TPN+
                EN+
               (1 | pid) +
                (1 | timebin)
get_prior(mod_original,data = meta )
```

```{r}
div_priors_original <- c(# for the food group variables 
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_veggie"),
            # interaction terms
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "abx1_fq_and_other:fg_veggie"),
                      # interaction terms
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "abx2_broad:fg_veggie"),
            # for the TPN
            prior(normal(0, 0.1), class = 'b', coef = "TPN"),
            # for the EN
            prior(normal(0, 0.1), class = 'b', coef = "EN"),
            # for the empirical
            prior(normal(0, 0.5), class = 'b', coef = "abx1_fq_and_other"),
            prior(normal(0, 0.5), class = 'b', coef = "abx2_broad"),
            # for the intensity
            prior(normal( 2, .1), class = 'b', coef = "intensityablative"),
            prior(normal( 2, .1), class = 'b', coef = "intensityreduced"),
            prior(normal( 2, .1), class = 'b', coef = "intensitynonablative")
            )

model_div_original <- brm( mod_original,  
              data = meta, 
              warmup = 1000, iter = 3000, 
              prior = div_priors_original,
              cores = ncores, 
              chains = 2, 
               control = list(adapt_delta = 0.99),
              seed = 123, sample_prior = T)

# save it for future use
post_res <- suppressWarnings(posterior_samples(model_div_original)) 
post_res %>%  write_csv('../data/18_model_div_original_three_abx_levels.csv')


```

```{r}
post_res %>% 
  select(starts_with('b')) %>% 
  select(!contains('intensity')) %>% 
  gather('shortname','coeff') %>% 
  group_by(shortname) %>% # Group by shortname to calculate medians per group
  mutate(median_coeff = median(coeff)) %>% # Calculate median for each shortname
  ungroup() %>% # Important to ungroup after calculating the median
  mutate(shortname = fct_reorder(shortname, median_coeff)) %>% 
  group_by(shortname) %>% # Group again to get intervals
  mutate(
    lower = quantile(coeff, 0.025), # Calculate lower bound of the interval
    upper = quantile(coeff, 0.975), # Calculate upper bound of the interval
    significant = ifelse(lower > 0 | upper < 0, "Significant", "Not Significant")
  ) %>%
  ungroup() %>%
  ggplot() +
    stat_pointinterval(
      aes(x = coeff, y = shortname, color = significant), # Color by significance
      .width = .95,
      fatten_point = 1.2
    ) +
    geom_vline(xintercept = 0, col = 'blue', size = 0.5) +
    scale_color_manual(values = c("Significant" = "red", "Not Significant" = "black")) + # Set colors
    labs(x = "Coefficient", y = "Shortname", color = "95% CI") + # Improved labels
    theme_bw()
ggsave('../data/18_three_abx.pdf', width = 7)
```
```{r}
library(wesanderson) 

dat_conditional <- conditional_effects(model_div_original, surface = T)

condi_dat <- dat_conditional %>% 
  keep(.p = str_detect(names(.), ':')) %>% 
  bind_rows(.id = 'grp')
```


```{r}
key <- read_csv('../data/food_group_color_key_final.csv', col_types = 'ccccc')
conditional <- condi_dat %>% 
  mutate(grp = str_replace(grp, '\\:empirical','')) %>% 
  left_join(key %>% select(grp = fg1_name, shortname)) %>% 
  mutate(shortname = factor(shortname, levels = rev( c('Vegetables',
                 'Oils',
                 'Fruits',
                 'Meats',
                 'Legumes',
                 'Eggs',
                 'Milk',
                 'Grains',
                 'Sweets'))))
palette <- wes_palette("Royal1", 4)[c(1,2,4)]
# the color for the strip background
strip_color <- wes_palette("Royal1", 3)[3]

conditional9 <- conditional %>%   
  ggplot() +
  geom_smooth(aes(x = effect1__, y = estimate__, ymin = lower__, ymax = upper__, fill = effect2__, color = effect2__),
              stat = "identity",
              alpha = .3, linewidth = 1.5)+ 
  scale_fill_manual('antibiotics', values = palette) +
  scale_colour_manual('antibiotics', values = palette) +
  facet_wrap(~ shortname, nrow = 2, scales = 'free_x') +
  ylim(0, 3.3) +
  labs(y = 'Predicted log(diversity)', x = 'Food group consumed (grams)') +
  theme_classic() +
  theme(legend.position = 'bottom', legend.title = element_text(size = 8, face="bold"),
        legend.text = element_text( size=8, face="bold"),
        panel.background = element_rect(fill = "#FAEFD140"),
        legend.background=element_rect(fill = alpha("white", 0)),
       legend.key=element_rect(fill = alpha("white", 0)),
        aspect.ratio = 1/1.5, 
        strip.background = element_rect(color="white", fill='#ffecdc', size=1.5,linetype="solid"),
     axis.text=element_text(size=8), axis.title=element_text(size=10))    +
  guides(fill = guide_legend(direction = "vertical"), color = guide_legend(direction = "vertical"))
conditional9
```

## Enterococcus 

```{r}
clr_res <- read_csv('../data/171_genus_CLR_res.csv')
clr_Enterococcus <- clr_res %>% 
  filter(genus %in% c('Enterococcus')) %>%
  spread('genus','clr')

meta <- read_csv('../data/153_combined_META.csv') %>% 
  mutate(timebin = cut_width(sdrt, 7, boundary=0, closed = 'left')) %>% 
  mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>% 
  mutate(pid = factor(pid)) %>% 
  inner_join(clr_Enterococcus)%>% 
  mutate(empirical = if_else(empirical == 'TRUE', '2_broad', if_else(sampleid %in% fg_and_other, '1_fq_and_other','0_no_abx')),
         empirical = factor(empirical)) 

entero_model_original <- Enterococcus ~ 0 +
             intensity+
                empirical+
               fg_fruit:empirical+
                fg_meat:empirical+
                fg_milk:empirical+
                fg_oils:empirical+
                fg_egg:empirical+
                fg_grain:empirical+
                fg_sweets:empirical+
                fg_legume:empirical+
                fg_veggie:empirical+
                TPN+
                EN+
               (1 | pid) +
                (1 | timebin)
get_prior(formula = entero_model_original, data = meta)
```


```{r}
priors_enterococcus_original <- c(# for the food group variables
         prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "empirical0_no_abx:fg_veggie"),
            # interaction terms
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "empirical1_fq_and_other:fg_veggie"),
                      # interaction terms
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "empirical2_broad:fg_veggie"),
            # for the TPN
            prior(normal(0, 0.1), class = 'b', coef = "TPNTRUE"),
            # for the EN
            prior(normal(0, 0.1), class = 'b', coef = "ENTRUE"),
            # for the empirical
            prior(normal(0, 0.5), class = 'b', coef = "empirical1_fq_and_other"),
            prior(normal(0, 0.5), class = 'b', coef = "empirical2_broad")
            
            )

mod_entero =  brm(entero_model_original,  
                data = meta, 
              warmup = 1000, iter = 3000, 
              prior = priors_enterococcus_original,
              cores = ncores, 
              chains = 2, 
              control = list(adapt_delta = 0.99),
              seed = 456)

post_res <- suppressWarnings(posterior_samples(mod_entero)) 
```
```{r}
post_res %>% 
  select(starts_with('b')) %>% 
  select(!contains('intensity')) %>% 
  gather('shortname','coeff') %>% 
  #filter(!shortname %in% c('b_empirical2_broad','b_empirical1_fq_and_other', 'b_ENTRUE', 'b_TPNTRUE')) %>% 
  group_by(shortname) %>% # Group by shortname to calculate medians per group
  mutate(median_coeff = median(coeff)) %>% # Calculate median for each shortname
  ungroup() %>% # Important to ungroup after calculating the median
  mutate(shortname = fct_reorder(shortname, median_coeff)) %>% 
  group_by(shortname) %>% # Group again to get intervals
  mutate(
    lower = quantile(coeff, 0.025), # Calculate lower bound of the interval
    upper = quantile(coeff, 0.975), # Calculate upper bound of the interval
    significant = ifelse(lower > 0 | upper < 0, "Significant", "Not Significant")
  ) %>%
  ungroup() %>%
  ggplot() +
    stat_pointinterval(
      aes(x = coeff, y = shortname, color = significant), # Color by significance
      .width = .95,
      fatten_point = 1.2
    ) +
    geom_vline(xintercept = 0, col = 'blue', size = 0.5) +
    scale_color_manual(values = c("Significant" = "red", "Not Significant" = "black")) + # Set colors
    labs(x = "Coefficient", y = "Shortname", color = "95% CI") + # Improved labels
    theme_bw()
```
# reclassifiy the sample exposure type using the newly cleaned med data

```{r}
exposures <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/191_all_medication_overlapping_with_prior_2_day_of_stool_samples_upated.csv') %>%    group_by(sampleid, sdrt, pid) %>%
  summarize(
    exposure_category = case_when(
      any(drug_category_for_this_study == "broad_spectrum") ~ "broad_spectrum",
      any(drug_category_for_this_study == "fluoroquinolones") ~ "fluoroquinolones",
      any(drug_category_for_this_study == "other_antibacterials") ~ "other_antibacterial",
      TRUE ~ "no_abx"  # Default case if neither "apple" nor "orange" is found
    )
  ) 
 
exposures %>% count(exposure_category!= med_exposure_type)

no_abx_and_fq <- exposures %>% 
  filter(exposure_category %in% c('fluoroquinolones','no_abx'))


```

```{r}
expodat <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/191_all_medication_overlapping_with_prior_2_day_of_stool_samples_upated_2.csv') 

```

# no_abx VS FQ


```{r}
meta <- read_csv('../data/153_combined_META.csv') %>% 
  mutate(timebin = cut_width(sdrt, 7, boundary=0, closed = 'left')) %>% 
  mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>% 
  mutate(pid = factor(pid)) %>% 
  mutate(fg_egg = fg_egg/100,
         fg_fruit = fg_fruit/100,
         fg_grain = fg_grain/100,
         fg_legume = fg_legume/100,
         fg_meat = fg_meat/100,
         fg_milk = fg_milk/100,
         fg_oils = fg_oils/100,
         fg_sweets = fg_sweets/100,
         fg_veggie = fg_veggie/100) %>% 
  full_join(exposures) %>% 
  filter(exposure_category %in% c('no_abx','fluoroquinolones')) %>% 
  mutate(abx = if_else(exposure_category == 'no_abx', '0_no_abx', '1_FQ'),abx = factor(abx)) %>% 
  #mutate(abx = if_else(empirical == 'TRUE', '2_broad', if_else(sampleid %in% fg_and_other, '1_fq_and_other','0_no_abx')),abx = factor(abx)) %>% 
  mutate(TPN = if_else(TPN == 'TRUE', 1, 0),
         EN = if_else(EN == 'TRUE', 1, 0)) 

num_annot <- meta %>% 
  split(.$abx) %>% 
  imap_dfr(function(df, name_) {
    df %>% distinct(pid) %>% nrow
  }) %>% gather('abx','pt_num') %>% 
  full_join(meta %>% count(abx) %>% mutate(abx = as.character(abx))) %>% 
  rename(sample_num = n) %>% 
  #mutate(antibiotics = if_else(empirical == 'FALSE', 'not exposed', 'exposed')) %>% 
  mutate(lbl = str_glue('{sample_num} samples\nfrom {pt_num} patients'))

mod_original <- log(simpson_reciprocal) ~ 0 +        
                intensity+
                abx+
               fg_fruit:abx+
                fg_meat:abx+
                fg_milk:abx+
                fg_oils:abx+
                fg_egg:abx+
                fg_grain:abx+
                fg_sweets:abx+
                fg_legume:abx+
                fg_veggie:abx+
                TPN+
                EN+
               (1 | pid) +
                (1 | timebin)
get_prior(mod_original,data = meta )
```


```{r}
div_priors_original <- c(# for the food group variables 
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "abx0_no_abx:fg_veggie"),
            # interaction terms
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_egg"),
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_fruit"),
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_grain"),
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_legume"),
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_meat"),
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_milk"),
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_oils"),
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_sweets"),
            prior(normal(0, 1), class = 'b', coef = "abx1_FQ:fg_veggie"),
    
            # for the TPN
            prior(normal(0, 0.1), class = 'b', coef = "TPN"),
            # for the EN
            prior(normal(0, 0.1), class = 'b', coef = "EN"),
            # for the empirical
            prior(normal(0, 0.5), class = 'b', coef = "abx1_FQ"),
            # for the intensity
            prior(normal( 2, .1), class = 'b', coef = "intensityablative"),
            prior(normal( 2, .1), class = 'b', coef = "intensityreduced"),
            prior(normal( 2, .1), class = 'b', coef = "intensitynonablative")
            )

model_div_original <- brm( mod_original,  
              data = meta, 
              warmup = 1000, iter = 3000, 
              prior = div_priors_original,
              cores = ncores, 
              chains = 2, 
               control = list(adapt_delta = 0.99),
              seed = 123, sample_prior = T)

# save it for future use
post_res <- suppressWarnings(posterior_samples(model_div_original)) 
post_res %>%  write_csv('../data/18_model_div_original_no_abx_and_FQ.csv')


```

```{r}
post_res <- read_csv('../data/18_model_div_original_no_abx_and_FQ.csv') %>% 
   select(starts_with('b')) %>% 
  select(!contains('intensity')) %>% 
  gather('shortname','coeff') %>% 
  mutate(shortname = str_replace(shortname, 'b_abx._',''))

post_res %>% 
  group_by(shortname) %>% # Group by shortname to calculate medians per group
  mutate(median_coeff = median(coeff)) %>% # Calculate median for each shortname
  ungroup() %>% # Important to ungroup after calculating the median
  mutate(shortname = fct_reorder(shortname, median_coeff)) %>% 
  group_by(shortname) %>% # Group again to get intervals
  mutate(
    lower = quantile(coeff, 0.025), # Calculate lower bound of the interval
    upper = quantile(coeff, 0.975), # Calculate upper bound of the interval
    significant = ifelse(lower >= 0 | upper <= 0, "Significant", "Not Significant")
  ) %>%
  ungroup() %>%
  ggplot() +
    stat_pointinterval(
      aes(x = coeff, y = shortname, color = significant), # Color by significance
      .width = .95,
      fatten_point = 1.2
    ) +
    geom_vline(xintercept = 0, col = 'blue', size = 0.5) +
    scale_color_manual(values = c("Significant" = "red", "Not Significant" = "black")) + # Set colors
    labs(x = "Coefficient", y = "Shortname", color = "95% CI") + # Improved labels
    theme_bw()
ggsave('../data/18_model_div_original_no_abx_and_FQ.pdf', width = 7)  
```
```{r}
# for the above model I also wanna show that what time were they collected     
meta %>%  
  mutate(log_div = log(simpson_reciprocal)) %>% 
  mutate(antibiotics = if_else(exposure_category == 'no_abx', 'not exposed', 'exposed'),
         antibiotics = factor(antibiotics, levels = c('not exposed','exposed'))) %>% 
  ggscatter(x = 'sdrt', y = 'log_div', color = 'antibiotics',
            ylab = 'log(diversity)',xlab = 'sdrt',
            alpha = 0.35, shape = 16, size = 1,
            add = "reg.line",  # Add regressin line
           add.params = list(color = 'antibiotics', fill = 'antibiotics', alpha = 0.3, size = 1.5), # Customize line
           conf.int = TRUE, # Add confidence interval
           cor.coef = TRUE, # Add correlation coefficient.
           cor.coeff.args = list(method = "spearman",  label.sep = "\n", cor.coef.name = c("rho"),p.accuracy = 0.01, r.accuracy = 0.01,label.x.npc = "left", label.y.npc = "bottom", size = 3.5)) +
  #geom_text(data = num_annot, aes(label = lbl), 
   #           x = Inf, y = Inf, hjust = 0.6, vjust = 1.5) +
  scale_fill_manual('antibiotics', values = palette, labels=c('not exposed', 'exposed')) +
  scale_colour_manual('antibiotics',values = palette, labels=c('not exposed', 'exposed')) +
  facet_wrap(~ antibiotics, labeller = 'label_both', dir = 'h') +
  theme(aspect.ratio = 1/1.15,   
        #legend.position = 'none', 
          strip.background = element_blank(),
        strip.text.x = element_blank(), 
     axis.text=element_text(size=9), axis.title=element_text(size=9)) 
```
```{r}
# what if only limiting the above model to pre transplant             
```

# heatmap how the switch happened 

Heatmap: Rows could represent patients, columns represent days, and cells could be colored to indicate the antibiotic status (e.g., FQ = blue, Empiric = red, Both = purple, Neither = white).

```{r}
#Data Preparation: You need your data in a format where, for each patient, you have:
# 
# Patient_ID: A unique identifier for each patient.
# Day: The day relative to transplant (e.g., -2, -1, 0, 1, 2...).
# FQ_Exposure: A binary variable (0 or 1) indicating whether the patient received fluoroquinolones on that day. Crucially, this should be coded as 1 only if they received FQ and not empiric antibiotics.
# Empiric_Antibiotic_Exposure: A binary variable (0 or 1) indicating whether the patient received empiric broad-spectrum antibiotics on that day. Again, be precise in your definition of "empiric."
# 


# I think I need to do this from the patient all med data itself and not the stool sample overlapped 

```

```{r}
# classify all of the meds the pt had into the categories
route_clean <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/routes_simple_2025-01-29.csv')

ptb <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/156_combined_PTB.csv')

types <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/191_types_of_medication.csv') 
```


```{r}
meds <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/169_all_medication.csv')  %>% 
  select(mrn, drug_name_clean,class,  route,together, start_date, stop_date  ) %>% 
  mutate(drug_name_clean = if_else(drug_name_clean == 'levofloxacin (levaquin)', 'levofloxacin', drug_name_clean)) %>% 
  left_join(route_clean %>% select(-n)) %>% 
  left_join(ptb %>% select(mrn, hct)) %>% 
  mutate(startday = as.numeric(start_date - hct), stopday = as.numeric(stop_date - hct)) %>% 
  filter(stopday > -5 & startday < 40) %>% 
  filter(startday > -20 & stopday < 40) %>% 
  left_join(types %>% select(-n, -class)) %>% 
  select(-c(start_date, stop_date, hct, route)) %>% 
  mutate(together = str_glue('{drug_name_clean}__{route_clean}'))

# annotate the medication that are NA
what <- meds %>% 
  filter(is.na(drug_category_for_this_study)) %>% distinct(drug_name_clean, route_clean, class)

# ertapenem  
# the most important few
few <- what %>% 
  mutate(together = str_glue('{drug_name_clean}__{route_clean}')) %>%  
  mutate(drug_category_for_this_study = if_else(together %in% c('amikacin__IV','gentamicin__IV','ampicillin__IV','penicillin v potassium__oral','cefdinir__oral','cephalexin__oral', 'azithromycin__IV', 'azithromycin__oral'), 'other_antibacterials', if_else(together == 'ertapenem__IV', 'broad_spectrum','not_antibacterial')))

link <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/152_pid_match.csv')
meds_updated <- bind_rows(
  meds %>% filter(is.na(drug_category_for_this_study)) %>% select(-drug_category_for_this_study) %>% left_join(few  ),
  meds %>% filter(!is.na(drug_category_for_this_study))
) %>% select(-class, -route_clean) %>%
   full_join(link) %>% select(-mrn) %>% relocate(pid, .before = 'drug_name_clean')

```

```{r}
# make the df with day exposure type 
meds_expanded <- meds_updated %>%
    distinct() %>% 
    group_by(pid, drug_name_clean, together, startday, stopday, drug_category_for_this_study) %>%
    summarize(day = seq(startday, stopday), .groups = "drop") %>% 
    group_by(pid, day) %>%
  summarise(
    exposure_category = case_when(
      any(drug_category_for_this_study == "broad_spectrum") & any(drug_category_for_this_study == "fluoroquinolones") ~ "both",  # Both present
      any(drug_category_for_this_study == "broad_spectrum") ~ "broad_spectrum",  # Only broad_spectrum
      any(drug_category_for_this_study == "fluoroquinolones") ~ "fluoroquinolones",
      any(drug_category_for_this_study == "other_antibacterials") ~ "other_antibacterials",
      TRUE ~ "not_antibacterial"
    ), .groups = "drop"
  )

meta <- read_csv('../data/153_combined_META.csv') %>% 
  select(pid, sdrt)

# what are in the other_antibacterials
other_antibacterials <- meds_updated %>%
  filter(drug_category_for_this_study == 'other_antibacterials') %>% 
  distinct(drug_name_clean, together) %>% 
  summarise(text = str_c(drug_name_clean, collapse = ', '))
other_antibacterials$text
```


```{r}
# heatmap
df<- meds_expanded


# 2. Find the first broad-spectrum day for each patient
first_broad_spectrum <- df %>%
  filter(exposure_category == "broad_spectrum") %>%
  group_by(pid) %>%
  summarize(FirstBroadSpectrumDay = min(day), .groups = 'drop') %>% 
  full_join(df %>% distinct(pid)) %>% 
  mutate(FirstBroadSpectrumDay = if_else(is.na(FirstBroadSpectrumDay), 40, FirstBroadSpectrumDay))

# 3. Sort PatientIDs based on the first broad-spectrum day
sorted_patient_ids <- first_broad_spectrum %>%
  arrange(FirstBroadSpectrumDay) %>%
  pull(pid)
```

```{r}
# 1. Ensure 'pid' is a factor (for proper ordering on the y-axis)
df$pid <- factor(df$pid)

# 2. Create a complete grid of pid and day combinations
#    This is *crucial* to handle missing days and ensure consistent plotting
df_complete <- df %>%
  complete(pid, day = full_seq(day, 1), fill = list(exposure_category = "not_antibacterial")) %>% 
  mutate(has_stool = if_else(paste(pid, day) %in% paste(meta$pid, meta$sdrt), T, F)) %>% 
  mutate(pid = factor(pid, levels = sorted_patient_ids)) 
#df_complete <- df
 
# 3. Define the color palette
color_palette <- c(
  "broad_spectrum" = "#FF000070",
  "fluoroquinolones" = "#0000FF70",
  "both" = "#A020F090",
  "not_antibacterial" = "#FFFFFF70",
  "other_antibacterials" = "#80808070"
)
 
# --- Create the Heatmap using ggplot2 ---
heatmap <- ggplot(df_complete, aes(x = day, y = pid, fill = exposure_category)) +
  geom_tile(aes(fill = exposure_category, color = has_stool),  linewidth = 0.7, show.legend = FALSE     )+  
  scale_fill_manual(values = color_palette) +  # Apply the defined color palette
  scale_color_manual(values = c(NA,'brown')) + 
  scale_x_continuous(breaks = seq(min(df_complete$day), max(df_complete$day), by = 5),
                     minor_breaks = seq(min(df_complete$day),max(df_complete$day), by = 1)) + # Show every 5th day, with minor breaks
  labs(title = "",
       x = "Transplant day",   
       y = "Patient ID",
       fill = "Exposure Category") +  # Add labels and title
  theme_minimal() +  # Use a cleaner theme
  geom_vline(xintercept = 0,  # The x-value where the line should be
                   linetype = "dashed",  # Optional: Line style (solid, dashed, dotted, etc.)
                   color = "black",     # Optional: Line color
                   linewidth = 1)  +
  theme(
    panel.grid.major = element_blank(),    # Remove major grid lines
    panel.grid.minor = element_blank(), 
    legend.position = "bottom"   ,      # Place the legend at the bottom
    axis.text.x = element_text(size = 6, angle = 0, hjust = 1, vjust=0.5),
  axis.text.y = element_blank(),
  axis.title = element_text(size = 8),
  legend.text = element_text(size = 6),
  legend.title = element_text(size = 8)
  )
      
#ggsave('../data/18_heatmpa_both.pdf', width = 9, height = 18)
title <- ggdraw() + 
  draw_label(
    "Fig. S4",
    fontface = 'bold',
    x = 0,
    hjust = 0
  ) +
  theme(
    plot.margin = margin(0, 0, 0, 0)
  )

combined <- plot_grid(
  title, heatmap,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.01, 1)
)+theme(plot.margin = unit(c(6.5,6.5,6.5,6.5), "cm"))

ggsave('../data/S04_abx_heatmap.pdf',
      width = 215.9, height = 279.4, units = "mm", device = 'pdf', plot = combined, dpi = 300)  
```

```{r}
# Phi Coefficient (φ): This is the most appropriate measure of association for two binary variables. It's mathematically equivalent to Pearson's correlation when applied to two binary variables, but it's conceptually clearer.  You can easily calculate this in most statistical software (R, Python, etc.). A negative phi coefficient indicates the negative association you expect.
# 
# Contingency Table and Chi-Squared Test: Create a 2x2 contingency table:
# 
#               | Empiric Antibiotic = 1 | Empiric Antibiotic = 0 |
# -------------------------------------------------------------------
# FQ = 1        |          a              |           b             |
# -------------------------------------------------------------------
# FQ = 0        |          c              |           d             |
# a: Number of patient-days with both FQ and empiric antibiotics. This should be relatively low, ideally zero, if your definitions are exclusive.
# b: Number of patient-days with FQ but no empiric antibiotics.
# c: Number of patient-days with empiric antibiotics but no FQ.
# d: Number of patient-days with neither FQ nor empiric antibiotics.
# You can then perform a chi-squared test of independence. A significant p-value indicates that the two variables are not independent (i.e., they are associated).  The direction of the association (positive or negative) is determined by examining the table itself.  You'd expect b and c to be larger than a and d if there's a negative association, as it means one treatment tends to preclude the other.

# to justify that they are correlated and we choose to include one that has more effects on the patients microbiome        
```

```{r}
num_days <- meds_expanded %>% 
  count(exposure_category)

# perform the chi squared test of independence  
```

```{r}
# --- Input: Your a, b, c, d values ---
# Replace these with your *actual* counts
a <- 183  # Example: Both FQ and empiric
b <- 1512  # Example: FQ only
c <- 1849  # Example: Empiric only
d <- 3007 # Example: Neither

# --- Create the Contingency Table ---
contingency_table <- matrix(c(a, b, c, d),
                           nrow = 2,
                           byrow = TRUE,
                           dimnames = list(c("FQ = 1", "FQ = 0"),
                                           c("Empiric = 1", "Empiric = 0")))

print(contingency_table)
```


```{r}
# --- Perform the Chi-Squared Test ---
chi_squared_result <- chisq.test(contingency_table)

# --- Print the Results ---
print(chi_squared_result)

# --- Extract and Interpret Key Values ---
cat("\n--- Interpretation ---\n")

# 1. Chi-squared statistic
cat("Chi-squared statistic:", chi_squared_result$statistic, "\n")

# 2. Degrees of freedom
cat("Degrees of freedom:", chi_squared_result$parameter, "\n")

# 3. P-value
cat("P-value:", chi_squared_result$p.value, "\n")

# 4. Interpretation of the p-value
if (chi_squared_result$p.value < 0.05) {
  cat("The p-value is less than 0.05.  There is a statistically significant association between FQ use and empiric antibiotic use.\n")
  # Additional analysis (Odds Ratio) would be needed here to determine the *direction* of the association.

} else {
  cat("The p-value is greater than or equal to 0.05. There is no statistically significant association between FQ use and empiric antibiotic use.\n")
}



# 6. Odds Ratio and Confidence Interval (for direction and strength of association)

# Calculate Odds Ratio (OR)
odds_ratio <- (a * d) / (b * c)
cat("\nOdds Ratio for direction and strength of association:", odds_ratio, "\n")

# Calculate 95% Confidence Interval for OR
# We use the log odds ratio for the CI calculation, then exponentiate
log_or <- log(odds_ratio)
se_log_or <- sqrt(1/a + 1/b + 1/c + 1/d)  # Standard error of log OR
lower_ci_log <- log_or - 1.96 * se_log_or
upper_ci_log <- log_or + 1.96 * se_log_or

lower_ci <- exp(lower_ci_log)
upper_ci <- exp(upper_ci_log)

cat("95% Confidence Interval for Odds Ratio: [", lower_ci, ", ", upper_ci, "]\n", sep="")

# Interpret the Odds Ratio and CI:
if (odds_ratio > 1) {
    cat("An odds ratio greater than 1 suggests a positive association.\n")
} else if (odds_ratio < 1) {
    cat("An odds ratio less than 1 suggests a negative association.\n")
} else {
    cat("An odds ratio of 1 suggests no association.\n")
}

if (lower_ci > 1 | upper_ci < 1) { # Check that interval does *not* include 1.
  cat("The 95% confidence interval does *not* include 1, supporting the statistical significance of the association.\n")
} else{
    cat("The 95% CI includes 1. The association is not significant at p<0.05.\n")
}
```


```{r}
library(lme4)

# Create the long-format data frame (as before)
df_long <- meds_expanded %>%
    mutate(EmpiricUse = ifelse(exposure_category %in% c("broad_spectrum", "both"), 1, 0),
           FQUse = ifelse(exposure_category %in% c("fluoroquinolones", "both"), 1, 0))

# Fit the mixed-effects model (FQ use as outcome)
model_fq <- glmer(FQUse ~ EmpiricUse + (1 | pid), data = df_long, family = binomial)
summary(model_fq)
```


```{r}
# Fit the mixed-effects model (Empiric use as outcome) - for completeness
model_empiric <- glmer(EmpiricUse ~ FQUse + (1 | pid), data = df_long, family = binomial)
summary(model_empiric)
```

```{r} 
# fishers <- meds_expanded %>% 
#   split(.$pid) %>% 
#   map(function(df){
#   keep(~ nrow(.x) > 0) %>% 
#     df_days <- df %>% count(exposure_category) %>% 
#                       full_join(tibble(exposure_category = c('broad_spectrum','not_antibacterial','fluoroquinolones','both')), by = join_by(exposure_category)) %>% 
#                       mutate(n = if_else(is.na(n), 0, n))
#     a = df_days %>% filter(exposure_category == 'both') %>% pull(n)
#     b = df_days %>% filter(exposure_category == 'fluoroquinolones') %>% pull(n)
#     c = df_days %>% filter(exposure_category == 'broad_spectrum') %>% pull(n)
#     d = df_days %>% filter(exposure_category == 'not_antibacterial') %>% pull(n)
#     
#     contingency_table = matrix(c(a, b, c, d),
#                            nrow = 2,
#                            byrow = TRUE,
#                            dimnames = list(c("FQ = 1", "FQ = 0"),
#                                            c("Empiric = 1", "Empiric = 0")))
#     
#     test_result = fisher.test(contingency_table)
#     return(test_result$p.value)
#   }) %>% bind_rows(.id = 'pid') %>% 
#   gather('pid','pvalue') %>% 
#   mutate(padj =  p.adjust(pvalue))



```
 
```{r}
questions <- link %>% 
  filter(pid %in% c('P45','P94','P147','P87','P127', 'P11'))

 questions %>% write_csv('../data/18_check_medication.csv')
```

  