--- 
title: "Blast the asv seq from the Enterococcus genus thru the API myself"    
output: html_document
date: "2025-06-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rBLAST)
library(tidyverse)
library(Biostrings)
library(rentrez)
```


Step 1: Install NCBI BLAST+
Before running the R script, you must install the BLAST+ command-line tools from the NCBI. rBLAST will not work without them.

Go to the NCBI BLAST+ download page: https://ftp.ncbi.nlm.nih.gov/blast/executables/blast+/LATEST/

Download and run the installer for your operating system (e.g., ...-win64.exe for Windows, ...-x64-linux.tar.gz for Linux, or ...-universal-macosx.tar.gz for macOS).

Ensure it's in your system's PATH. The installers for Windows and macOS usually handle this automatically. For Linux, you may need to extract the archive and add the bin directory to your system's PATH. A quick way to check is to open a new terminal or command prompt and type blastn -version. If it prints a version number, you are all set.


```{r}
# --- Input Your Data Here ---
# Create a character vector of your ASV sequences.
# Using example Enterococcus faecalis 16S rRNA gene sequences below.
library(vdbR)
connect_database()
#get_table_from_database('asv_annotation_blast_detailed_ag')
get_table_from_database('asv_sequences_ag')
# list_table_from_database('detailed')
# 
# meta <- read_csv('../data/153_combined_META.csv')
# cts <- get_counts_subset(meta$sampleid)
# 
# cts_e <- cts  %>%
#   left_join(asv_annotation_blast_ag %>% select(asv_key, genus)) %>%
#   filter(genus == 'Enterococcus') %>%
#   select(asv_key, sampleid,  count, count_relative, genus)
# 
# cts_e %>% write_csv('../data/R25_all_enterococcus_asv_in_sample.csv')
```

```{r}
cts_e <- read_csv('../data/R25_all_enterococcus_asv_in_sample.csv')

unique_e_asv <- cts_e %>% 
  group_by(asv_key, genus) %>% summarise(total_abun = sum(count_relative)) %>% 
  arrange(desc(total_abun)) %>% 
  ungroup() %>% 
  inner_join(asv_sequences_ag %>% select(asv_key, asv_sequence))

asv_sequences <- unique_e_asv %>% select(asv_key, asv_sequence) %>% deframe()


# Convert the character vector to a DNAStringSet, which rBLAST uses.
# We also give them names so we can track them in the output.
query_sequences <- DNAStringSet(asv_sequences)
```

```{r}
#Download the prebuilt 16S Microbial data base from NCBI’s ftp server at: https://ftp.ncbi.nlm.nih.gov/blast/db/
tgz_file <- blast_db_get("16S_ribosomal_RNA.tar.gz")
untar(tgz_file, exdir = "../software/blast/16S_rRNA_DB")
```


```{r}
# Load the downloaded BLAST database.
bl <- blast(db = "../software/blast/16S_rRNA_DB/16S_ribosomal_RNA")
bl

# Run the query. This is now much faster as it doesn't use the internet.
blast_results <- predict(bl, query_sequences)

# extrememly confident
#Perfect Identity: You have a pident of 100%. Your sequence is a perfect match.

# Extremely Low E-value: The e-value of 1.52e-175 essentially rules out the possibility of a random match.
# 
# High Bitscore: The bitscore of 614 further confirms the quality of the alignment.
```

```{r}
# Apply the filtering logic   
top_hits_filtered <- blast_results %>%
  # Work on each ASV independently
  group_by(qseqid) %>%
  # Apply a conditional filter
  filter(
    # The 'if/else' gets evaluated once for the entire group
    if (any(pident == 100)) {
      # If any 100% match exists, this is the rule: keep only the 100s
      pident == 100
    } else {
      # Otherwise, this is the rule: keep the row(s) with the highest bitscore
      bitscore == max(bitscore)
    }
  ) %>%
  # It's good practice to ungroup after you're done with group operations
  ungroup()
 
# View the result
print(top_hits_filtered)

top_hits_filtered %>% write_csv('../data/R25_top_hits_filtered_enterococcus_asv.csv')
```

 
# find the taxa name from ncbi

```{r}
top_hits_filtered <- read_csv('../data/R25_top_hits_filtered_enterococcus_asv.csv')

accession_ids <- top_hits_filtered %>% 
  distinct(sseqid) %>% pull(sseqid)

# Step 3: Fetch summary records for all IDs at once
# This might take a few seconds
summary_records <- entrez_summary(db = "nuccore", id = accession_ids)

# Step 4: Extract just the title for each record, which contains the name
# The sapply() function is a clean way to pull the same piece of info from every item in a list
organism_names <- sapply(summary_records, "[[", "title") %>% enframe('name','taxa_desc') %>% 
  mutate(accession_id = accession_ids)

# Print the resulting vector of names
print(organism_names)
```

# organize them into a table

```{r}
top_entero_asv_taxa <- top_hits_filtered %>% 
  rename(accession_id = sseqid) %>% 
  left_join(organism_names) %>% 
  rename(asv_key = qseqid) %>% 
  left_join(unique_e_asv %>% select(asv_key, total_abun, asv_sequence))

top_entero_asv_taxa %>% write_csv('../data/R25_Enterococcus_asv_blast_res_full_sorted.csv')
```

