---
title: "sub groups of the sweets"
output: html_document
date: "2025-06-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(broom.mixed)
library(brms)   
library(ggpubr)
library(tidybayes)
library(cowplot)
library(ggridges)
library(brmstools)
library(bayesplot)
options(mc.cores = parallel::detectCores())
rstan::rstan_options(auto_write = TRUE)
theme_set(theme_tidybayes() + panel_border())
ncores <- parallel::detectCores()
```

# Major Critiques and Challenges

Lack of Interpretability (The "Black Box" Problem): This is the biggest critique. The model's output can be incredibly difficult to understand. What does the coefficient for "apples" mean after being adjusted for 499 other foods, many of which are highly correlated? The complexity often obscures, rather than clarifies, the biological story.

Extreme Collinearity: Dietary data is notoriously collinear (e.g., people who eat hot dogs often eat hot dog buns). Putting hundreds of correlated food items into one model can make the individual coefficient estimates highly unstable and unreliable. The model may struggle to assign the effect correctly between two highly correlated foods.

Risk of Overfitting: Without very careful and robust regularization, a model with so many predictors is at high risk of "overfitting"—finding spurious patterns in the random noise of your specific dataset that will not be reproducible in the future.

It May Be Overkill: If the dominant biological signal comes from a broad nutritional pattern (like high sugar intake), modeling every single food item individually might not provide any more insight than the more logical, hierarchical approach you are taking.

# the sub groups of sweets

```{r}
# what's the subgroup of sweets?
sweets3 <- read_tsv('../data/NodeLabels.txt', col_types = 'cc')  %>% 
  filter(str_detect(Level.code, '^9')) %>% 
  filter(str_length(Level.code) == 3) 
```


```{r}
dtb  <- read_csv('../data/152_combined_DTB.csv')

# testing the solid and liquid sugar foods association with abx !     
dtb_ <- dtb %>% 
  select(pid, fdrt, Food_NSC, Food_code, dehydrated_weight, Sugars_g) 
```
```{r}
daily_intake_summary <- dtb_ %>%

  # Create a new 'food_category' column based on FNDDS codes.
  # The case_when() function checks conditions in order.
  # This is why we put the more specific 3-digit rules before the general 1-digit rules.
  mutate(
    food_category = case_when(
      
      # Rule for high-sugar sports drinks, grouping them with other liquid sweets
      Food_code == "95320200" ~ "fg_Liquid_Sweets",
      # Rule for low-sugar/electrolyte drinks, creating a new category
      Food_code %in% c("95322200", "95330100") ~ "fg_Functional_Beverages",
      
      # --- Define Oral Nutritional Supplements (ONS) ---
      str_starts(Food_code, "951") | str_starts(Food_code, "952") ~ "fg_ONS",
      
      # Solid Sweets & Sugars
      str_starts(Food_code, "911") | str_starts(Food_code, "913") |
      str_starts(Food_code, "914") | str_starts(Food_code, "915") |
      str_starts(Food_code, "917") | str_starts(Food_code, "918") ~ "fg_Solid_Sweets",

      # Liquid Sugar (Sugar-Sweetened Beverages)
      str_starts(Food_code, "924") | str_starts(Food_code, "925") |
      str_starts(Food_code, "926") ~ "fg_Liquid_Sweets",
      
       # Define Coffee & Tea
      str_starts(Food_code, "921") | str_starts(Food_code, "922") |
      str_starts(Food_code, "923") ~ "fg_Coffee_Tea",

      # --- Define the 8 main food categories using the first digit ---
      str_starts(Food_code, "1") ~ "fg_milk",
      str_starts(Food_code, "2") ~ "fg_meat",
      str_starts(Food_code, "3") ~ "fg_egg",
      str_starts(Food_code, "4") ~ "fg_legume",
      str_starts(Food_code, "5") ~ "fg_grain",
      str_starts(Food_code, "6") ~ "fg_fruit",
      str_starts(Food_code, "7") ~ "fg_veggie",
      str_starts(Food_code, "8") ~ "fg_oils",

      # Any food code that doesn't match the rules above will be assigned NA
      TRUE ~ NA_character_
    )
  ) %>%

  # Remove any rows that were not assigned to a category (e.g., ONS, water, etc.)
  filter(!is.na(food_category)) %>%

  # Group the data by patient, day, and our new food category
  group_by(pid, fdrt, food_category) %>%

  # Calculate the total dehydrated weight for each group
  # The na.rm = TRUE argument handles any missing weight values gracefully
  summarise(
    total_dehydrated_weight = sum(dehydrated_weight, na.rm = TRUE),
    total_sugars = sum(Sugars_g, na.rm = TRUE),
    .groups = 'drop' # Ungroup the data after summarizing
  )

# --- View the final summary table ---
# This table is now in a "tidy" format, ready for analysis or plotting.
# Each row shows the total grams of a food category consumed by a patient on a specific day.

# You can also use pivot_wider() to create a "wide" table where each food category
# has its own column. This format is often required for statistical models.
daily_intake_wide <- daily_intake_summary %>%
  pivot_wider(
    names_from = food_category,
    values_from = total_dehydrated_weight,
    values_fill = 0 # Fill in 0 for categories a patient didn't eat on a given day
  )


# finding the prior two day average of the above groups
stool_samples_df <- read_csv('../data/153_combined_META.csv') %>% 
  select(pid, sdrt, sampleid, simpson_reciprocal, empirical, intensity, EN, TPN)
```


```{r}
# ---  Calculate the 2-Day Prior Average Intake ---

# This single pipe performs the entire calculation.
# It joins the two tables, filters for the correct time window,
# calculates the average, and reshapes the data into the final format.

stool_samples_with_window <- stool_samples_df %>%
  mutate(
    # The 2-day window starts 2 days before the stool sample...
    window_start = sdrt - 2,
    # ...and ends 1 day before the stool sample.
    window_end = sdrt - 1
  )

# Now, perform the join using these new, pre-calculated columns
final_model_data <- stool_samples_with_window %>%
  left_join(
    daily_intake_summary,
    # The join is performed using direct column comparisons
    by = join_by(
      pid,
      window_start <= fdrt,
      window_end >= fdrt
    )
  ) %>%
  # This step reshapes the data into the final wide format needed for modeling.
  # It calculates the average over the 2-day window.
  pivot_wider(
    id_cols = c(pid, sdrt), # The columns that identify a unique stool sample
    names_from = food_category,
    values_from = total_dehydrated_weight,
    values_fn = ~ sum(.x, na.rm = TRUE) / 2, # The function to apply: sum and average
    values_fill = 0, # If a category wasn't eaten, its value is 0
    names_prefix = "avg_intake_" # Adds a clear prefix to the new columns
  ) %>%
    # This divides all the new intake columns by 100.
  # The across() function efficiently applies the same operation to many columns.
  mutate(across(starts_with("avg_intake_"), ~ .x / 100)) %>%
  
  # Joins back the original stool sample info
  right_join(stool_samples_df, by = c("pid", "sdrt")) %>% 
  mutate(timebin = cut_width(sdrt, 7, boundary=0, closed = 'left')) %>% 
  mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>% 
  mutate(pid = factor(pid)) 

# repeat the above process but only calculate the prior two day of sugar in those food categories
final_model_sugar <- stool_samples_with_window %>%
  left_join(
    daily_intake_summary,
    # The join is performed using direct column comparisons
    by = join_by(
      pid,
      window_start <= fdrt,
      window_end >= fdrt
    )
  ) %>%
  # This step reshapes the data into the final wide format needed for modeling.
  # It calculates the average over the 2-day window.
  pivot_wider(
    id_cols = c(pid, sdrt), # The columns that identify a unique stool sample
    names_from = food_category,
    values_from = total_sugars,
    values_fn = ~ sum(.x, na.rm = TRUE) / 2, # The function to apply: sum and average
    values_fill = 0, # If a category wasn't eaten, its value is 0
    names_prefix = "avg_sugar_intake_" # Adds a clear prefix to the new columns
  ) %>%
    # This divides all the new intake columns by 100.
  # The across() function efficiently applies the same operation to many columns.
  mutate(across(starts_with("avg_sugar_intake_"), ~ .x / 100)) %>%
  
  # Joins back the original stool sample info
  right_join(stool_samples_df, by = c("pid", "sdrt")) %>% 
  mutate(timebin = cut_width(sdrt, 7, boundary=0, closed = 'left')) %>% 
  mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>% 
  mutate(pid = factor(pid)) 


```

# iterative modeling 

```{r}

# --- Step 2: Define the food categories for the iterative analysis ---

# Define the "sweets" subcategories that will be swapped in and out of the model
sweets_subcategories_to_test <- c(
  "avg_intake_fg_Solid_Sweets",
  "avg_intake_fg_Liquid_Sweets",
  "avg_intake_fg_ONS",
  "avg_intake_fg_Coffee_Tea",
  "avg_intake_fg_Functional_Beverages"
)

# Define the 8 main food groups that will be included as covariates in every model
main_food_groups <- c(
  "avg_intake_fg_milk", "avg_intake_fg_meat", "avg_intake_fg_egg", "avg_intake_fg_legume", "avg_intake_fg_grain",
  "avg_intake_fg_fruit", "avg_intake_fg_veggie", "avg_intake_fg_oils"
)


# --- Step 3: Create a function that runs one iteration of the Bayesian model ---
# This function takes one of the "sweets" subcategory names as input,
# builds the full model formula, sets all priors, runs the model,
# and returns the key results.
run_iterative_model <- function(sweet_variable, data) {

  # Combine the main food groups and the current sweet variable
  all_food_vars <- c(main_food_groups, sweet_variable)

  # Create the interaction terms for all food variables in this iteration
  interaction_terms <- paste(all_food_vars, "empirical", sep = ":")

  # Build the full formula string dynamically
  formula_string <- paste(
    "log(simpson_reciprocal) ~ 0 + intensity + empirical + TPN + EN +",
    paste(interaction_terms, collapse = " + "),
    "+ (1 | pid) + (1 | timebin)"
  )

  # Convert the string to a formula object
  formula <- brms::bf(as.formula(formula_string))

  # Build the priors by adding them together.
  # This single prior() call applies to all coefficients listed in `all_food_coefs`.
  priors <-
    prior(normal(0, 1), class = 'b') + # General prior for all food effects
    # Specific priors that override the general one for non-food covariates
    prior(normal(0, 0.1), class = 'b', coef = "TPNTRUE") +
    prior(normal(0, 0.1), class = 'b', coef = "ENTRUE") +
    prior(normal(0, 0.5), class = 'b', coef = "empiricalTRUE") +
    prior(normal(2, .1), class = 'b', coef = "intensityablative") +
    prior(normal(2, .1), class = 'b', coef = "intensityreduced") +
    prior(normal(2, .1), class = 'b', coef = "intensitynonablative")

  # Fit the model (using fewer iterations for this example to run quickly)
  model_fit <- brm(
    formula = formula,
    data = data,
    prior = priors,
    warmup = 1000, iter = 3000,
    chains = 2, cores = 10, # Adjust cores as needed
    seed = 123,
    silent = 2
  )

  # Tidy the output and return only the interaction term for the sweet subcategory
  tidy(model_fit, conf.int = TRUE) %>%
    filter(term == paste0("empiricalTRUE:", sweet_variable))
}


# --- Step 4: Run the iterative analysis using purrr::map ---
# This will apply the function to each "sweets" subcategory.
# This step will take a significant amount of time to run.
all_results <- map(
  sweets_subcategories_to_test,
  ~run_iterative_model(sweet_variable = .x, data = final_model_data)
)
```

```{r}
# Combine the list of results into a single, clean data frame
comparison_df <- bind_rows(all_results) %>%
  # Clean up the term name to be used as a label
  mutate(food_category = str_remove(term, "empiricalTRUE:"))

# Create a forest plot to visualize and compare the effects
ggplot(comparison_df, aes(x = estimate, y = fct_reorder(food_category, estimate))) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 4, color = "dodgerblue") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, linewidth = 1, color = "dodgerblue") +
  labs(
    title = "Interaction Effect of Sweet Subcategories with Antibiotics on Diversity",
    subtitle = "Each estimate is from a separate model controlling for other main food groups",
    x = "Interaction Coefficient (Effect per 100g intake during antibiotic exposure)",
    y = "Sweet Subcategory"
  ) +
  theme_bw(base_size = 14)
```

# coffee and tea sugar proportion 

```{r}
merged_sugar <- final_model_data %>% 
  full_join(final_model_sugar %>% select(pid : avg_sugar_intake_fg_Functional_Beverages))

```

```{r}
plot_data <- merged_sugar %>%
  # Select only the columns we need for this plot
  select(pid,sdrt,  starts_with("avg_intake_"), starts_with("avg_sugar_intake_")) %>%
  # Pivot the data from a wide format to a long format
  pivot_longer(
    cols = avg_intake_fg_Coffee_Tea:avg_sugar_intake_fg_Functional_Beverages,
    names_to = "metric",
    values_to = "grams"
  ) %>%
  # Extract the food category name from the column name
  mutate(
    food_category = str_remove(metric, "avg_intake_|avg_sugar_intake_"),
    metric_type = if_else(str_starts(metric, "avg_sugar_intake_"), "sugar", "total")
  ) %>%
  # Reshape again to have total intake and sugar intake in separate columns
  select(pid,sdrt, food_category, metric_type, grams) %>%
  # revert it back to original value by * 100
  mutate(grams = grams * 100) %>% 
  pivot_wider(
    names_from = metric_type,
    values_from = grams
  ) %>%
  # Calculate the overall average intake across all samples for each category
  group_by(food_category) %>%
  summarise(
    avg_total_intake = mean(total, na.rm = TRUE),
    avg_sugar_intake = mean(sugar, na.rm = TRUE)
  ) %>%
  # Calculate the proportion of sugar for sorting purposes
  mutate(
    sugar_proportion = avg_sugar_intake / avg_total_intake
  ) %>%
  # Calculate the non-sugar portion for the stacked bar
  mutate(
    avg_nonsugar_intake = avg_total_intake - avg_sugar_intake
  ) %>%
  # Reshape one last time for easy plotting with ggplot
  pivot_longer(
    cols = c(avg_sugar_intake, avg_nonsugar_intake),
    names_to = "intake_type",
    values_to = "grams"
  ) %>%
  # Clean up names for the plot legend
  mutate(
    intake_type = if_else(intake_type == "avg_sugar_intake", "Sugar", "Non-Sugar Mass"),
    # Clean up food category names for the axis labels
    food_category = str_remove(food_category, "fg_") %>% str_replace_all("_", " "),
    # Reorder the 'food_category' factor based on the sugar proportion
    # The '.desc = TRUE' ensures it's in descending order (highest % at the top)
    food_category = fct_reorder(food_category, sugar_proportion, .desc = F)
  )


# --- Step 3: Create the Sorted Stacked Bar Chart ---

ggplot(plot_data, aes(x = grams, y = food_category, fill = intake_type)) +
  # Use geom_col which is an alias for geom_bar(stat="identity")
  geom_col() +
  # Use a color palette that clearly distinguishes sugar
  scale_fill_manual(values = c("Sugar" = "#FF6B6B", "Non-Sugar Mass" = "#BDBDBD")) +
  # Add informative labels
  labs(
    title = "Average Sugar Content by Food Category",
    subtitle = "Categories are sorted by the percentage of sugar content",
    x = "Average Dehydrated Weight Consumed (grams)",
    y = "Food Category",
    fill = "Component"
  ) +
  # Use a clean theme
  theme_bw() +
  theme(legend.position = "bottom")

```

```{r}
# find the top foods in the coffee/tea sub group
#  Filter, Summarize, and Prepare Data for Plotting ---

# This data wrangling pipe will isolate the coffee/tea items, find the top
# contributors, and calculate their sugar content for visualization.
coffee_tea_plot_data <- dtb_ %>%
  # Ensure Food_code is a character for string matching
  mutate(Food_code = as.character(Food_code)) %>%

  # Filter for only the food codes corresponding to Coffee & Tea
  filter(str_starts(Food_code, "921") | str_starts(Food_code, "922") | str_starts(Food_code, "923")) %>%

  # Group by the specific food name to aggregate all consumption records
  group_by(Food_NSC) %>%

  # Calculate the total intake and total sugar for each specific food item
  summarise(
    total_intake = sum(dehydrated_weight, na.rm = TRUE),
    total_sugar = sum(Sugars_g, na.rm = TRUE)
  ) %>%
  
  # Keep only the top 10 most consumed items for a clean plot
  slice_max(order_by = total_intake, n = 10) %>%

  # Calculate the non-sugar portion for the stacked bar
  mutate(
    non_sugar_mass = total_intake - total_sugar
  ) %>%

  # Reshape the data into a "long" format for easy plotting with ggplot
  pivot_longer(
    cols = c(total_sugar, non_sugar_mass),
    names_to = "component",
    values_to = "grams"
  ) %>%
  # Clean up names for the plot legend
  mutate(
    component = if_else(component == "total_sugar", "Sugar", "Non-Sugar Mass")
  )
```


```{r}
# --- Step 3: Create the Stacked Bar Chart ---

ggplot(coffee_tea_plot_data, aes(x = grams, y = fct_reorder(Food_NSC, grams, .fun = sum), fill = component)) +
  # Use geom_col to create the bars
  geom_col() +
  # Use a color palette that clearly distinguishes sugar
  scale_fill_manual(values = c("Sugar" = "#FF6B6B", "Non-Sugar Mass" = "#BDBDBD")) +
  # Add informative labels
  labs(
    title = "Top 10 Consumed Items in the Coffee & Tea Category",
    subtitle = "Showing total intake and the proportion from sugar",
    x = "Total Dehydrated Weight Consumed (grams)",
    y = "Food Item",
    fill = "Component"
  ) +
  # Use a clean theme
  theme_bw() +
  theme(legend.position = "bottom")
```

