---
title: "New clinical data - BSI CD4"
output: html_document
date: "2025-07-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(broom)
```

Date ranges:

Blood Cx: Days 0 to 100

ALC: Day -365 to 365

CD4: Day -365 to 365


Caveats –

Blood Cx: working on a more robust tool to pull out cleaned names for species isolates from the culture reports. Right now, done with simple Regex. Result2 column contains additional bugs if identified.

CD4: I am still trying to hunt down all the different assays we use to describe CD4+ cells, this one is just using the most popular assay I could identify for absolute CD4 quantification.

# Immune reconstitution

```{r}
# For the ALC and CD4 data, the uln of 3.2 means that a result above 3.2 would be considered higher than normal. However, in the context of BMT patients who are recovering from zero, the primary focus is on seeing the result_value increase towards the normal range. The ULN is just there for completeness.
load('../data/angel_chrt_bsi_cd4_alc.RData') 

ptb <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/156_combined_PTB.csv')
link <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/152_pid_match.csv')
meta <- read_csv('../data/153_combined_META.csv')

d_alc <- alc %>% 
  mutate(mrn = as.numeric(mrn), dt_performed = ymd(dt_performed)) %>% 
  left_join(ptb %>% select(mrn, hct)) %>% 
  mutate(alc_day = as.numeric(dt_performed - hct)) %>% 
  inner_join(link) %>% 
  select(-mrn) %>% 
  select(pid, alc_day, result_value, uln, lab)

# deidentify the other two
d_cd4 <- cd4 %>% 
  mutate(mrn = as.numeric(mrn), dt_performed = ymd(dt_performed)) %>% 
  left_join(ptb %>% select(mrn, hct)) %>% 
  mutate(cd4_day = as.numeric(dt_performed - hct)) %>% 
  inner_join(link) %>% 
  select(-mrn) %>% 
  select(pid, cd4_day, result_value, uln, lab) %>% 
  mutate(lab = 'cd4')
```



```{r}
# A standard and easily defensible choice would be to match each stool sample with the closest lab draw within a +/- 2 day window.

# For a stool sample collected on sdrt = 5, you would look for any ALC/CD4 measurements taken on days 3, 4, 5, 6, or 7.
# 
# If there are multiple lab draws in that window, you would typically use the one that is closest in time to the stool sample.

# --- Combine ALC and CD4 data into a single long-format lab table ---
# This makes the matching process much more efficient.
all_labs_df <- bind_rows(
  rename(d_alc, day = alc_day),
  rename(d_cd4, day = cd4_day)
)
```


```{r}
# ---  Create the Density Plot ---
# This plot shows the distribution of collection days for each lab type.

ggplot(all_labs_df, aes(x = day, fill = lab)) +
  # geom_density creates a smooth curve representing the distribution.
  # The 'alpha' argument adds transparency so you can see overlapping areas.
  geom_density(alpha = 0.6) +
  
  # Add a vertical line at Day 0 (transplant day) for reference
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_vline(xintercept = -9, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red") +
  
  # Manually define colors for clarity
  scale_fill_manual(values = c("alc" = "#118AB2", "cd4" = "#FF6B6B")) +
  
  # Add informative labels
  labs(
    title = "Distribution of ALC and CD4 Collection Days",
    subtitle = "Density plot shows when lab measurements were most frequently taken",
    x = "Days Relative to Transplant",
    y = "Density",
    fill = "Lab Type"
  ) +
  
  # Use a clean theme
  theme_bw(base_size = 14) +
  theme(legend.position = "bottom")

```

```{r}
# This plot shows the immune reconstitution trajectory for each patient,
# separated by lab type (ALC vs. CD4).
pseudocount <- 1

ggplot(all_labs_df, aes(x = day, y = result_value + pseudocount, color = lab, fill = lab)) +
  # Use geom_smooth to show the average trend line and its 95% confidence interval.
  # This replaces the individual patient lines for a much cleaner look.
  geom_smooth(method = "loess", alpha = 0.2) +
  
  # Use facet_wrap to create separate plots for 'alc' and 'cd4'.
  # 'scales = "free_y"' is crucial because ALC and CD4 are on different scales.
  facet_wrap(~ lab, scales = "free_y", nrow = 2) +
  
  # Use a log10 scale on the y-axis to better visualize the wide range of counts
  scale_y_log10() +
  annotation_logticks(sides = "l") + # Adds minor tick marks for the log scale
  
  # Manually define colors for clarity
  scale_color_manual(values = c("alc" = "#118AB2", "cd4" = "#FF6B6B")) +
  scale_fill_manual(values = c("alc" = "#118AB2", "cd4" = "#FF6B6B")) +
  
  # Add a vertical line at Day 0 (transplant day) for reference
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
    geom_vline(xintercept = -9, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red") +
  
  # Add informative labels
  labs(
    title = "Average Immune Reconstitution Trajectories for ALC and CD4",
    subtitle = "Line shows the average trend (LOESS smooth) across all patients",
    x = "Days Relative to Transplant",
    y = "Cell Count (log10 scale)",
    color = "Lab Type",
    fill = "Lab Type"
  ) +
  
  # Use a clean theme
  theme_bw(base_size = 14) +
  theme(legend.position = "bottom")

```

```{r}
# --- Match Stool Samples to Closest Lab Draw ---
# stool_df_with_window <- meta %>%
#   select(sampleid, sdrt, simpson_reciprocal, pid,intensity, empirical ) %>% 
#   mutate(
#     # The 2-day window starts 2 days before the stool sample...
#     window_start = sdrt - 2,
#     # ...and ends 2 days after the stool sample.
#     window_end = sdrt + 2
#   )
# 
# 
# # This single command performs the entire matching process for all lab types.
# matched_data <- stool_df_with_window %>%
#   # Perform a non-equi join. This finds all lab draws for each patient
#   # that fall within the +/- 2 day window of a stool sample.
#   left_join(
#     all_labs_df,
#     by = join_by(
#       pid,
#       window_start <= day,
#       window_end >= day
#     )
#   ) %>%
#   # If a stool sample had multiple lab draws in its window, this section
#   # will select only the one closest in time *for each lab type*.
#   mutate(
#     # Calculate the absolute time difference in days
#     time_diff = abs(sdrt - day)
#   ) %>%
#   # Group by the original stool sample information AND the lab type
#   group_by(sampleid, pid, sdrt, simpson_reciprocal, lab) %>%
#   # For each group, keep only the row with the minimum time difference
#   slice_min(order_by = time_diff, n = 1, with_ties = FALSE) %>%
#   # Ungroup to finish the operation
#   ungroup() %>%
#   # Now, pivot the data to a "wide" format, so ALC and CD4 have their own columns
#   select(-time_diff) %>% # Remove intermediate columns
#   pivot_wider(
#     names_from = lab,
#     values_from = result_value
#   )
# 
# 
# # Print the final merged table
# print(matched_data)


```

# CD4

peri engraftment diversity and the day 100/day 180 CD4

```{r}
#Day 100: Look for a measurement between Day +85 and Day +115.

#Day 180: Look for a measurement between Day +150 and Day +210." 
#and For any patient who has multiple CD4 measurements within your chosen window, you should select the one that is closest in time to the landmark day (i.e., closest to Day 100 or Day 180).
cd4_day100 <- all_labs_df %>%
  # Keep only the CD4 lab results
  filter(lab == "cd4") %>%
  # Filter for rows within the Day 100 window (+85 to +115)
  filter(day >= 85 & day <= 115) %>%
  # For each patient, calculate the time difference to the landmark day
  mutate(time_diff = abs(day - 100)) %>%
  # Group by patient ID to select the best measurement for each one
  group_by(pid) %>%
  # Keep only the row with the minimum time difference
  # with_ties = FALSE ensures we only get one row if there's a tie
  slice_min(order_by = time_diff, n = 1, with_ties = FALSE) %>%
  # Ungroup to finish the operation
  ungroup() %>%
  # Add a column to identify which landmark this data belongs to
  mutate(landmark = "Day_100") %>%
  # Rename columns for clarity
  rename(cd4_value = result_value, cd4_day = day) %>%
  select(pid, landmark, cd4_day, cd4_value, time_diff)
```


```{r}
# --- Step 3: Extract CD4 Count for Day 180 Landmark ---

cd4_day180 <- all_labs_df %>%
  # Keep only the CD4 lab results
  filter(lab == "cd4") %>%
  # Filter for rows within the Day 180 window (+150 to +210)
  filter(day >= 150 & day <= 210) %>%
  # For each patient, calculate the time difference to the landmark day
  mutate(time_diff = abs(day - 180)) %>%
  # Group by patient ID
  group_by(pid) %>%
  # Keep only the row with the minimum time difference
  slice_min(order_by = time_diff, n = 1, with_ties = FALSE) %>%
  # Ungroup to finish the operation
  ungroup() %>%
  # Add a column to identify which landmark this data belongs to
  mutate(landmark = "Day_180") %>%
  # Rename columns for clarity
  rename(cd4_value = result_value, cd4_day = day) %>%
  select(pid, landmark, cd4_day, cd4_value, time_diff)


# --- Step 4: Combine and View the Results ---

# Combine the results from both landmarks into a single, clean table
final_landmark_cd4 <- bind_rows(cd4_day100, cd4_day180) %>%
  arrange(pid, landmark)

# Print the final table
print(final_landmark_cd4)

```

```{r}
stool_df_with_window <- meta %>%
  select(sampleid, sdrt, simpson_reciprocal, pid,intensity, empirical ) %>%
  mutate(
    # The 2-day window starts 2 days before the stool sample...
    window_start = sdrt - 2,
    # ...and ends 2 days after the stool sample.
    window_end = sdrt + 2
  )

peri_engraftment_diversity <- stool_df_with_window %>%
  # Filter for rows within the peri-engraftment window
  filter(sdrt >= 7 & sdrt <= 21) %>%
  
  # Group by patient ID to calculate the average for each one
  group_by(pid) %>%
  
  # Calculate the mean of the diversity metric for all samples in the window
  # na.rm = TRUE will handle any missing diversity values gracefully
  summarise(
    peri_engraftment_diversity = median(simpson_reciprocal, na.rm = TRUE),
    # It's also useful to know how many samples were averaged
    n_samples_in_window = n()
  ) %>%
  
  # Ungroup to finish the operation
  ungroup() 

d100_df <- peri_engraftment_diversity %>% 
  inner_join(cd4_day100)

d180_df <- peri_engraftment_diversity %>% 
  inner_join(cd4_day180)
```
```{r}
correlation_model <- lm(log1p(cd4_value) ~ peri_engraftment_diversity, data = d100_df)

# Get a summary of the model to extract statistics
model_summary <- summary(correlation_model)

# Extract the R-squared and p-value for annotating the plot
r_squared <- round(model_summary$r.squared, 3)
p_value <- format(glance(model_summary)$p.value, digits = 2, scientific = TRUE)
```


```{r}
# --- Step 4: Create the Visualization ---
# A scatter plot is the best way to visualize the correlation.

ggplot(d100_df, aes(x = peri_engraftment_diversity, y = cd4_value)) +
  # Add the points for each patient
  geom_point(size = 3, color = "#118AB2", alpha = 0.8) +
  # Add the regression line to show the trend
  geom_smooth(method = "lm", se = TRUE, color = "#073B4C") +
  
  # Use a log scale for the y-axis for better visualization of CD4 counts
  # scale_y_log10() +
  # annotation_logticks(sides = "l") +
  
  # Add informative labels and a title that includes the correlation results
  labs(
    title = "Peri-Engraftment Diversity vs. Day 100 CD4 Count",
    subtitle = paste0("Estimate = ", 0.02758, " (p = ", p_value, ")"),
    x = "Median Peri-Engraftment Diversity (Simpson Reciprocal)",
    y = "CD4 Count at Day 100"
  ) +
  # Use a clean theme
  theme_bw(base_size = 14)

```
```{r}
correlation_model <- lm(log1p(cd4_value) ~ peri_engraftment_diversity, data = d180_df)

# Get a summary of the model to extract statistics
model_summary <- summary(correlation_model)

# Extract the R-squared and p-value for annotating the plot
# Estimate <- round(model_summary$, 3)
p_value <- format(glance(model_summary)$p.value, digits = 2, scientific = TRUE)

ggplot(d180_df, aes(x = peri_engraftment_diversity, y = cd4_value)) +
  # Add the points for each patient
  geom_point(size = 3, color = "#118AB2", alpha = 0.8) +
  # Add the regression line to show the trend
  geom_smooth(method = "lm", se = TRUE, color = "#073B4C") +
  
  # Use a log scale for the y-axis for better visualization of CD4 counts
  # scale_y_log10() +
  # annotation_logticks(sides = "l") +
  
  # Add informative labels and a title that includes the correlation results
  labs(
    title = "Peri-Engraftment Diversity vs. Day 180 CD4 Count",
    subtitle = paste0("Estimate = ", 0.02878, " (p = ", p_value, ")"),
    x = "Median Peri-Engraftment Diversity (Simpson Reciprocal)",
    y = "CD4 Count at Day 180"
  ) +
  # Use a clean theme
  theme_bw(base_size = 14)

```

