---
title: "Germ free"
output: html_document
date: "2025-07-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(ggrepel)
library(scales)
library(patchwork)
library(ggpubr)
library(cowplot)
```
# design

```{r}
# Manually create a tibble that contains every event for every group.
# This "long" data format is ideal for plotting with ggplot2.
# This version includes the stool plating at 144h for the E. faecalis groups.
experiment_design_df <- tribble(
  ~group_name, ~y_position, ~color_group, ~time_h, ~event_type,
  # Group 1: E. faecalis + Sucrose Water
  "E. faecalis + Sucrose Water", 4, "Sucrose Water",   0, "BHI Anaerobic Culture",
  "E. faecalis + Sucrose Water", 4, "Sucrose Water",   4, "Stool Plating",
  "E. faecalis + Sucrose Water", 4, "Sucrose Water",   8, "Stool Plating",
  "E. faecalis + Sucrose Water", 4, "Sucrose Water",  24, "Stool Plating",
  "E. faecalis + Sucrose Water", 4, "Sucrose Water",  48, "Stool Plating",
  "E. faecalis + Sucrose Water", 4, "Sucrose Water", 144, "Stool Plating", # Final stool plating

  # Group 2: E. faecalis + Regular Water
  "E. faecalis + Regular Water", 3, "Regular Water",   0, "BHI Anaerobic Culture",
  "E. faecalis + Regular Water", 3, "Regular Water",   4, "Stool Plating",
  "E. faecalis + Regular Water", 3, "Regular Water",   8, "Stool Plating",
  "E. faecalis + Regular Water", 3, "Regular Water",  24, "Stool Plating",
  "E. faecalis + Regular Water", 3, "Regular Water",  48, "Stool Plating",
  "E. faecalis + Regular Water", 3, "Regular Water", 144, "Stool Plating", # Final stool plating

  # Group 3: PBS + Sucrose Water
  "PBS + Sucrose Water", 2, "Sucrose Water",   0, "BHI Anaerobic Culture",
  "PBS + Sucrose Water", 2, "Sucrose Water",   4, "Stool Plating",
  "PBS + Sucrose Water", 2, "Sucrose Water",  48, "BHI Anaerobic Culture",
  "PBS + Sucrose Water", 2, "Sucrose Water", 144, "", # No final stool plating for PBS group

  # Group 4: PBS + Regular Water
  "PBS + Regular Water", 1, "Regular Water",   0, "BHI Anaerobic Culture",
  "PBS + Regular Water", 1, "Regular Water",   4, "Stool Plating",
  "PBS + Regular Water", 1, "Regular Water",  48, "BHI Anaerobic Culture",
  "PBS + Regular Water", 1, "Regular Water", 144, ""  # No final stool plating for PBS group
)

# Create a data frame for the background rectangles.
rect_data <- experiment_design_df %>%
  distinct(group_name, y_position, color_group)
```

## ggplot

```{r}
# Now, create the diagram using ggplot.
design <- ggplot(experiment_design_df, aes(x = time_h, y = y_position)) +
    # Layer 1: The background rectangles to show continuous treatment.
  geom_rect(
    data = rect_data,
    aes(xmin = 0, xmax = 144, ymin = y_position - 0.2, ymax = y_position + 0.2, fill = color_group),
    alpha = 0.2, # Make the rectangles semi-transparent
    inherit.aes = FALSE
  ) +
  
 # Layer 2: The horizontal timelines, colored by treatment.
  geom_line(aes(group = group_name, color = color_group), linewidth = 0.8) +
  
  # Layer 3: The action points (squares and triangles), now all black.
  geom_point(aes(shape = event_type), size = 2, color = "black") +
  
  # Manually define the fill and color for the rectangles and lines.
  scale_fill_manual(values = c("Sucrose Water" = "#F8766D", "Regular Water" = "#00BFC4")) +
  scale_color_manual(values = c("Sucrose Water" = "#F8766D", "Regular Water" = "#00BFC4")) +
  
  # Manually define the shapes for the action points.
  scale_shape_manual(values = c("BHI Anaerobic Culture" = 15, "Stool Plating" = 17, "Sacrifice" = 16)) +
  
  # Add the group labels to the left of the timelines.
  geom_text(
    data = experiment_design_df %>% distinct(group_name, y_position),
    aes(x = -2, y = y_position, label = group_name),
    hjust = 1, size = 3
  ) +
  
  # Use a clean theme for a diagrammatic look.
  theme_classic(base_size = 11) +
  theme(
    axis.line.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) +
  
  # Add informative titles and labels.
  labs(
    x = "Time Post-Gavage (Hours)",
    fill = "Water Treatment", # Legend title for the fill color
    shape = "Action"
  ) +
  
  # Customize the x-axis breaks to show all key time points.
  scale_x_continuous(
    breaks = c(0, 4, 8, 24, 48, 144),
    labels = c("0", "4", "8", "24", "48", "144 (Sacrifice)")
  ) +
  
  # Expand plot limits so labels on the left are not cut off.
  coord_cartesian(xlim = c(-45, 150)) +
  
  # This ensures the color legend for the lines is not displayed, as it's redundant.
  guides(color = "none")
design

#ggsave('../data/R33_gf_design_new.pdf', width = 10)
```

# the CFU

for this figure:


-plot with a thickish line for the group median and thin semi-transparent spaghetti tracings for inividual mice
-add p value; can be with Vardi test or GEE

```{r}
# Load the data from the CSV file.
# The `read_csv` function is great because it handles column types automatically.
raw_data <- read_csv("../data/7_29_25_germ_free_faecalis_gavage - sheet1.csv") %>% 
  # We can only plot values greater than 0 on a log scale.
  mutate(CFUs_per_GramStool = if_else(CFUs_per_GramStool == 0, 1, CFUs_per_GramStool)) %>% 
  mutate(log10_CFUs_per_GramStool = log10(CFUs_per_GramStool))

# add the pre-innoculation (hour 0) enterococcus abundance (0 from bhi anaerobic culturing)  
baseline_data <- tibble(
  # The 'Time Point' is the same for all baseline measurements.
  `Time Point` = rep("0h (pre-innoculation)", 10),
  
  # A vector containing the unique group IDs for each of the 10 mice.
  `Mouse Group` = c("1-1", "1-3", "1-10", 
                    "2-1", "2-3", "2-10", 
                    "3-1", "3-3", 
                    "4-1", "4-3"),
                    
  # The notes column specifies the source of the germ-free confirmation.
  Notes = rep("BHI anaerobic culturing", 10),
  
  # The CFU/gram is set to 1, representing the limit of detection (no growth).
  `CFUs_per_GramStool` = rep(1, 10),
  
  # The log10 of 1 is 0.
  `log10_CFUs_per_GramStool` = rep(0, 10)
)


no_gavage_culturing_48 <- tibble(
  # The 'Time Point' is the same for all baseline measurements.
  `Time Point` = rep("48h", 4),
  
  # A vector containing the unique group IDs for each of the 10 mice.
  `Mouse Group` = c(
                    "3-1", "3-3", 
                    "4-1", "4-3"),
                    
  # The notes column specifies the source of the germ-free confirmation.
  Notes = rep("BHI anaerobic culturing", 4),
  
  # The CFU/gram is set to 1, representing the limit of detection (no growth).
  `CFUs_per_GramStool` = rep(1, 4),
  
  # The log10 of 1 is 0.
  `log10_CFUs_per_GramStool` = rep(0, 4)
)

# add the repeat experiment data 
repeat_data <- read_csv("../data/8_13_25_germ_free_faecalis_gavage_repeat - sheet1.csv") %>% 
  mutate(CFUs_per_GramStool = (((((Colonies_counted*(100/Plated_volume_ul))/100)/(10^Dilution_factor) )*1000)/Stool_weight_mg)+1)  %>%
  mutate(log10_CFUs_per_GramStool = log10(CFUs_per_GramStool)) 

## add the pre-innoculation (hour 0) enterococcus abundance (0 from bhi anaerobic culturing)  for the repeat experiment
baseline_data_repeat <- tibble(
  # The 'Time Point' is the same for all baseline measurements.
  `Time Point` = rep("0h (pre-innoculation)", 5),
  
  # A vector containing the unique group IDs for each of the 10 mice.
  `Mouse Group` = c("1-1", "1-3", "1-10", 
                    "2-1", "2-3"),
                    
  # The notes column specifies the source of the germ-free confirmation.
  Notes = rep("BHI anaerobic culturing", 5),
  
  # The CFU/gram is set to 1, representing the limit of detection (no growth).
  `CFUs_per_GramStool` = rep(1, 5),
  
  # The log10 of 1 is 0.
  `log10_CFUs_per_GramStool` = rep(0, 5)
)


all_raw <- bind_rows(
  bind_rows(baseline_data, raw_data , no_gavage_culturing_48) %>% mutate(experiment_no = 1, Total_weight_stool_and_tube_mg = as.numeric(Total_weight_stool_and_tube_mg)),
  bind_rows(baseline_data_repeat, repeat_data )%>% mutate(experiment_no = 2)
) %>% 
  mutate(mouse_identifier = str_glue('E{experiment_no}__{`Mouse Group`}'))

# Clean and transform the data for plotting.
plot_data <- all_raw %>%
  # Use mutate to create new, cleaner columns.
  mutate(
    # Convert the 'Time Point' text into a numerical hour for the x-axis.
    Time_h = case_when(
      `Time Point` == "0h (pre-innoculation)" ~ 0,
      `Time Point` == "4h post gavage" ~ 4,
      `Time Point` == "48h" ~ 48,
      `Time Point` == "8h post gavage" ~ 8,
      `Time Point` == "24h post gavage" ~ 24,
      `Time Point` == "48h post gavage" ~ 48,
      `Time Point` == "144h post gavage" ~ 144
    ),
    # Create descriptive labels for the four treatment groups.
    Treatment = case_when(
      str_starts(`Mouse Group`, "1-") ~ "E. faecalis + Sucrose Water",
      str_starts(`Mouse Group`, "2-") ~ "E. faecalis + Regular Water",
      str_starts(`Mouse Group`, "3-") ~ "PBS + Sucrose Water",
      str_starts(`Mouse Group`, "4-") ~ "PBS + Regular Water"
    ),
    # Ensure the treatment groups are ordered correctly in the plot panels.
    Treatment = factor(Treatment, levels = c(
      "E. faecalis + Sucrose Water", "E. faecalis + Regular Water",
      "PBS + Sucrose Water", "PBS + Regular Water"
    ))
  ) %>%
  # Remove any rows where time could not be parsed.
  filter(!is.na(Time_h)) %>% 
  mutate( experiment_no = factor( experiment_no))

plot_data %>% write_csv('../data/R33_germ_free_data_for_plot.csv')
```


## slope for the growth trajectory in the first 48 hours 

```{r}
library(lme4)
library(lmerTest)
# test the slope
slope_df <- plot_data %>% 
  select(mouse_identifier, Treatment, experiment_no, log10_CFUs_per_GramStool, Time_h) %>% 
  drop_na() %>% 
  filter(Time_h <= 48 & Time_h > 0) %>% 
  mutate(Treatment = as.character(Treatment)) %>% 
  filter(str_detect(Treatment, '^E. faecalis')) %>% 
  mutate(Treatment = factor(Treatment, levels = c('E. faecalis + Regular Water','E. faecalis + Sucrose Water')),
         Time_h = factor(Time_h))

  

# Fit the linear mixed-effects model.
# The formula `log10_CFUs_per_GramStool ~ Treatment * Time_h` tests for the main effects
# of Treatment and Time, as well as their interaction. The interaction term is key.
# `(1 | experiment_no)` accounts for baseline differences between experiments.
# `(1 | mouse_identifier)` accounts for repeated measurements from the same mouse.
mixed_model <- lmer(log10_CFUs_per_GramStool ~ Treatment * Time_h + 
                      (1 | experiment_no) + (1 | mouse_identifier), 
                    data = slope_df)

# Get the summary of the model, which includes p-values for the fixed effects.
model_summary <- summary(mixed_model)

# Print the summary to the console.
print(model_summary)


```

## CFU plot

```{r}
# Calculate the median CFU for each group at each time point.
# This will be used to draw the dashed median line.
median_data <- plot_data %>%
  group_by(Treatment, Time_h) %>%
  summarize(median_CFU = median(`CFUs_per_GramStool`, na.rm = TRUE), .groups = 'drop')

# Create the improved plot.
CFU <- ggplot() +
  # Layer 1: Individual mouse trajectories (now thinner and more transparent).
  geom_line(data = plot_data, aes(x = Time_h, y = `CFUs_per_GramStool`, group = mouse_identifier, color = Treatment),
            linewidth = 0.5, alpha = 0.4) +
            
  # Layer 2: Individual data points (now bigger).
  geom_point(data = plot_data, aes(x = Time_h, y = `CFUs_per_GramStool`, color = Treatment),
             size = 2.5, alpha = 0.6) +
             
  # Layer 3: The median line (thicker and dashed).
  geom_line(data = median_data, aes(x = Time_h, y = median_CFU, color = Treatment),
            linetype = "dashed", linewidth = 1.2) +
            
  # Facet_wrap creates separate panels for each treatment group.
  facet_wrap(~Treatment) +
  
  # Set the Y axis to a log10 scale and format the labels as 10^x.
  scale_y_log10(
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x)),
    limits = c(1, NA) # Start the axis at 1 (log10=0).
  ) +
  
  # Manually set the colors to match the previous plot.
  scale_color_manual(values = c(
    "E. faecalis + Sucrose Water" = "#F8766D", # Reddish-Pink
    "PBS + Sucrose Water" = "#F8766D",
    "E. faecalis + Regular Water" = "#00BFC4", # Teal/Cyan
    "PBS + Regular Water" = "#00BFC4"
  )) +
    scale_x_continuous(
    breaks = c(0, 4, 8, 24, 48, 144),
    labels = c("0", "4", "8", "24", "48", "144")
  ) +
  #scale_shape_manual(values = c("1" = 16, "2" = 18)) + # Circle and diamond 
  # Use a clean theme and add informative titles and labels.
  theme_bw(base_size = 14) +
  labs(
    #title = "Bacterial Colonization Timeline in Germ-Free Mice",
    #subtitle = "Individual mouse trajectories with group medians",
    x = "Time Post-Gavage (Hours)",
    y = "CFU / gram of feces"
  ) +
  
  # Customize the plot's appearance.
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_text(size = 9),
    strip.text = element_text(face = "bold", size = 12)
  )+
  
  # This is the key part to remove the fill legend but keep the shape legend.
  guides(color = "none", shape = 'none')
CFU

ggsave('../data/R33_gf_full_course_with_median_combined2.jpg', width = 10, height = 7)
```

## zoom in on the 24 hour only  between the two conditions

```{r}
# zoom in on the 24 hour only 
cfu_24h_data <- plot_data %>%
  filter( Time_h == 24 ) %>% 
  # Convert the Experiment column to a factor for plotting.
  mutate(experiment_no = as.factor(experiment_no))

# Create the boxplot.
ggplot(cfu_24h_data, aes(x = Treatment, y = log10_CFUs_per_GramStool)) +
  # Add the boxplot layer. `outlier.shape = NA` hides the default outlier points
  # since we will be plotting all points ourselves.
  geom_boxplot(width = 0.5, alpha = 0.4, outlier.shape = NA) +
  
  # Add the individual data points, jittered to prevent overlap.
  # The shape of the points is mapped to the 'Experiment' column.
  geom_jitter(aes(shape = experiment_no), width = 0.25, size = 4) +
  
  # This function from ggpubr runs the Wilcoxon test and adds the p-value.
  stat_compare_means(
    method = "wilcox.test",
    comparisons = list(c("E. faecalis + Sucrose Water", "E. faecalis + Regular Water")),
    label = "p.format" # Displays the p-value in a standard format.
  ) +
  
  
  # Manually define the shapes for the two experiments.
  scale_shape_manual(values = c("1" = 16, "2" = 17)) + # Circle and Triangle
  
  # Add informative titles and labels.
  theme_bw(base_size = 14) +
  labs(
    title = "E. faecalis Colonization at 24 Hours",
    x = "Treatment Group",
    y = "log10 CFU / gram of feces",
    shape = "Experiment" # Legend title for the shapes.
  ) +
  
  # Customize the plot's appearance.
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 10, hjust = 1),
    legend.position = "bottom"
  ) 


```

# median VS mean   

```{r}
# # one figure where I only plot the median and the mean line
# summary_data <- bind_rows(
#   gavage_median_data %>% mutate(Metric = 'summarized with median') %>% rename(CFU = median_CFU),
#   gavage_mean_data %>% mutate(Metric = 'summarized with mean')%>% rename(CFU = Mean_CFU)
# )
# 
# ggplot(summary_data, aes(x = Time_h, y = CFU, color = Treatment, linetype = Metric)) +
#   # Draw lines connecting the points. The linetype (solid vs. dashed) will distinguish mean from median.
#   geom_line(linewidth = 1.2, alpha = 0.8) +
#   # Add points to show the exact measurement at each time point.
#   geom_point(size = 3.5, alpha = 0.8) +
#   
#   # Set the Y axis to a log10 scale and format the labels as 10^x.
#   scale_y_log10(
#     breaks = trans_breaks("log10", function(x) 10^x),
#     labels = trans_format("log10", math_format(10^.x)),
#     # We add a limit of 1 so the zero values are included at the baseline.
#     limits = c(1, NA)
#   ) +
#   
#   # Use a clean theme and add informative titles and labels.
#   theme_bw(base_size = 14) +
#   facet_grid(~ Metric) +
#   labs(
#     title = "Mean vs. Median E. faecalis Colonization",
#     subtitle = "Comparing Sucrose vs. Regular Water Groups",
#     x = "Time Post-Gavage (Hours)",
#     y = "CFU / gram of feces",
#     color = "Treatment Group",
#     linetype = "Metric"
#   ) +
#   
#   # Customize the plot's appearance.
#   theme(
#     plot.title = element_text(face = "bold", hjust = 0.5),
#     plot.subtitle = element_text(hjust = 0.5),
#     legend.position = "right",
#     legend.title = element_text(face = "bold")
#   )
# 
# ggsave('../data/R33_gm_summaries_2.jpg', width = 10, height = 5)
```

# the AUC test    

```{r}
library(DescTools) 
library(ggpubr)

analysis_data <- plot_data %>%
  # First, filter for only the groups of interest (E. faecalis gavage).
  filter(str_starts(`Mouse Group`, "1-") | str_starts(`Mouse Group`, "2-")) %>% 
  mutate(mouse_identifier = str_glue('{experiment_no}__{`Mouse Group`}'))

# Now, calculate the AUC for each mouse.
# We group by mouse ID and treatment, then summarize to get one AUC value per mouse.
auc_results <- analysis_data %>%
  group_by(mouse_identifier, Treatment,experiment_no ) %>%
  summarize(
    # The AUC function from DescTools calculates the area under the curve.
    # We use the log-transformed CFU values for this calculation.
    auc_value = AUC(x = Time_h, y = log10_CFUs_per_GramStool),
    # The .groups = 'drop' argument cleans up the grouping structure afterward.
    .groups = 'drop'
  ) %>%
  # It's important to convert the grouping variables to factors for ANOVA.
  mutate(
    Treatment = as.factor(Treatment),
    experiment_no = as.factor(experiment_no)
  )

```

## test the AUC difference adjusting for the experiment number

```{r}
# Fit the two-way ANOVA model.
# The formula `AUC ~ Treatment + experiment_no` tests for the main effects
# of both Treatment and Experiment number.
# You could also test for an interaction with `AUC ~ Treatment * experiment_no`.
anova_model <- aov(auc_value ~ Treatment + experiment_no, data = auc_results)

# Get the summary of the model, which includes the p-values for each factor.
anova_summary <- summary(anova_model)

# Print the summary to the console.
print(anova_summary)

# Look at the p-value (in the `Pr(>F)` column) for the 'Treatment' row.
# p > 0.05: we do not have enough evidence to conclude that there is a
#   significant difference between the treatment groups.



# Extract the p-value for the 'Treatment' effect from the summary table.
# The [1, 5] index gets the value from the first row, fifth column (`Pr(>F)`).
p_value <- anova_summary[[1]]$`Pr(>F)`[1]

# Format the p-value text for plotting, adding the asterisk.
p_label <- paste0("p = ", format(p_value, digits = 2, nsmall = 2), "*")
```

## AUC panel

```{r}
# To visualize the result, a simple boxplot of the AUC values is helpful.
auc_plot <- ggplot(auc_results, aes(x = Treatment, y = auc_value, fill = Treatment)) +
  geom_boxplot(alpha = 0.7,  outlier.shape = NA) +
  geom_jitter(width = 0.35, size = 2) + 
  # Manually add the p-value annotation.
  annotate(
    "text",
    x = 1.5, y = max(auc_results$auc_value) + 50, # Position the text
    label = p_label,
    size = 3,
    fontface = "bold"
  ) +
    # Add a bracket to show which groups are being compared.
  annotate(
    "segment",
    x = 1, xend = 2, y = max(auc_results$auc_value) + 30, yend = max(auc_results$auc_value) + 30,
    linewidth = 0.8
  ) +
  #scale_shape_manual(values = c("1" = 16, "2" = 18)) + # Circle and diamond 
  labs(
    #title = "Comparison of Total Bacterial Load (AUC)",
    x = "",
    y = "Area Under Curve\nlog10(CFU/g) * hours",
    #shape = "Experiment", # This will be the title for the shape legend.
    caption = "*p-value from two-way ANOVA adjusting for experiment number"
  ) +
  theme_classic(base_size = 11) +
  theme(legend.position = "bottom", plot.caption = element_text(hjust = 0, face = "italic"))+
  
  # This is the key part to remove the fill legend but keep the shape legend.
  guides(fill = "none", shape = 'none') + 
  expand_limits(y = 0)

auc_plot 

ggsave('../data/R33_GF_AUC_starts_at_0.pdf', width = 4, height = 4)
```



```{r}
# interactive model test
anova_model_interaction <- aov(auc_value ~ Treatment * experiment_no, data = auc_results)

# Get the summary of the model, which includes the p-values for each factor.
anova_summary_interaction <- summary(anova_model_interaction)

# the interaction term (Treatment:experiment_no) is actually testing: "Did the effect of sucrose water depend on which experiment it was in?"
# not the question I am interested in but it is good to confirm that whatever effect (or lack of effect) the sucrose had, it was the same in both of your independent experiments. This demonstrates that your experimental system is reproducible.
# Because the interaction is not significant, it is statistically valid to remove that term from the model and look at the simpler "main effects" model (which is what you had before). This gives you more power to detect a true effect of the Treatment itself.
```

# assemble 

```{r}
# --- 4. Stack All Three Plots Vertically (A, B, C) ---

final_plot <-  plot_grid( design , CFU , auc_plot,
                 label_size = 12, nrow  =3,
                 rel_heights = c(1.2,2,1.8),labels = "auto", 
                align = 'vh', axis = 'lrtb')


title <- ggdraw() + 
  draw_label("Rebuttal Fig X",fontface = 'bold',x = 0,hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 0))

combined <- plot_grid(
  title, final_plot,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.05, 1)
) +theme(plot.margin = unit(c(1,1,1,1), "cm"))

ggsave('../data/R33_germ_free_mouse.pdf',
      width = 8.5, height = 11, units = "in", device = 'pdf', 
      dpi = 300)  
```