---
title: "Enterococcus -- stratify by abx class"
output: html_document
date: "2025-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(cowplot)
library(broom.mixed)
library(brms)   
library(ggpubr)
library(tidybayes)
library(brmstools)
library(bayesplot)
library(gt)
library(rcompanion)
library(janitor)
library(ggtext)
rstan::rstan_options(auto_write = TRUE)
theme_set(theme_tidybayes() + panel_border())
```

Was the increased Enterococcus abundance also stratified by antibiotic class?

# stratify fecal sample exposure by abx class like below:

anaerobic-targeting broad-spectrum: piperacillin-tazobactam, carbapenems, metronidazole, linezolid, and oral vancomycin
anaerobic-sparing broad-spectrum: cefepime

```{r}
# It's good practice to define these lists separately to make the code cleaner
# and easier to update later.
anaerobic_targeting <- c(
  "piperacillin_tazobactam",
  "meropenem", 
  "metronidazole",
  "linezolid"
)

anaerobic_sparing <- c("cefepime")

exposures <- read_csv('../../MSS_pipeline-/scripts/food_tree/data/191_all_medication_overlapping_with_prior_2_day_of_stool_samples_upated_2.csv') %>%
  # Now, create the new classification column using conditional logic.
  mutate(
    broad_spectrum_subtype = case_when(
      
      # Rule 1: Handle the anaerobic-targeting drugs by name.
      drug_name_clean %in% anaerobic_targeting ~ "anaerobic_targeting",
      
      # Rule 2: Handle the special case for oral vancomycin.
      drug_name_clean == "vancomycin" & route_clean == "oral" ~ "anaerobic_targeting",
      
      # Rule 3: Handle the anaerobic-sparing drugs.
      drug_name_clean %in% anaerobic_sparing ~ "anaerobic_sparing",
      
      # For any other drug that doesn't match the rules above,
      # it will be assigned NA (Not Available).
      TRUE ~ 'non_broad_spectrum'
    )
  )

# take a look at the broad spectrum exposure : drug name and the route
broad <- exposures %>% 
  filter(drug_category_for_this_study == 'broad_spectrum') %>% distinct(drug_name_clean, route_clean)
```
```{r}
# Now group_by(sampleid, sdrt, pid)  and if there is any anaerobic-targeting exposure then consider it anaerobic-targeting, otherwise if any  anaerobic-sparing, then consider the stool sample exposed to anaerobic-sparing, otherwise non-broad_spectrum
sample_summary <- exposures %>%
  group_by(sampleid, sdrt, pid) %>%
  summarise(
    # This logic establishes a hierarchy for classification.
    # If any drug in the group is anaerobic-targeting, the whole sample is classified as such.
    # Otherwise, if any is anaerobic-sparing, it's classified that way.
    # If neither, it's considered non-broad spectrum.
    exposure_class_category = case_when(
      any(broad_spectrum_subtype == "anaerobic_targeting") ~ "anaerobic_targeting",
      any(broad_spectrum_subtype == "anaerobic_sparing") ~ "anaerobic_sparing",
      TRUE ~ "non_broad_spectrum"
    ),
    .groups = 'drop' # It's good practice to drop the grouping after summarizing.
  )

sample_summary |> write_csv('../data/R46_sample_summary.csv')
```

## tally the exposure numbers to different class 

```{r}
# Calculate the total number of unique patients and samples for percentages.
total_samples <- nrow(sample_summary)
total_patients <- n_distinct(sample_summary$pid)

# First, create the summary data, including clean labels for the final table.
summary_data <- sample_summary %>%
  group_by(exposure_class_category) %>%
  summarise(
    n_samples = n(),
    n_patients = n_distinct(pid),
    .groups = 'drop'
  ) %>%
  mutate(
    percent_samples = (n_samples / total_samples) * 100,
    percent_patients = (n_patients / total_patients) * 100,
    # Create the clean, descriptive labels for the final table rows.
    exposure_category_label = recode(exposure_class_category,
                                    "anaerobic_targeting" = "Anaerobic-targeting",
                                    "anaerobic_sparing" = "Anaerobic-sparing",
                                    "non_broad_spectrum" = "Non-broad-spectrum")
  )

# Next, create the final table structure with the 'Total' row.
summary_table_for_gt <- summary_data %>%
  # Arrange the data in the desired order before adding the total row.
  arrange(factor(exposure_class_category, levels = c("non_broad_spectrum", "anaerobic_sparing", "anaerobic_targeting"))) %>%
  # Select and reorder the final columns for the table.
  select(exposure_category_label, n_samples, percent_samples, n_patients, percent_patients)


# --- Create the Publication-Ready Table with gt ---
gt_table <- summary_table_for_gt %>%
  # Initialize the gt object, using the label column for row names.
  gt(rowname_col = "exposure_category_label") %>%
  
  # Add a main title for the table.
  tab_header(
    title = 'Cohort Antibiotic Class Exposure Summary'
  ) %>%
  
  # Create spanner headings to group related columns together.
  tab_spanner(
    label = md("**Samples**"),
    columns = c(n_samples, percent_samples)
  ) %>%
  tab_spanner(
    label = md("**Patients**"),
    columns = c(n_patients, percent_patients)
  ) %>%
  
  # Customize the final column labels.
  cols_label(
    n_samples = "N",
    percent_samples = "%",
    n_patients = "N",
    percent_patients = "%"
  ) %>%
  
  # Format the large count numbers to use comma separators.
  fmt_number(
    columns = c(n_samples, n_patients),
    decimals = 0
  ) %>%
  
  # Format the percentage columns to one decimal place.
  fmt_number(
    columns = c(percent_samples, percent_patients),
    decimals = 1
  ) %>%
  
  # Center align the content of all numeric columns.
  cols_align(
    align = "center",
    columns = where(is.numeric)
  ) %>%

  # Add some final styling options for a polished look.
  opt_table_font(font = "sans-serif") %>%
  opt_table_outline() %>% 
  gt::gtsave(filename = "../data/R46_number_tally_abx_exposure_class.png") 
```

# ASV1 model 

```{r}
clr_res <- read_csv('../data/R29_all_ASV_CLR_counts.csv')

clr_wide <- clr_res %>% 
   select(sampleid, asv_1) %>% rename(asv_1_CLR = asv_1)

meta_ASV1 <- read_csv('../data/153_combined_META.csv') %>% 
  mutate(intensity = factor(intensity, levels = c('nonablative','reduced','ablative'))) %>%
  mutate(pid = factor(pid)) %>%
    # This divides all the new intake columns by 100.
  mutate(across(starts_with("fg_"), ~ .x / 100)) %>% 
  inner_join(sample_summary) %>% 
  mutate(exposure_class_category = factor(exposure_class_category, levels = c("non_broad_spectrum", "anaerobic_sparing", "anaerobic_targeting"))) %>% 
  mutate(timebin = cut_width(sdrt, 7, boundary=0, closed = 'left')) %>% 
  inner_join(clr_wide)
```

## model 

```{r}
all_food_vars <- meta_ASV1 %>% select(starts_with('fg')) %>% colnames()

# Create the interaction terms for all food variables in this iteration
interaction_terms <- paste(all_food_vars, "exposure_class_category", sep = ":")

# Build the full formula string dynamically
formula_string <- paste(
  "asv_1_CLR ~ 0 + intensity + exposure_class_category + TPN + EN +",
  paste(interaction_terms, collapse = " + "),
  "+ (1 | pid) + (1 | timebin)"
)

# Convert the string to a formula object
formula <- brms::bf(as.formula(formula_string))

# Build the priors by adding them together.
# This single prior() call applies to all coefficients listed in `all_food_coefs`.
priors <-
    prior(normal(0, 1), class = 'b') + # General prior for all food effects
    # Specific priors that override the general one for non-food covariates
    prior(normal(0, 0.1), class = 'b', coef = "TPNTRUE") +
    prior(normal(0, 0.1), class = 'b', coef = "ENTRUE") +
    prior(normal(0, 0.5), class = 'b', coef = "exposure_class_categoryanaerobic_sparing") +
    prior(normal(0, 0.5), class = 'b', coef = "exposure_class_categoryanaerobic_targeting") 

# Fit the model (using fewer iterations for this example to run quickly)
model_fit <- brm(
  formula = formula,
  data = meta_ASV1,
  prior = priors,
    warmup = 1000, iter = 3000,
    chains = 4, cores = 4,
    seed = 123,
    silent = 2,
  control = list(adapt_delta = 0.99)
)

results_df <- tidy(model_fit, conf.int = TRUE) %>%
  mutate(conf.low = round(conf.low, 2),
         conf.high = round(conf.high, 2))

```

## forest plot

```{r}
# --- Data Wrangling ---
# Clean up the term names to make them usable for plotting.
# This involves separating the food group from the antibiotic category.
plot_data <- results_df %>%
  # Filter out the main antibiotic effects for now, we're focusing on diet.
  filter(str_detect(term, "fg_")) %>%
  # Separate the 'term' column into two new columns: 'antibiotic_group' and 'food_group'.
  separate(term, into = c("antibiotic_group", "food_group"), sep = ":fg_") %>%
  # Clean up the names to be more readable for the plot labels.
  mutate(
    antibiotic_group = str_remove(antibiotic_group, "exposure_class_category") %>%
                       str_replace("non_broad_spectrum", "Non-broad-spectrum") %>%
                       str_replace("anaerobic_targeting", "Anaerobic-targeting") %>%
                       str_replace("anaerobic_sparing", "Anaerobic-sparing"),
    
    # Standardize the food group names to match the desired labels exactly.
    food_group_clean = case_when(
        food_group == "sweets" ~ "Sweets",
        food_group == "grain"  ~ "Grains",
        food_group == "milk"   ~ "Milk",
        food_group == "egg"    ~ "Eggs",
        food_group == "legume" ~ "Legumes",
        food_group == "meat"   ~ "Meats",
        food_group == "fruit"  ~ "Fruits",
        food_group == "oils"   ~ "Oils",
        food_group == "veggie" ~ "Vegetables",
        TRUE ~ str_to_title(food_group) # A fallback for any other groups
    )
  ) %>%
  # Set the order of the facets so "No Antibiotics" comes first.
  mutate(food_group_ordered = factor(food_group_clean, levels = rev(c(
        "Sweets", "Grains", "Milk", "Eggs", "Legumes", 
        "Meats", "Fruits", "Oils", "Vegetables"
    ))),
    antibiotic_group = factor(antibiotic_group, 
                                   levels = c("Non-broad-spectrum", "Anaerobic-sparing", "Anaerobic-targeting"))) %>%
  # Create a new column to identify significant results
  # The condition is TRUE if conf.low and conf.high have the same sign (i.e., don't cross zero)
  mutate(is_significant = (conf.low * conf.high) > 0) 

# --- Plotting ---
# Create the faceted dot-and-whisker plot.
plot_stratify_abx_class <- ggplot(plot_data, aes(x = estimate, y = food_group_ordered)) +
  # Add the horizontal line for the confidence interval.
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, color = "gray50") +
   # Use geom_pointrange to show the estimate (point) and confidence interval (line)
   geom_pointrange(
   aes(xmin = conf.low, xmax = conf.high, color = is_significant),
     size = 0.25,linewidth = 1,
  ) +
  # Add a vertical line at zero, which is the "no effect" line.
   geom_vline(xintercept = 0, linetype = "solid", color = "blue",size = 0.8) +
  # This is the key part: create the three separate panels.
  facet_wrap(~antibiotic_group, ncol = 3) +
  # Add labels and a title.
  labs(
    title = "Diet Predictor effects on ASV 1 (E. faecium),\nstratifying samples by exposure to different antibiotcs class",
    x = "CLR(ASV 1) Change per 100g Food Intake",
    y = "Food Group"
  ) +
  scale_color_manual(
    values = c(
      "TRUE" = "red",
      "FALSE" = "black"
    ), name = "Effect Status" # Legend title
  ) +
  # Use a clean, publication-ready theme.
  theme_bw() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = 'none',
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 11)
  )

plot_stratify_abx_class
```

# assemble 

```{r}
table_tally <- ggdraw() + draw_image("../data/R46_number_tally_abx_exposure_class.png", scale = 1)

final_plot <- plot_grid( table_tally ,  plot_stratify_abx_class , 
                nrow  = 2,
                 rel_heights = c(1.6,2),
                labels = 'auto',
                align = 'vh', axis = 'lrtb')


title <- ggdraw() + 
  draw_label("Rebuttal Fig X",fontface = 'bold', x = 0,hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 0))

combined <- plot_grid(
  title, final_plot,
  ncol = 1,
  # rel_heights values control vertical title margins
  rel_heights = c(0.05, 1)
) +theme(plot.margin = unit(c(1,1,1,1), "cm"))

ggsave('../data/R46_abx_three_levels_class_stratification.pdf',
      width = 8.5, height = 11, units = "in", device = 'pdf', 
      dpi = 300)        
```
